### 2016-06-23 submission/Westcott_Opticlust_PeerJ_2016.Rmd
Initial commit (borrowed a lot from Schloss_Cluster_PeerJ_2015)

---
title: "**Need a title...**"
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
csl: peerj.csl
fontsize: 11pt
geometry: margin=1.0in
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
	print(x)
	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
		}

	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)
```

\begin{center}
\vspace{25mm}
Sarah L. Westcott and Patrick D. Schloss${^\dagger}$

\vspace{30mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
\end{center}

\newpage
\linenumbers

## Abstract

\newpage

## Introduction

## Methods

## Results and Discussion

## Conclusions

\newpage

## Figures

\newpage

## References

### 2016-06-24 code/mice.R
Add mice processing R script

@@ -1,0 +1,19 @@
+get_contigsfile <- function(p){
+	p <- gsub("\\/$", "", p)
+	f <- list.files(p)
+	f <- f[!grepl("Mock", f)] #let's ignore the mock community data
+
+	r1 <- f[grep("_R1_", f)]
+	r1.group <- sub("_S.*", "", r1)
+
+	r2 <- f[grep("_R2_", f)]
+	r2.group <- sub("_S.*", "", r2)
+
+	stopifnot(r1.group == r2.group)
+
+	files.file <- paste0(gsub(".*\\/(.*)", "\\1", p), ".files")
+	p.files.file <- paste0(p, "/", files.file)
+
+	write.table(file=p.files.file, cbind(r1.group, r1, r2), sep="\t", quote=F, row.names=F, col.names=F)
+}
+

### 2016-06-24 code/even.R
Add code and generalize rule to generate even data

@@ -1,0 +1,12 @@
+get_seq_names <- function(fasta_file){
+	fasta_data <- scan(fasta_file, what=character(), sep='\n', quiet=T)
+	is_seq_name <- grepl("^>", fasta_data)
+	gsub('>', '', fasta_data[is_seq_name])
+}
+
+even <- function(fasta_file, n_seqs=100){
+	seq_names <- get_seq_names(fasta_file)
+	count_data <- data.frame(Representative_Sequence=seq_names, total=100)
+	count_file <- gsub('fasta', 'count_table', fasta_file)
+	write.table(count_data, count_file, row.names=F, quote=F, sep='\t')
+}

### 2016-06-24 code/staggered.R
Add code to build staggered communities

@@ -1,0 +1,14 @@
+get_seq_names <- function(fasta_file){
+	fasta_data <- scan(fasta_file, what=character(), sep='\n', quiet=T)
+	is_seq_name <- grepl("^>", fasta_data)
+	gsub('>', '', fasta_data[is_seq_name])
+}
+
+staggered <- function(fasta_file, max_n_seqs=200){
+	set.seed(1)
+	seq_names <- get_seq_names(fasta_file)
+	freqs <- floor(runif(length(seq_names), 1, max_n_seqs))
+	count_data <- data.frame(Representative_Sequence=seq_names, total=freqs)
+	count_file <- gsub('fasta', 'count_table', fasta_file)
+	write.table(count_data, count_file, row.names=F, quote=F, sep='\t')
+}

### 2016-06-24 code/soil.R
Add code files to process He soil dataset

@@ -1,0 +1,28 @@
+library(rentrez)
+NCBI_data <- entrez_link(dbfrom = "pubmed", id = "18043639", db = "all")
+n_popsets <- length(NCBI_data$pubmed_popset)
+
+fetched_data <- character()
+for(i in 1:n_popsets){
+    fetched_data[i] <- entrez_fetch(db = "popset", rettype = 'fasta', id = NCBI_data$pubmed_popset[i])
+}
+
+full_fasta_file <- paste(fetched_data, collapse="")
+full_fasta_file <- gsub("\n\n", "\n", full_fasta_file)
+split_fasta_file <- unlist(strsplit(full_fasta_file, ">"))[-1]
+
+seq_ids <- sub("^gi\\|\\d*\\|gb\\|EF(\\d*).1\\|.*", "\\1", split_fasta_file)
+seq_ids <- as.numeric(seq_ids)
+
+seq_bases <- sub("^.*partial sequence\n", "\\1", split_fasta_file)
+seq_bases <- gsub("\\n", "", seq_bases)
+
+used_ids <- 308591:361836
+
+keep_sequences <- seq_ids %in% used_ids
+
+keep_ids <- seq_ids[keep_sequences]
+keep_bases <- seq_bases[keep_sequences]
+keep_fasta <- paste0(">EF", keep_ids, "\n", keep_bases)
+
+write(keep_fasta, "data/soil/soil.raw.fasta")

### 2016-06-24 code/mice.R
Fix problems with generation of mice.files

@@ -1,6 +1,6 @@
 get_contigsfile <- function(p){
 	p <- gsub("\\/$", "", p)
-	f <- list.files(p)
+	f <- list.files(path=p, pattern="*.fastq.gz")
 	f <- f[!grepl("Mock", f)] #let's ignore the mock community data
 
 	r1 <- f[grep("_R1_", f)]

### 2016-06-24 code/marine.R
Pool duplicate marine samples

@@ -1,15 +1,15 @@
 make_files_file <- function(){
-	sample_map <- c("SRR3085688" = "JLB_A",
-									"SRR3085689" = "ARD_A",
+	sample_map <- c("SRR3085688" = "JLB",
+									"SRR3085689" = "ARD",
 									"SRR3085690" = "CJ",
 									"SRR3085691" = "TBON",
 									"SRR3085692" = "CJ2",
 									"SRR3085693" = "LBK",
-									"SRR3085694" = "FWC_A",
-									"SRR3085695" = "CJ_A",
-									"SRR3085696" = "TBON_A",
-									"SRR3085697" = "CJ2_A",
-									"SRR3085698" = "LBK_A",
+									"SRR3085694" = "FWC",
+									"SRR3085695" = "CJ",
+									"SRR3085696" = "TBON",
+									"SRR3085697" = "CJ2",
+									"SRR3085698" = "LBK",
 									"SRR3085699" = "JLB",
 									"SRR3085700" = "ARD",
 									"SRR3085701" = "FWC")

### 2016-06-24 code/sediment.R
Add code to run sediment data

@@ -1,0 +1,18 @@
+make_files_file <- function(){
+
+	mimarks <- read.table(file="data/sediment/sediment.metadata", header=T,
+	 											stringsAsFactors=FALSE)
+
+	sample_map <- mimarks$Sample_Name_s
+	names(sample_map) <- mimarks$Run_s
+
+	read_1 <- list.files(path="data/sediment/", pattern="*1.fastq.gz")
+	read_2 <- gsub("_1", "_2", read_1)
+
+	stub <- gsub("_1.*", "", read_1)
+	sample <- sample_map[stub]
+
+	files <- data.frame(sample, read_1, read_2)
+	write.table(files, "data/sediment/sediment.files", row.names=F, col.names=F, quote=F, sep='\t')
+
+}

### 2016-06-24 code/landfill.R
Add code for processing landfill data

@@ -1,0 +1,20 @@
+library(openxlsx)
+
+get_oligos_file <- function(){
+	metadata <- read.xlsx('data/landfill/landfill.xlsx')
+	metadata <- metadata[!is.na(metadata$Barcode.Sequence),]
+
+	barcode <- metadata$Barcode.Sequence
+	for_primer <- metadata$"Linker/Primer.Sequence"
+	rev_primer <- metadata$Reverse.Primer.Sequence
+	sample <- paste(metadata$Sample, metadata$Replicate, sep="_")
+
+	forward <- cbind("forward", unique(for_primer), "")
+	reverse <- cbind("reverse", unique(rev_primer), "")
+	barcodes <- cbind("barcode", barcode, sample)
+
+	oligos <- rbind(forward, reverse, barcodes)
+
+	write.table(oligos, file="data/landfill/landfill.oligos", quote=F,
+ 							row.names=F, col.names=F, sep='\t')
+}

### 2016-06-25 code/sediment.R
Fix delimeter in read.table function call

@@ -1,7 +1,7 @@
 make_files_file <- function(){
 
 	mimarks <- read.table(file="data/sediment/sediment.metadata", header=T,
-	 											stringsAsFactors=FALSE)
+	 											stringsAsFactors=FALSE, sep='\t')
 
 	sample_map <- mimarks$Sample_Name_s
 	names(sample_map) <- mimarks$Run_s

### 2016-08-02 code/aggregate_stats.R
Aggregate timing and sens/spec data

@@ -1,0 +1,91 @@
+cat_cols_files <- function(x){
+	x["frac"] <- gsub("\\.", "_", x["frac"])
+	x["rep"] <- gsub(" ", "0", x["rep"])
+	paste(x["dataset"], x["frac"], x["rep"], x["method"], sep='.')
+}
+
+build_cluster_stats_files <- function(x){
+	seed <- cat_cols_files(x)
+	paste0("data/", x["dataset"], "/", seed, ".stats")
+}
+
+build_cluster_sensspec_files <- function(x){
+	seed <- cat_cols_files(x)
+	paste0("data/", x["dataset"], "/", seed, ".sensspec")
+}
+
+build_dist_stats_files <- function(x){
+	if(x["method"] %in% c("accuracy", "f1score", "fn", "mcc", "nn")){
+		x["method"] <- 'sm'
+	} else if(x["method"] == 'an'){
+		x["method"] <- 'lg'
+	} else {
+		x["method"] <- NA
+	}
+
+	seed <- cat_cols_files(x)
+
+	stats <- NA
+	if(!is.na(x["method"])){
+		stats <- paste0("data/", x["dataset"], "/", seed, ".stats")
+	}
+	unlist(stats)
+}
+
+
+get_stats <- function(file_name){
+
+	data <- scan(file_name, what="", sep='\n', quiet=T)
+
+	stats <- c(NA,NA)
+
+	if(sum(z <- grepl("^Out of memory", data) != 0)){
+		stats <- c(NA,NA)
+	} else if(sum(z <- grepl("^FINISHED", data) != 0)){
+		parsed <- unlist(strsplit(data[z], " "))
+		stats <- as.numeric(parsed[c(3,11)])
+	} else if(sum(z <- grepl("^TIMEOUT", data) != 0)){
+		stats <- c(NA,NA)
+	} else if(sum(z <- grepl("^MEM_RSS", data) != 0)){
+		stats <- c(NA,NA)
+	}
+
+	return(stats)
+}
+
+
+datasets <- c("even", "human", "marine", "mice", "sediment", "soil", "staggered")
+
+fractions <- seq(0.2,1,0.2)
+n_fracs <- length(fractions)
+
+replicates <- 1:10
+n_reps <- length(replicates)
+
[…]

### 2016-08-02 submission/Westcott_Opticlust_mSystems_2016.Rmd
Target to mSystems

--- 2016-06-23 submission/Westcott_Opticlust_PeerJ_2016.Rmd
@@
       in_header: header.tex
-csl: peerj.csl
+csl: msystems.csl
 fontsize: 11pt

### 2016-08-04 code/fire_off_cluster_jobs.sh
Generalize code to generate other file types

@@ -1,6 +1,7 @@
 #!/bin/bash
 
 LIBRARY=$1
+TYPE=$2
 
 FRACTIONS=(1_0 0_8 0_6 0_4 0_2)
 REPS=(01 02 03 04 05 06 07 08 09 10)
@@ -11,27 +12,27 @@ do
 for R in ${REPS[@]}
 do
 cat head8.batch > $LIBRARY.$F.$R.qsub
-echo "make data/$LIBRARY/$LIBRARY.$F.$R.accuracy.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.an.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.f1score.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.fn.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.mcc.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.nn.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.swarm.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.vagc_1.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.vdgc_1.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.vagc_8.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.vdgc_8.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.uagc.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.udgc.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.vdgc_split5_1.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.vdgc_split5_8.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.an_split5_1.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.an_split5_8.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.mcc_split5_1.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.mcc_split5_8.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.otuclust.list;
-make data/$LIBRARY/$LIBRARY.$F.$R.sumaclust.list;" >> $LIBRARY.$F.$R.qsub
[…]

### 2016-08-05 code/sediment.R
Remove sediment dataset from the analysis

@@ -1,18 +1,0 @@
-make_files_file <- function(){
-
-	mimarks <- read.table(file="data/sediment/sediment.metadata", header=T,
-	 											stringsAsFactors=FALSE, sep='\t')
-
-	sample_map <- mimarks$Sample_Name_s
-	names(sample_map) <- mimarks$Run_s
-
-	read_1 <- list.files(path="data/sediment/", pattern="*1.fastq.gz")
-	read_2 <- gsub("_1", "_2", read_1)
-
-	stub <- gsub("_1.*", "", read_1)
-	sample <- sample_map[stub]
-
-	files <- data.frame(sample, read_1, read_2)
-	write.table(files, "data/sediment/sediment.files", row.names=F, col.names=F, quote=F, sep='\t')
-
-}

### 2016-08-05 code/run_otuclust.R
Remove sequence number to generate a uniqued list file

@@ -1,13 +1,28 @@
 # the otuclust format puts otus on separate rows and the sequences within each
-# otu on the same line separated by tabs
+# otu on the same line separated by tabs. this will return a unique'd list file
+# that only has the names of the unique sequences
 
-otuclust_to_list <- function(otuclust_file_name){
+parse_otus <- function(otu_line){
+	otu_line <- gsub("_\\d*(\\s)", "\\1", otu_line)
+	otu_line <- gsub("_\\d*$", "", otu_line)
+	sequences <- unique(unlist(strsplit(otu_line, "\t")))
+	otus <- paste(sequences, collapse=',')
+}
+
+otuclust_to_list <- function(otuclust_file_name, count_file_name){
 	otuclust_data <- scan(otuclust_file_name, what="", sep="\n", quiet=TRUE)
-	otuclust_data <- gsub('\t', ',', otuclust_data)
+
+	count_data <- read.table(count_file_name, stringsAsFactors=F, header=T)
+	n_unique_seqs <- nrow(count_data)
+
+	otuclust_data <- sapply(otuclust_data, parse_otus)
+	n_seqs <- sum(nchar(otuclust_data) - nchar(gsub(",", "", otuclust_data)) + 1)
+
+	stopifnot(n_unique_seqs == n_seqs)
+
 	n_otus <- length(otuclust_data)
 	list_data <- paste(c("userLabel", n_otus, otuclust_data), collapse='\t')
 	list_file_name <- gsub("\\.clust", ".list", otuclust_file_name)
 	list_file_name <- gsub("\\.redundant", "", list_file_name)
 	write(list_data, list_file_name)
 }
-

### 2016-08-05 code/run_otuclust.sh

@@ -23,6 +23,6 @@ OTUCLUST_REP=$(echo $OTUCLUST_FASTA | sed 's/ng.redundant.fasta/rep/')
 
 otuclust -f fasta $OTUCLUST_FASTA --out-clust $OTUCLUST_CLUST --out-rep $OTUCLUST_REP -s 0.97 -m 1
 
-R -e "source('code/run_otuclust.R'); otuclust_to_list('$OTUCLUST_CLUST')"
+R -e "source('code/run_otuclust.R'); otuclust_to_list('$OTUCLUST_CLUST', '$COUNT')"
 
 rm $TEMP_FASTA $OTUCLUST_CLUST $OTUCLUST_FASTA $NG_FASTA $OTUCLUST_REP $REDUNDANT_GROUPS

### 2016-08-16 code/run_otuclust.R
Fix regular expression to remove trailing _1 rather than internal _1s

@@ -5,7 +5,7 @@
 parse_otus <- function(otu_line){
 	seqs <- unlist(strsplit(otu_line, '\t'))
 	seqs <- seqs[grepl("_1$", seqs)]
-	seqs <- gsub("_1", "", seqs)
+	seqs <- gsub("_1$", "", seqs)
 	otu <- paste(unique(seqs), collapse=',')
 }
 

### 2016-08-17 submission/Westcott_Opticlust_mSystems_2016.Rmd
Rough in Results section

--- 2016-06-23 submission/Westcott_Opticlust_PeerJ_2016.Rmd
@@
 ---
-title: "**Need a title...**"
+title: "**Opticlust: Improved method for assigning amplicon-based sequence data to operational taxonomic units**"
 bibliography: references.bib
@@
 
+Amplicon-based sequencing has provided incredible insights into Earth's microbial biodiversity.
+
+Numerous methods have been proposed
+
+Needed an objective criteria to evaluate the various methods methods
+
+Average neighbor is consistently the best method, but others including USEARCH and VSEARCH resulted in comparable results
+
+Difficulty with amount of memory and time required to complete average neighbor algorithm is a significant hurdle to completion of the method.
+
+Instead of retrospectively evaluating methods for their ability to correctly assign sequences we sought to develop a method that would prospectively assign sequences to OTUs to optimize classification metrics.
+
+## Results
+
+***Opticlust algorithm.***
+
+***Opticlust generates more robust OTUs than other methods.***
+
+***Opticlust is faster and uses less memory than other methods.***
+
+***The speed and memory requirements of Opticlust scale XXXXXly.***
+
+## Discussion
+Restate most important contributions
+* Optimized clustering based on an objective criteria. Results are meaningfully better than existing methods
+* Added benefits include smaller CPU and RAM footprint
+* Result is efficient analysis of large datasets without sacrificing clustering quality as has been experienced in heuristic methods.
+
+Choice of objective criteria
+* Value of using a metric that is based on all four parameters
+* Preference is for Matthew's Correlation Coefficient because... other options include F1 Score and accuracy
+
+Preference for Opticlust over cluster splitting approach
+* Risks are mis-classification and artificially splitting similar sequences between OTUs because they classify to different taxa
+* Still potential to merge the methods for more efficient processing.
+
+Based on our model, we propose the following...
+
 ## Methods
 
-## Results and Discussion
+***Sequence data and processing steps.***
 
-## Conclusions
+***Implementation of clustering algorithms.***
+
+***Benchmarking strategy.***
+
+## Results
+
+***Opticlust algorithm (Figure 1A).*** The Opticlust algorithm uses a list of all sequence names in the dataset, the pairs of sequences that are within a desired threshold of each other (e.g.
+0.03), and the metric that should be used to assess clustering quality (e.g.
+MCC).
+The sequence pairs that are within the user-specified threshold are used to indicate those sequences that should appear within the same OTU and those that have a distance larger than the threshold should not appear within the same OTU.
+The algorithm starts by placing each sequence within its own OTU and calculating the number of true positives, true negatives, false positives, and false negatives based on the pairwise distances.
+These counts are used to calculate the metric used for optimization.
+The algorithm proceeds by interrogating each sequence and re-calculating the metric for the cases where the sequence stays in its current OTU, is moved to each of the other OTUs, or is moved into a new OTU.
+The location that results in the largest value of the optimization metric indicates whether the sequence remains in its current OTU or is moved.
+Each iteration consists of interrogating every sequence in the dataset.
+The algorithm continues until the optimization metric stabilizizes.
+
+***Opticlust stopping criteria (Figure 1B).*** By default, the mothur-based implementation of the algorithm stops when the optimization metric does not change between iterations.
+It also allows the user to stop the algorithm if a maximum number of iterations is exceeded or if the change in the optimization metric is less than a stopping threshold.
+For each of the datasets we evaluated, between **XXX** and **XXX** iterations were needed to reach full conversion of the algorithm and between **XXX* and **XXX** iterations were needed to see a change in the metric that was less than 0.01%.
+To achieve full stabilization of the optimization metric, the **XXXX** algorithm required an average of **XX**% more iterations than the **XXXX** algorithm, which needed **XX**% more iterations than the **XXXX** algorithm.
+These results suggest that achieving full stabilization of the optimization metric adds computational effort, but that it is relatively small.
+
+***Opticlust-generated OTUs are as stable as those from other algorithms (Figure 1C).*** One concern that many have with *de novo* clustering algorithms is that their output is sensitive to the initial order of the sequences.
+An additional concern with the Opticlust algorithm is that it may stabilize at a local optimum.
+To evaluate these concerns we compared the results obtained using ten randomizations of the order that sequences were given to the algorithm.
+Similar to our previous analysis, we found that the MCC values obtained from the replicate clusterings did not vary by more than **0.XX**% and were actually less variable than what we observed using the other algorithms.
+The variation observed in clustering quality suggests that the algorithm generates OTU assignments with subtle variation, but that the assignments are of a similar quality.
+
+***Optimization of clustering using composite metrics significantly improves clustering quality (Figure 2, 3, 4).*** There are multiple metrics available to assess clustering quality.
+The mothur-based implementation of Opticlust allows the user to use the MCC, F1 score, accuracy, sensitivity, specificity, and the sum of true positives and negatives.
+The MCC, F1 score, and accuracy are preferred because each of them incorporates all four values from the confusion matrix while others only utilize two values.
+It is relatively straightforward to implement other metrics.
+**Comparison of MCC, accuracy, and F1 score to each other for each dataset using the full set of sequences.** Each metric outperformed the observed values for the other clustering algorithms indicating that regardless of the metric one uses to evaluate clustering quality, the Opticlust algorithm generates better OTU assignments than any of the other methods.
+The number of OTUs generated by an algorithm is often used as a metric for clustering quality.
+We did not observe a significant correlation between clustering quality as measured by MCC, F1 score, or accuracy and the number of OTUs generated by each algorithm.
+**Interestingly, the values of the MCC, F1 score, and accuracy generated when using each metric to optimize clustering were similar across implementations.** Based on these results and its previous use in the OTU assignment literature, the remainder of our analysis uses the MCC metric for optimization.
+
+***Optimization-based clustering is faster and uses less memory than current methods (Figure 5).*** Although not as important as the quality of clustering, the amount of time and memory required to assign sequences to OTUs is a legitimate concern.
+**XXXX**, **XXXX**, and **XXXX** were not able to complete within 50 hours for the **XXXX** and **XXXX** datasets.
+The Opticlust algorithm finished **XXX** to **XXX**% faster than the next fastest algorithm.
+Similarly, **XXXX**, **XXXX**, and **XXXX** were not able to complete using less than 48 GB of RAM for the **XXXX** and **XXXX** datasets.
+The Opticlust algorithm used **XXX** to **XXX**% less RAM than the next fastest algorithm.
+To evaluate howt the speed and memory usage scaled with the number of sequences in the dataset, we measured the time required and maximum RAM usage to cluster 20, 40, 60, 80, and 100% of the unique sequences in each dataset by each algorithm.
+Using the full datasets, we observed that the speed and memory requirements of the Opticlust algorithm scaled **XXXXXXX**.
+These results clearly demonstrate that Opticlust is faster and uses less memory than other algorithms currently in use.
+This complements the higher quality Opticlust OTU assignments relative to other methods.
+
+***Cluster splitting heuristic generates OTUs that are as good as non-split approach (Figure 6).*** We previously described a heuristic to accelerate OTU assignments where sequences were classified to taxonomic groupings and within each taxon sequences were assigned to OTUs.
+Using the average neighbor algorithm we found that this method was effective as clustering without using the classification information.
+We again assessed this approach using the average neighbor, VSEARCH abundance-based greedy clustering, and Opticlust by splitting the data at the family level and then clustering the sequences into OTUs.
+As before, we found that the quality of the clustering was as good using the classification data as without.
+An advantage of this approach is that because each taxon is distinct, it is possible to parallelize the assignment by clustering each taxon on a separate processor.
+Although the taxa are not equally abundant, there was still a considerable improvement in execution time and the amount of memory required to execute the analysis.
+The risk of this approach is that the initial classification may introduce error; however, with the datasets included in our analysis, this source of error did not appear to be significant.
 
@@
 
+**Figure 1. The Opticlust algorithm.**
+A. Flowchart describing algorithm with an example (?)
+B. Plot of MCC/F1Score/Accuracy change with iterations
+C. Plot of MCC/F1Score/Accuracy for the different datasets
+
+**Figure 2.** MCC values for comparison of *de novo* clustering algorithms when applied to five natural and two synthetic datasets.
+For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.
+The median of 10 re-orderings of the data is presented for each method and dataset.
+
+**Figure 3.** F1-score values for comparison of *de novo* clustering algorithms when applied to five natural and two synthetic datasets.
+For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.
+The median of 10 re-orderings of the data is presented for each method and dataset.
+
+**Figure 4.** Accuracy values for comparison of *de novo* clustering algorithms when applied to five natural and two synthetic datasets.
+For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.
+The median of 10 re-orderings of the data is presented for each method and dataset.
+
+**Figure 5.** Demonstration of how execution time (A) and memory usage (B) scale with the number of unique sequences for each clustering algorithm.
+For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.
+
+**Figure 6.** Comparison of cluster.split and cluster for average neighbor, VSEARCH-based abundance-based greedy clustering, and Opticlust.
+
 \newpage

### 2016-08-19 submission/Westcott_Opticlust_mSystems_2016.Rmd
First draft of Materials and Methods section

@@
 ---
-title: "**Opticlust: Improved method for assigning amplicon-based sequence data to operational taxonomic units**"
+title: "**OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units**"
 bibliography: references.bib
@@
 
-***Opticlust algorithm.***
+***OptiClust algorithm (Figure 1A).*** The OptiClust algorithm uses a list of all sequence names in the dataset, the pairs of sequences that are within a desired threshold of each other (e.g.
+0.03), and the metric that should be used to assess clustering quality (e.g.
+MCC).
+The sequence pairs that are within the user-specified threshold are used to indicate those sequences that should appear within the same OTU and those that have a distance larger than the threshold should not appear within the same OTU.
+The algorithm starts by placing each sequence within its own OTU and calculating the number of true positives, true negatives, false positives, and false negatives based on the pairwise distances.
+These counts are used to calculate the metric used for optimization.
+The algorithm proceeds by interrogating each sequence and re-calculating the metric for the cases where the sequence stays in its current OTU, is moved to each of the other OTUs, or is moved into a new OTU.
+The location that results in the largest value of the optimization metric indicates whether the sequence remains in its current OTU or is moved.
+Each iteration consists of interrogating every sequence in the dataset.
+The algorithm continues until the optimization metric stabilizizes.
 
-***Opticlust generates more robust OTUs than other methods.***
+***OptiClust stopping criteria (Figure 1B).*** By default, the mothur-based implementation of the algorithm stops when the optimization metric does not change between iterations.
+It also allows the user to stop the algorithm if a maximum number of iterations is exceeded or if the change in the optimization metric is less than a stopping threshold.
+For each of the datasets we evaluated, between **XXX** and **XXX** iterations were needed to reach full conversion of the algorithm and between **XXX* and **XXX** iterations were needed to see a change in the metric that was less than 0.01%.
+To achieve full stabilization of the optimization metric, the **XXXX** algorithm required an average of **XX**% more iterations than the **XXXX** algorithm, which needed **XX**% more iterations than the **XXXX** algorithm.
+These results suggest that achieving full stabilization of the optimization metric adds computational effort, but that it is relatively small.
 
-***Opticlust is faster and uses less memory than other methods.***
+***OptiClust-generated OTUs are as stable as those from other algorithms (Figure 1C).*** One concern that many have with *de novo* clustering algorithms is that their output is sensitive to the initial order of the sequences.
+An additional concern with the OptiClust algorithm is that it may stabilize at a local optimum.
+To evaluate these concerns we compared the results obtained using ten randomizations of the order that sequences were given to the algorithm.
+Similar to our previous analysis, we found that the MCC values obtained from the replicate clusterings did not vary by more than **0.XX**% and were actually less variable than what we observed using the other algorithms.
+The variation observed in clustering quality suggests that the algorithm generates OTU assignments with subtle variation, but that the assignments are of a similar quality.
 
-***The speed and memory requirements of Opticlust scale XXXXXly.***
+***Optimization of clustering using composite metrics significantly improves clustering quality (Figure 2, 3, 4).*** There are multiple metrics available to assess clustering quality.
+The mothur-based implementation of OptiClust allows the user to use the MCC, F1 score, accuracy, sensitivity, specificity, and the sum of true positives and negatives.
+The MCC, F1 score, and accuracy are preferred because each of them incorporates all four values from the confusion matrix while others only utilize two values.
+It is relatively straightforward to implement other metrics.
+**Comparison of MCC, accuracy, and F1 score to each other for each dataset using the full set of sequences.** Each metric outperformed the observed values for the other clustering algorithms indicating that regardless of the metric one uses to evaluate clustering quality, the OptiClust algorithm generates better OTU assignments than any of the other methods.
+The number of OTUs generated by an algorithm is often used as a metric for clustering quality.
+We did not observe a significant correlation between clustering quality as measured by MCC, F1 score, or accuracy and the number of OTUs generated by each algorithm.
+**Interestingly, the values of the MCC, F1 score, and accuracy generated when using each metric to optimize clustering were similar across implementations.** Based on these results and its previous use in the OTU assignment literature, the remainder of our analysis uses the MCC metric for optimization.
+
+***Optimization-based clustering is faster and uses less memory than current methods (Figure 5).*** Although not as important as the quality of clustering, the amount of time and memory required to assign sequences to OTUs is a legitimate concern.
+**XXXX**, **XXXX**, and **XXXX** were not able to complete within 50 hours for the **XXXX** and **XXXX** datasets.
+The OptiClust algorithm finished **XXX** to **XXX**% faster than the next fastest algorithm.
+Similarly, **XXXX**, **XXXX**, and **XXXX** were not able to complete using less than 48 GB of RAM for the **XXXX** and **XXXX** datasets.
+The OptiClust algorithm used **XXX** to **XXX**% less RAM than the next fastest algorithm.
+To evaluate howt the speed and memory usage scaled with the number of sequences in the dataset, we measured the time required and maximum RAM usage to cluster 20, 40, 60, 80, and 100% of the unique sequences in each dataset by each algorithm.
+Using the full datasets, we observed that the speed and memory requirements of the OptiClust algorithm scaled **XXXXXXX**.
+These results clearly demonstrate that OptiClust is faster and uses less memory than other algorithms currently in use.
+This complements the higher quality OptiClust OTU assignments relative to other methods.
+
+***Cluster splitting heuristic generates OTUs that are as good as non-split approach (Figure 6).*** We previously described a heuristic to accelerate OTU assignments where sequences were classified to taxonomic groupings and within each taxon sequences were assigned to OTUs.
+Using the average neighbor algorithm we found that this method was effective as clustering without using the classification information.
+We again assessed this approach using the average neighbor, VSEARCH abundance-based greedy clustering, and OptiClust by splitting the data at the family level and then clustering the sequences into OTUs.
+As before, we found that the quality of the clustering was as good using the classification data as without.
+An advantage of this approach is that because each taxon is distinct, it is possible to parallelize the assignment by clustering each taxon on a separate processor.
+Although the taxa are not equally abundant, there was still a considerable improvement in execution time and the amount of memory required to execute the analysis.
+The risk of this approach is that the initial classification may introduce error; however, with the datasets included in our analysis, this source of error did not appear to be significant.
 
@@
 
-Preference for Opticlust over cluster splitting approach
+Preference for OptiClust over cluster splitting approach
 * Risks are mis-classification and artificially splitting similar sequences between OTUs because they classify to different taxa
@@
 
-## Methods
+## Materials and Methods
 
 ***Sequence data and processing steps.***
+To evaluate the OptiClust and the other algorithms we created two synthetic sequence collections and four sequence collections generated from previously published studies.
+The V4 region of the 16S rRNA gene was used from all datasets because it is a popular region that can be fully sequenced with two-fold coverage using the commonly used MiSeq sequencer from Illumina [@Kozich2013].
+The method for generating the simulated datasets followed the approach used by Kopylova et al.
+[-@Kopylova2016] and Schloss [@Schloss2016].
+Briefly, we randomly selected 10,000 uniques V4 fragments from 16S rRNA gene sequences that were unique from the SILVA non-redundant database [@Pruesse2007].
+A community with an even relative abundance profile was generated by specifying that each sequence had a frequency of 100 reads.
+A community with a staggered relative abundance profile was generated by specifying that the abundance of each sequence was a randomly drawn integer sampled from a uniform distribution between 1 and 200.
+Sequence collections collected from human feces [@Baxter2016], murine feces [@Schloss2012], soil [@Johnston2016], and seawater [@Henson2016] were used to characterize the algorithms' performance with natural communities.
+These sequence collections were all generated using paired 150 or 250 nt reads of the V4 region.
+We re-processed all of the reads using a common analysis pipeline that included quality score-based error correction [@Kozich2013], alignment against a SILVA reference database [@Pruesse2007; @Schloss2010], screening for chimeras using UCHIME [@Edgar2011], and classification using a naive Bayesian classifier with the RDP training set [@Wang2007].
 
 ***Implementation of clustering algorithms.***
+In addition to the OptiClust algorithm we evaluated ten different *de novo* clustering algorithms.
+These included three hierarchical algorithms, average neighbor (AN), nearest neighbor (NN), and furthest neighbor (FN), which are implemented in mothur [v.1.39.0; @Schloss2009].
+Seven heuristic methods were also used including abundance-based greedy clustering (UAGC) and distance-based greedy clustering (UDGC) as implemented in USEARCH [v.6.1; @Edgar2010], abundance-based greedy clustering (VAGC) and distance-based greedy clustering (VDGC) as implemented in VSEARCH [v.X.X.X; @Rognes2015], OTUClust [v.X.X.X; @XXXXXX], SumaClust [v.X.X.X; @XXXXXX], and Swarm [v.2.1.1; @Mah2014].
+With the exception of Swarm each of these methods uses distance-based thresholds to report OTU assignments.
+To judge the quality of the Swarm-generated OTU assignments we calculated the MCC value using thresholds incremented by 1% between 0 and 5% and selected the threshold that provided the optimal MCC value [@Westcott2015].
+We also assessed the ability of a previously describe heuristic to cluster sequences using the UDGC and OptiClust algorithms.
+In this heuristic sequences are split into bacterial families based on sequence classification and clustered within the taxonomic family [@Westcott20XX].
+Finally, for our benchmarking analysis, we evaluate the memory and time requirements when using 8 processors with the UDGC and VDGC algorithms and with the taxonomic splitting heuristic.
 
-***Benchmarking strategy.***
+***Benchmarking.*** We evaluated the quality of the sequence clustering, reproducibility of the clustering, the speed of clustering, and the amount of memory required to complete the clustering.
+To assess the quality of the clusters generated by each method, we counted the cells within a confusion matrix that indicated how well the clusterings represented the distances between the pair of sequences [@Schloss2011Assessing].
+Pairs of sequences that were in the same OTU and had a distance less than 3% were true positives (TPs), those that were in different OTUs and had a distance greater than 3% were true negatives (TNs), those that were in the same OTU and had a distance greater than 3% were false positives (FPs), and those that were in different OTUs and had a distance less than 3% were false negatives (FNs).
+To synthesize the matrix into a single metric we used the Matthew's Correlation Coefficient, F1 score, and accuracy using the `sens.spec` command in mothur using the following equations.
 
-## Results
+$$
+MCC = \frac{TP \times TN-FP \times FN}{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)} }
+$$
 
-***Opticlust algorithm (Figure 1A).*** The Opticlust algorithm uses a list of all sequence names in the dataset, the pairs of sequences that are within a desired threshold of each other (e.g.
-0.03), and the metric that should be used to assess clustering quality (e.g.
-MCC).
-The sequence pairs that are within the user-specified threshold are used to indicate those sequences that should appear within the same OTU and those that have a distance larger than the threshold should not appear within the same OTU.
-The algorithm starts by placing each sequence within its own OTU and calculating the number of true positives, true negatives, false positives, and false negatives based on the pairwise distances.
-These counts are used to calculate the metric used for optimization.
-The algorithm proceeds by interrogating each sequence and re-calculating the metric for the cases where the sequence stays in its current OTU, is moved to each of the other OTUs, or is moved into a new OTU.
-The location that results in the largest value of the optimization metric indicates whether the sequence remains in its current OTU or is moved.
-Each iteration consists of interrogating every sequence in the dataset.
-The algorithm continues until the optimization metric stabilizizes.
+$$
+F1_score = \frac{2 \times TP}{2 \times TP+FP+FN}
+$$
 
-***Opticlust stopping criteria (Figure 1B).*** By default, the mothur-based implementation of the algorithm stops when the optimization metric does not change between iterations.
-It also allows the user to stop the algorithm if a maximum number of iterations is exceeded or if the change in the optimization metric is less than a stopping threshold.
-For each of the datasets we evaluated, between **XXX** and **XXX** iterations were needed to reach full conversion of the algorithm and between **XXX* and **XXX** iterations were needed to see a change in the metric that was less than 0.01%.
-To achieve full stabilization of the optimization metric, the **XXXX** algorithm required an average of **XX**% more iterations than the **XXXX** algorithm, which needed **XX**% more iterations than the **XXXX** algorithm.
-These results suggest that achieving full stabilization of the optimization metric adds computational effort, but that it is relatively small.
+$$
+Accuracy = \frac{TP + TN}{(TP+FP+TN+FN)}
+$$
 
-***Opticlust-generated OTUs are as stable as those from other algorithms (Figure 1C).*** One concern that many have with *de novo* clustering algorithms is that their output is sensitive to the initial order of the sequences.
-An additional concern with the Opticlust algorithm is that it may stabilize at a local optimum.
-To evaluate these concerns we compared the results obtained using ten randomizations of the order that sequences were given to the algorithm.
-Similar to our previous analysis, we found that the MCC values obtained from the replicate clusterings did not vary by more than **0.XX**% and were actually less variable than what we observed using the other algorithms.
-The variation observed in clustering quality suggests that the algorithm generates OTU assignments with subtle variation, but that the assignments are of a similar quality.
+To assess the reproducibility of the algorithms we randomized the starting order of each sequence collection ten times and ran each algorithm on each randomized collection.
+We then measured the MCC, F1 score, and accuracy for each randomization and quantified their coefficient of variation (CV; the ratio of the standard deviation to the mean).
 
-***Optimization of clustering using composite metrics significantly improves clustering quality (Figure 2, 3, 4).*** There are multiple metrics available to assess clustering quality.
-The mothur-based implementation of Opticlust allows the user to use the MCC, F1 score, accuracy, sensitivity, specificity, and the sum of true positives and negatives.
-The MCC, F1 score, and accuracy are preferred because each of them incorporates all four values from the confusion matrix while others only utilize two values.
-It is relatively straightforward to implement other metrics.
-**Comparison of MCC, accuracy, and F1 score to each other for each dataset using the full set of sequences.** Each metric outperformed the observed values for the other clustering algorithms indicating that regardless of the metric one uses to evaluate clustering quality, the Opticlust algorithm generates better OTU assignments than any of the other methods.
-The number of OTUs generated by an algorithm is often used as a metric for clustering quality.
-We did not observe a significant correlation between clustering quality as measured by MCC, F1 score, or accuracy and the number of OTUs generated by each algorithm.
-**Interestingly, the values of the MCC, F1 score, and accuracy generated when using each metric to optimize clustering were similar across implementations.** Based on these results and its previous use in the OTU assignment literature, the remainder of our analysis uses the MCC metric for optimization.
+To assess how the the memory and time requirements scaled with the number of sequences included in each sequence collection, we randomly subsampled 20, 40, 60, or 80% of the unique sequences in each collection.
+We obtained 10 subsamples at each depth for each dataset and ran each collection (N= 50 = 5 sequencing depths x 10 replicates) through each of the algorithms.
+We used the timeout script to quantify the maximum RAM used and the amount of time required to process each sequence collection (https://github.com/pshved/timeout).
+We limited each algorithm to 48 GB of RAM, 50 hours, and unless otherwise specified, a single processor.
 
-***Optimization-based clustering is faster and uses less memory than current methods (Figure 5).*** Although not as important as the quality of clustering, the amount of time and memory required to assign sequences to OTUs is a legitimate concern.
-**XXXX**, **XXXX**, and **XXXX** were not able to complete within 50 hours for the **XXXX** and **XXXX** datasets.
-The Opticlust algorithm finished **XXX** to **XXX**% faster than the next fastest algorithm.
-Similarly, **XXXX**, **XXXX**, and **XXXX** were not able to complete using less than 48 GB of RAM for the **XXXX** and **XXXX** datasets.
-The Opticlust algorithm used **XXX** to **XXX**% less RAM than the next fastest algorithm.
-To evaluate howt the speed and memory usage scaled with the number of sequences in the dataset, we measured the time required and maximum RAM usage to cluster 20, 40, 60, 80, and 100% of the unique sequences in each dataset by each algorithm.
-Using the full datasets, we observed that the speed and memory requirements of the Opticlust algorithm scaled **XXXXXXX**.
-These results clearly demonstrate that Opticlust is faster and uses less memory than other algorithms currently in use.
-This complements the higher quality Opticlust OTU assignments relative to other methods.
-
-***Cluster splitting heuristic generates OTUs that are as good as non-split approach (Figure 6).*** We previously described a heuristic to accelerate OTU assignments where sequences were classified to taxonomic groupings and within each taxon sequences were assigned to OTUs.
-Using the average neighbor algorithm we found that this method was effective as clustering without using the classification information.
-We again assessed this approach using the average neighbor, VSEARCH abundance-based greedy clustering, and Opticlust by splitting the data at the family level and then clustering the sequences into OTUs.
-As before, we found that the quality of the clustering was as good using the classification data as without.
-An advantage of this approach is that because each taxon is distinct, it is possible to parallelize the assignment by clustering each taxon on a separate processor.
-Although the taxa are not equally abundant, there was still a considerable improvement in execution time and the amount of memory required to execute the analysis.
-The risk of this approach is that the initial classification may introduce error; however, with the datasets included in our analysis, this source of error did not appear to be significant.
+***Data and code availability.*** The workflow utilized commands in GNU make (v.3.81), GNU bash (v.4.1.2), mothur [v.1.39.0; @Schloss2009], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015].
+A reproducible version of this manuscript and analysis is available at https://github.com/SchlossLab/Westcott_OptiClust_mSystems_2015.
 
@@
 
-**Figure 1. The Opticlust algorithm.**
+**Figure 1. The OptiClust algorithm.**
 A. Flowchart describing algorithm with an example (?)
@@
 
-**Figure 6.** Comparison of cluster.split and cluster for average neighbor, VSEARCH-based abundance-based greedy clustering, and Opticlust.
+**Figure 6.** Comparison of cluster.split and cluster for average neighbor, VSEARCH-based abundance-based greedy clustering, and OptiClust.
 

### 2016-09-27 code/build_figure2.R
Add seeding figure adn change figure names

@@ -1,92 +1,0 @@
-library(dplyr)
-library(ggplot2)
-library(cowplot)
-library(wesanderson)
-
-basic_methods <- c('an', 'fn',
-									'nn', 'otuclust', 'sumaclust', 'swarm',
-									'uagc', 'udgc', 'vagc_1', 'vdgc_1', 'mcc')
-pretty_methods <- c('Average\nNeighbor', 'Furthest\nNeighbor',
- 										'Nearest\nNeighbor', 'OTUCLUST', 'Sumaclust', 'Swarm',
-										'AGC with\nUSEARCH', 'DGC with\nUSEARCH',
-										'AGC with\nVSEARCH', 'DGC with\nVSEARCH', 'OptiClust')
-names(pretty_methods) <- basic_methods
-
-datasets <- c('soil', 'marine', 'mice', 'human', 'even', 'staggered')
-pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
-names(pretty_datasets) <- datasets
-
-data <- read.table(file="data/processed/cluster_data.summary", header=T)
-
-subset <- data %>%
-			filter(frac==1.0 & method %in% basic_methods) %>%
-			group_by(dataset, method) %>%
-			summarize(avg_mcc=mean(mcc, na.rm=T),
-								sd_mcc=sd(mcc, na.rm=T),
-								avg_sobs=mean(num_otus, na.rm=T),
-								sd_sobs=sd(num_otus, na.rm=T),
-								n=sum(!is.na(mcc))) %>%
-			mutate(cov_mcc=sd_mcc/avg_mcc)
-			
-method_ordering <- data %>%
-			filter(frac==1.0 & method %in% basic_methods) %>%
-			group_by(method) %>%
-			summarize(avg_mcc=mean(mcc, na.rm=T)) %>%
-			arrange(desc(avg_mcc))
-
-subset$method <- factor(subset$method, method_ordering$method)
-subset$dataset <- factor(subset$dataset, datasets)
-
-my_theme <- theme_classic() +
-	theme(axis.text.x =element_blank(),
-		axis.text.y=element_text(size=8),
-		axis.ticks.x = element_blank(),
-		axis.title.y=element_text(size=8),
-		panel.grid.major = element_blank(),
-		panel.grid.minor = element_blank(),
-		panel.border = element_rect(color = "black", fill=NA, size=1),
[…]

### 2016-09-27 code/build_performance_figure.R

@@ -1,0 +1,92 @@
+library(dplyr)
+library(ggplot2)
+library(cowplot)
+library(wesanderson)
+
+basic_methods <- c('an', 'fn',
+									'nn', 'otuclust', 'sumaclust', 'swarm',
+									'uagc', 'udgc', 'vagc_1', 'vdgc_1', 'mcc')
+pretty_methods <- c('Average\nNeighbor', 'Furthest\nNeighbor',
+ 										'Nearest\nNeighbor', 'OTUCLUST', 'Sumaclust', 'Swarm',
+										'AGC with\nUSEARCH', 'DGC with\nUSEARCH',
+										'AGC with\nVSEARCH', 'DGC with\nVSEARCH', 'OptiClust')
+names(pretty_methods) <- basic_methods
+
+datasets <- c('soil', 'marine', 'mice', 'human', 'even', 'staggered')
+pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
+names(pretty_datasets) <- datasets
+
+data <- read.table(file="data/processed/cluster_data.summary", header=T)
+
+subset <- data %>%
+			filter(frac==1.0 & method %in% basic_methods) %>%
+			group_by(dataset, method) %>%
+			summarize(avg_mcc=mean(mcc, na.rm=T),
+								sd_mcc=sd(mcc, na.rm=T),
+								avg_sobs=mean(num_otus, na.rm=T),
+								sd_sobs=sd(num_otus, na.rm=T),
+								n=sum(!is.na(mcc))) %>%
+			mutate(cov_mcc=sd_mcc/avg_mcc)
+
+method_ordering <- data %>%
+			filter(frac==1.0 & method %in% basic_methods) %>%
+			group_by(method) %>%
+			summarize(avg_mcc=mean(mcc, na.rm=T)) %>%
+			arrange(desc(avg_mcc))
+
+subset$method <- factor(subset$method, method_ordering$method)
+subset$dataset <- factor(subset$dataset, datasets)
+
+my_theme <- theme_classic() +
+	theme(axis.text.x =element_blank(),
+		axis.text.y=element_text(size=8),
+		axis.ticks.x = element_blank(),
+		axis.title.y=element_text(size=8),
+		panel.grid.major = element_blank(),
+		panel.grid.minor = element_blank(),
+		panel.border = element_rect(color = "black", fill=NA, size=1),
[…]

### 2016-09-27 code/build_seeding_figure.R

@@ -1,0 +1,53 @@
+library(dplyr)
+library(ggplot2)
+library(tidyr)
+library(wesanderson)
+
+data <- read.table(file="data/processed/cluster_data.summary", header=TRUE, stringsAsFactors=FALSE)
+
+mcc_methods <- c("mcc", "mcc_agg")
+pretty_methods <- c("Separate OTUs", "Single OTU")
+
+datasets <- c('soil', 'marine', 'mice', 'human', 'even', 'staggered')
+pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
+names(pretty_datasets) <- datasets
+
+agg_test <- data %>%
+			filter(frac==1.0 & method %in% mcc_methods) %>%
+			group_by(dataset, method) %>%
+			summarize(avg_mcc=mean(mcc, na.rm=T),
+								avg_secs=mean(cluster_secs, na.rm=T))
+agg_test$dataset <- factor(agg_test$dataset, datasets)
+
+my_theme <- theme_classic() +
+	theme(
+		panel.grid.major = element_blank(),
+		panel.grid.minor = element_blank(),
+		panel.border = element_rect(color = "black", fill=NA, size=1),
+		panel.background = element_rect(fill=NA),
+		strip.background = element_blank(),
+		strip.text.y = element_text(size = 11),
+		axis.text.x=element_text(size=10),
+		legend.position=c(.75, .7),
+		legend.title=element_text(size=9),
+		legend.text = element_text(size=9),
+		legend.key.size = unit(0.75, "line")
+	)
+
+
+gathered <- gather(agg_test, parameter, variable, avg_mcc, avg_secs)
+
+facet_labels <- as_labeller(c('avg_mcc' = "Mean Matthew's\nCorrelation Coefficient", 'avg_secs' = "Execution Time\n(seconds)"))
+
+ggplot(gathered, aes(dataset, variable, col=dataset, shape=method)) +
+	geom_point(position = position_dodge(0.1), size=2) +
+	facet_grid(parameter~., scales = 'free_y', switch = 'y', labeller = facet_labels) +
+	scale_x_discrete(breaks=levels(agg_test$dataset),
+		labels=pretty_datasets[levels(agg_test$dataset)]) +
+	scale_color_manual(breaks=datasets, labels=pretty_datasets, values=
+		c('black', wes_palette("Darjeeling")), guide=F)+
[…]

### 2016-09-27 code/aggregate_steps.R
Add code to count number of steps required for MCC

@@ -1,0 +1,57 @@
+#this function concatenates together the dataset, fraction, replicate number,
+#and method to generate the filename stub
+cat_cols_files <- function(x){
+	x["frac"] <- gsub("\\.", "_", x["frac"])
+	x["frac"] <- gsub("^1$", "1_0", x["frac"])
+	x["rep"] <- gsub("^\\s*(\\d)$", "0\\1", x["rep"])
+	paste(x["dataset"], x["frac"], x["rep"], x["method"], sep='.')
+}
+
+
+#takes file name stub from results and returns the steps files
+build_steps_files <- function(x){
+	seed <- cat_cols_files(x)
+	paste0("data/", x["dataset"], "/", seed, ".steps")
+}
+
+
+#return the total number of steps to reach point where change is less than #0.01% or to convergence (0.00% change)
+get_required_steps <- function(file){
+	steps <- read.table(file=file, header=T)
+
+	mcc_a <- steps[-nrow(steps),"mcc"]
+	mcc_b <- steps[-1, "mcc"]
+
+	c(min(steps$iter[(mcc_b - mcc_a) / mcc_a < 0.0001]), max(steps$iter))
+}
+
+# Six datasets - two synthetic and four biological
+datasets <- c("even", "human", "marine", "mice", "soil", "staggered")
+n_datasets <- length(datasets)
+
+# these are the fractions of *unique* sequences that were used for each dataset
+fractions <- 1#seq(0.2,1,0.2)
+n_fracs <- length(fractions)
+
+# we randomized the sequences (and their order) in each fraction ten times
+replicates <- 1:10
+n_reps <- length(replicates)
+
+# focus on the mcc method
+methods <- c("mcc", "mcc_agg")
+n_methods <- length(methods)
+
+dataset_vector <- rep(datasets, each=n_reps*n_fracs*n_methods)
+frac_vector <- rep(fractions, each=n_reps*n_methods, times=n_datasets)
+rep_vector <- rep(replicates, times=n_datasets*n_fracs*n_methods)
+method_vector <- rep(methods, each=n_reps, times=n_fracs*n_datasets)
+
+results <- data.frame(dataset=dataset_vector, frac=frac_vector, rep=rep_vector, method=method_vector, stringsAsFactors=F)
+
+steps_names <- apply(results, 1, build_steps_files)
+all_stats <- t(sapply(steps_names, get_required_steps))
+results$close_enough <- all_stats[,1] #within 0.01%
[…]

### 2016-09-27 code/build_seeding_figure.R
Fix variable names to be more meaningful

@@ -3,7 +3,7 @@ library(ggplot2)
 library(tidyr)
 library(wesanderson)
 
-data <- read.table(file="data/processed/cluster_data.summary", header=TRUE, stringsAsFactors=FALSE)
+cluster_data <- read.table(file="data/processed/cluster_data.summary", header=TRUE, stringsAsFactors=FALSE)
 
 mcc_methods <- c("mcc", "mcc_agg")
 pretty_methods <- c("Separate OTUs", "Single OTU")
@@ -12,12 +12,12 @@ datasets <- c('soil', 'marine', 'mice', 'human', 'even', 'staggered')
 pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
 names(pretty_datasets) <- datasets
 
-agg_test <- data %>%
+cluster <- cluster_data %>%
 			filter(frac==1.0 & method %in% mcc_methods) %>%
 			group_by(dataset, method) %>%
 			summarize(avg_mcc=mean(mcc, na.rm=T),
 								avg_secs=mean(cluster_secs, na.rm=T))
-agg_test$dataset <- factor(agg_test$dataset, datasets)
+cluster$dataset <- factor(cluster$dataset, datasets)
 
 my_theme <- theme_classic() +
 	theme(
@@ -35,7 +35,7 @@ my_theme <- theme_classic() +
 	)
 
 
-gathered <- gather(agg_test, parameter, variable, avg_mcc, avg_secs)
+gathered <- gather(cluster, parameter, variable, avg_mcc, avg_secs)
 
 facet_labels <- as_labeller(c('avg_mcc' = "Mean Matthew's\nCorrelation Coefficient", 'avg_secs' = "Execution Time\n(seconds)"))
 

### 2016-09-28 submission/supplemental_text.Rmd
Add detailed description of algorithm

---
title: "Supplemental Text"
output:
  pdf_document:
    latex_engine: xelatex
mainfont: Arial
fontsize: 12pt
geometry: margin=1.0in
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
library(knitr, quiet=T)
library(diagram, quiet=T)

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)
opts_chunk$set("fig.height" = 3)
opts_chunk$set("fig.align" = 'center')

par(mar=c(2,2,2,2), xpd=T)

mcc <- function(tp, tn, fp, fn){
	format(round((tp * tn - fp * fn) /	sqrt((tp+fp)*(tp+fn)*(tn+fp)*(tn+fn)),digits=2), digits=2, nsmall=2L)
}

#initially
tp <- 0
tn <- 1210
fp <- 0
fn <- 15
```

To provide a detailed demonstration of how the OptiClust algorithm works consider a relatively simple example where there are 50 sequences. After aligning the sequences and calculating the pairwise distances between the sequences there are 15 pairs of sequences with a distance below the desired threshold of 0.03:

```{r distances}
data <- matrix(c("D", "B", 0.02381,
								"F", "E", 0.02767,
								"G", "C", 0.02756,
								"H", "A", 0.02745,
								"I", "B", 0.01581,
								"I", "D", 0.02372,
								"J", "A", 0.02756,
								"J", "H", 0.02362,
								"N", "M", 0.02390,
								"O", "L", 0.02353,
								"P", "K", 0.01587,
								"P", "L", 0.02745,
								"P", "O", 0.02745,
								"Q", "E", 0.02390,
								"Q", "F", 0.02767), nrow=15, byrow=T)
data.df <- data.frame(data, stringsAsFactors=F)
colnames(data.df) <- c("Seq. 1", "Seq. 2", "Distance")
data.df$Distance <- round(as.numeric(data.df$Distance), 3)
kable(data.df, align='c')
```

The other 1210 distances were larger than 0.03 and are not needed for the analysis because they are taken to be true negatives. For example, the distance between sequences A and B is larger than 0.03. So, if they are in different OTUs then that would be a true negative (TN) and if they are in the same OTU then that would be a false positive (FP). Alternatively, because sequences B and D are closer to each other than 0.03 (i.e. 0.024) if they are in separate OTUs then that would be a false negative (FN) and if they are in the same OTU then that would be a true positive (TP). It is important to note that the algorithm assumes that there are no duplicate sequences and the actual abundance is saved elsewhere to be substituted later when counting the frequency distribution of each OTU across the samples included in the analysis.

The algorithm starts by seeding sequences either into individual OTUs or into a single OTU. As demonstrated in Figure 1, seeding the sequences into randomly ordered individual OTUs generates better results and is faster than starting with a single OTU. Among the 15 pairwise distances that are smaller than 0.03, there are 17 sequences that are labeled A through Q. A separate pool is created for the 33 other sequences that are not within 0.03 of any other sequence and are thus to be placed into 33 separate OTUs. In the diagrams below, this pool is designated as "...". Having seeded the initial OTUs there are `r tp` TPs, `r tn` TNs, `r fp` FPs, and `r fn` FNs. Initially the number of FNs corresponds to the number of distances less than 0.03, the number of TNs is the number of total distances (i.e. `r (50*49)/2`) minus the number of distances less than 0.03. The number of TPs, TNs, FPs, and FNs should sum to the total number of distances. The resulting Matthew's Correlation Coefficient (MCC) is 0.00. The algorithm next goes through each sequence sequentially to determine whether the MCC value would be increased by removing it from its current OTU to join other sequences in a new OTU or to create its own OTU.

----

The demonstration of the algorithm starts with sequence A. Notice that it is within 0.03 of sequences H and J. There are three options: sequence A could remain as its own OTU, it could join with sequence H, or it could join with sequence J.

```{r A}

names <- c(	"A", "B", "C", "D", "E", "F", "G", "H", "I", "J", "K", "L", "M", "N", "O", "P", "Q", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names
M["H","A"] <- mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)
M["J","A"] <- mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)
C["H","A"] <- 0.3
C["J","A"] <- -0.3

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.025, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.8, self.shiftx =-0.02, self.shifty=-0.03, xpd=T, box.cex=0.6, arr.length=0.3)

tp <- tp+1
fn <- fn-1
```

If it stays as its own OTU, the MCC value would remain 0.00. If it joined with sequences H or J the number of TPs would increase by 1 and the number of FNs would decrease by 1. In either case the MCC would be `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`. In this case, joining H or J would result in an improved MCC and so the algorithm randomly selects which sequences to join. For this demonstration, it will form a new OTU with sequence H. This results in `r tp` TPs, `r tn` TNs, `r fp` FPs, `r fn` FNs, and an MCC of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`.

----

Sequence B is processed by the same process as sequence A. Sequence B is within 0.03 of sequences D and I. Again, there are three options: sequence B could remain as its own OTU, it could join with sequence D, or it could join with sequence I.

```{r B}
names <- c(	"B", "C", "D", "E", "F", "G", "A,H", "I", "J", "K", "L", "M", "N", "O", "P", "Q", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

#M[1,1] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["B","B"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["D", "B"] <- mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)
M["I", "B"] <- mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)
C["D", "B"] <- 1
C["I", "B"] <- -0.3

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.025, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.8, self.shiftx =-0.02, self.shifty=-0.03, xpd=T, box.cex=0.6, arr.length=0.3)

tp <- tp+1
fn <- fn-1
```

If it remains as its own OTU, the MCC value would remain `r mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)`. If it joined with sequences D or I the number of TPs would increase by 1 and the number of FNs would decrease by 1. In either case the MCC would be `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`. Joining D or I would result in an improved MCC and so the algorithm randomly selects which sequences to join. For this demonstration it will form a new OTU with sequence D. This results in `r tp` TPs, `r tn` TNs, `r fp` FPs, `r fn` FNs, and an MCC of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`.

----

Sequence C is within 0.03 of sequence G creating two options: remain as its own OTU or join with sequence G.

```{r C}
names <- c(	"C", "B,D", "E", "F", "G", "A,H", "I", "J", "K", "L", "M", "N", "O", "P", "Q", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

#M[1,1] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["C","C"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["G", "C"] <- mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)
C["G", "C"] <- -0.5

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.025, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.8, self.shiftx =-0.02, self.shifty=-0.03, xpd=T, box.cex=0.6, arr.length=0.3)

tp <- tp+1
fn <- fn-1
```

If it remains in its own OTU, the MCC value would remain `r mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)`. If it joined with sequence G the number of TPs would increase by 1 and the number of FNs would decrease by 1 resulting in an MCC value of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`. Because of the improved MCC value, sequence C joins with sequence G to form a new OTU. This results in `r tp` TPs, `r tn` TNs, `r fp` FPs, `r fn` FNs, and an MCC of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`.

----

For sequence D the algorithm gets more complicated since sequence D is already in an OTU with sequence B. As seen when considering sequence B, sequence D is within 0.03 of sequence B and it is also similar to sequence I. Now there are three options: sequence D could remain with sequence B in an OTU, it could leave that OTU and join with sequence I, or it could form a new OTU where it is the sole member.

```{r D}
names <- c(	"B,D", "E", "F", "C,G", "A,H", "I", "J", "K", "L", "M", "N", "O", "P", "Q", "...", "New")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["B,D","B,D"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["I", "B,D"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["New", "B,D"] <- mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)
C["I", "B,D"] <- -0.5
C["New", "B,D"] <- 0.2

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.025, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.8, self.shiftx =-0.02, self.shifty=-0.03, xpd=T, box.cex=0.6, arr.length=0.3)
```

If sequence D remains with B, the MCC value would remain `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`. If it joined with sequence I the number of TPs and FNs would stay constant resulting in an MCC value of `r mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)`. If it formed a new OTU by itself, the number of TPs would decrease by one and the number of FNs would increase by 1 resulting in an MCC value of `r mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)`. Because the MCC values for staying in the OTU with B or leaving the OTU to join the OTU with I are the same, the algorithm would again randomly chose between the two options. For demonstration purposes, sequence D will remain in its OTU with sequence B. This results in no changes in the four parameters or the MCC value.

----

For sequence E, the same type of options are available as when sequences A and B were processed. Sequence E could remain on its own or it could join with sequences F or Q.

```{r E}
names <- c(	"B,D", "E", "F", "C,G", "A,H", "I", "J", "K", "L", "M", "N", "O", "P", "Q", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["E","E"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["F", "E"] <- mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)
M["Q", "E"] <- mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)
C["F", "E"] <- 1.5
C["Q", "E"] <- -0.2

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.025, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.8, self.shiftx =-0.02, self.shifty=-0.03, xpd=T, box.cex=0.6, arr.length=0.3)

tp <- tp+1
fn <- fn-1
```

Similar to the earlier cases, the MCC value for joining another sequence is larger than staying on its own. Because the MCC values for joining F or Q are the same, the algorithm randomly selects which sequences to join. For this demonstration sequence E will form a new OTU with sequence F. This results in `r tp` TPs, `r tn` TNs, `r fp` FPs, `r fn` FNs, and an MCC of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`.

---

For sequence F, the steps taken by the algorithm are the same as earlier for sequence D. Here, sequence F could remain in an OTU with sequence E, it could leave and form an OTU with sequence Q, or it could form a new OTU on its own.

```{r F}
names <- c(	"B,D", "E,F", "C,G", "A,H", "I", "J", "K", "L", "M", "N", "O", "P", "Q", "...", "New")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["E,F","E,F"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["Q", "E,F"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["New", "E,F"] <- mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)
C["Q", "E,F"] <- -0.2
C["New", "E,F"] <- 0.2

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.025, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.8, self.shiftx =-0.02, self.shifty=-0.03, xpd=T, box.cex=0.6, arr.length=0.3)
```

Again, the MCC value for sequence F remaining with sequence E is the same as for leaving to form an OTU with sequence Q. Both options are superior to leaving to form a new OTU on its own. The algorithm randomly chooses between the two options. For demonstration purposes, sequence F will remain in its OTU with sequence E. This results in no changes in the four parameters or the MCC value.

---

The decisions for sequence G are similar to sequence F, with the exception that the only choices are to stay in an OTU with another sequence (C) or to form a new OTU on its own.

```{r G}
names <- c(	"B,D", "E,F", "C,G", "A,H", "I", "J", "K", "L", "M", "N", "O", "P", "Q", "...", "New")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["C,G","C,G"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["New","C,G"] <- mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)
C["New","C,G"] <- -0.2

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.025, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.8, self.shiftx =-0.02, self.shifty=-0.03, xpd=T, box.cex=0.6, arr.length=0.3)
```

Leaving to form a new OTU results in a lower MCC value and so the algorithm leaves sequence G with sequence C. This results in no changes in the four parameters or the MCC value.

---

For sequence H the steps taken are the same as seen earlier for sequence B.

```{r H}
names <- c(	"B,D", "E,F", "C,G", "A,H", "I", "J", "K", "L", "M", "N", "O", "P", "Q", "...", "New")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["A,H","A,H"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["J", "A,H"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["New", "A,H"] <- mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)
C["J", "A,H"] <- -0.9
C["New", "A,H"] <- 0.2

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.025, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.8, self.shiftx =-0.02, self.shifty=-0.03, xpd=T, box.cex=0.6, arr.length=0.3)
```

For demonstration purposes sequence H will remain in its OTU with sequence A.

---

Moving to sequence I, the process is similar to what was done earlier with sequences A and B. The only difference is that because sequence I is similar to both sequences B and D the increase in TP and decrease in FN will be double.

```{r I}
names <- c(	"B,D", "E,F", "C,G", "A,H", "I", "J", "K", "L", "M", "N", "O", "P", "Q", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["I","I"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["B,D", "I"] <- mcc(tp=tp+2, tn=tn, fp=fp, fn=fn-2)
C["B,D", "I"] <- 0.4

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.025, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.8, self.shiftx =-0.02, self.shifty=-0.03, xpd=T, box.cex=0.6, arr.length=0.3)

tp <- tp+2
fn <- fn-2
```

If sequence I remains in an OTU by itself, the MCC value will be `r mcc(tp=tp-2, tn=tn, fp=fp, fn=fn+2)`. If it joined the OTU with sequences B and D, then the number of TPs would increase by 2 and the number of FNs would decrease by 2 resulting in an improved MCC value of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`. This is the choice that is taken resulting in `r tp` TPs, `r tn` TNs, `r fp` FPs, and `r fn` FNs.

---

Processing sequence J is the same as sequence I since sequence J is close to both A and H.

```{r J}
names <- c(	"B,D,I", "E,F", "C,G", "A,H", "J", "K", "L", "M", "N", "O", "P", "Q", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["J","J"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["A,H", "J"] <- mcc(tp=tp+2, tn=tn, fp=fp, fn=fn-2)
C["A,H", "J"] <- 1.5

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.03, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.8, self.shiftx =-0.02, self.shifty=-0.04, xpd=T, box.cex=0.6, arr.length=0.3)

tp <- tp+2
fn <- fn-2
```

Again, by joining the OTU containing sequences A and H the number of TPs increases by 2 and the number of FNs decreases by 2 resulting in `r tp` TPs, `r tn` TNs, `r fp` FPs, `r fn` FNs, and an improved MCC value of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`.

---

Processing sequence K is the same as for sequence A.

```{r K}
names <- c(	"B,D,I", "E,F", "C,G", "A,H,J", "K", "L", "M", "N", "O", "P", "Q", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["K","K"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["P", "K"] <- mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)
C["P", "K"] <- 0.4

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.03, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.8, self.shiftx =-0.02, self.shifty=-0.04, xpd=T, box.cex=0.6, arr.length=0.3)

tp <- tp+1
fn <- fn-1
```

Merging sequences K and P into the same OTU results in `r tp` TPs, `r tn` TNs, `r fp` FPs, `r fn` FNs, and an improved MCC value of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`.

---

Processing sequence L presents a more complicated set of decisions. Again, there are three choices. Because sequence L is similar to sequences O and P it could form an OTU with sequence O or with the OTU containing sequences K and P.

```{r L}
names <- c(	"B,D,I", "E,F", "C,G", "A,H,J", "L", "M", "N", "O", "K,P", "Q", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["L","L"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["O", "L"] <- mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)
M["K,P", "L"] <- mcc(tp=tp+1, tn=tn-1, fp=fp+1, fn=fn-1)
C["O", "L"] <- -0.4
C["K,P", "L"] <- 0.4

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.03, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.5, self.shiftx =-0.02, self.shifty=-0.04, xpd=T, box.cex=0.6, arr.length=0.3)

tp <- tp+1
fn <- fn-1
```

If sequence L remains on its own, the MCC value would remain `r mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)`. If it joined with sequence O the number of TPs would increase by 1 and the number of FNs would decrease by 1 resulting in an MCC value of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`. The subtlety of this step is found in when considering the possibility of sequence L joining an OTU with sequences K and P. It would increase the number of TPs by one and decrease the number of FN by one and by joining with sequence P; however, because O is not close to K the number of TNs would decrease by one and the number of FPs would increase by one. This would result in an MCC value of `r mcc(tp=tp, tn=tn-1, fp=fp+1, fn=fn)`. Of the three options forming an OTU with sequences L and O provides the maximal MCC value. This results in `r tp` TPs, `r tn` TNs, `r fp` FPs, `r fn` FNs, and an improved MCC value of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`.

---

The steps for processing sequence M is the same as earlier for sequence C.

```{r M}
names <- c(	"B,D,I", "E,F", "C,G", "A,H,J", "M", "N", "L,O", "K,P", "Q", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["M","M"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["N", "M"] <- mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)
C["N", "M"] <- -1.0

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.03, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.5, self.shiftx =-0.02, self.shifty=-0.04, xpd=T, box.cex=0.6, arr.length=0.3)

tp <- tp+1
fn <- fn-1
```

Merging sequences M and N into the same OTU results in `r tp` TPs, `r tn` TNs, `r fp` FPs, `r fn` FNs, and an improved MCC value of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`.

---

Moving on to sequence N, the two options are to stay in an OTU with sequence M or to spit off and form a new OTU on its own.

```{r N}
names <- c(	"B,D,I", "E,F", "C,G", "A,H,J", "M,N", "L,O", "K,P", "Q", "...", "New")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["M,N","M,N"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["New","M,N"] <- mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)
C["New","M,N"] <- 0.3

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.03, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.5, self.shiftx =-0.02, self.shifty=-0.04, xpd=T, box.cex=0.6, arr.length=0.3)
```

Remaining in an OTU with sequence M provides the larger MCC value and so the OTU memberships do not change.

---

Processing of sequence O presents three options that have been explored before. Sequence O can stay in its OTU with sequence L, it can join the OTU with sequences K and P, or it can form a new OTU on its own.

```{r O}
names <- c(	"B,D,I", "E,F", "C,G", "A,H,J", "M,N", "L,O", "K,P", "Q", "...", "New")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["L,O","L,O"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["K,P", "L,O"] <- mcc(tp=tp, tn=tn-1, fp=fp+1, fn=fn)
M["New","L,O"] <- mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)
C["K,P", "L,O"] <- -1.0
C["New", "L,O"] <- 0.3

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.03, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.5, self.shiftx =-0.02, self.shifty=-0.04, xpd=T, box.cex=0.6, arr.length=0.3)
```

If sequence O remains in the OTU with sequence L, the MCC value would be `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`. If it leaves that OTU to join sequences K and P in their OTU then the number of TNs would decrease by 1, but the number of FPs would increase by 1 because O is similar to P, but not to K. This would result in an MCC value of `r mcc(tp=tp, tn=tn-1, fp=fp+1, fn=fn)`. If sequence O forms a new OTU on its own, then the number of TPs would decrease by one and the number of FNs would increase by one resulting in an MCC value of `r mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)`. The best option is for sequence O to remain in its OTU with sequence L.

---

For sequence P the steps taken are similar to those used to evaluate clusters for sequence O; however, the final decision is different. Sequence P can stay with sequence K in their OTU, it can leave to join the OTU with sequences L and O, or it can form a new OTU on its own.

```{r P}
names <- c(	"B,D,I", "E,F", "C,G", "A,H,J", "M,N", "L,O", "K,P", "Q", "...", "New")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["K,P","K,P"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["L,O", "K,P"] <- mcc(tp=tp+1, tn=tn, fp=fp, fn=fn-1)
M["New", "K,P"] <- mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)
C["L,O", "K,P"] <- 0.7
C["New", "K,P"] <- -0.4

par(mar=c(1,2,1,2), xpd=T)

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.03, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.5, self.shiftx =-0.02, self.shifty=-0.04, xpd=T, box.cex=0.6, arr.length=0.3)

tp <- tp+1
fn <- fn-1
```

If sequence P remains in the OTU with sequence K, the MCC value would be `r mcc(tp=tp-1, tn=tn, fp=fp, fn=fn+1)`. Alternatively, P could leave that OTU to join sequences L and O in their OTU. If P leaves its OTU with K then the number of TPs would decrease by one and the number of FNs would increase by one. By joining with L and O the number of TPs would increase by two and the number of FNs would decrease by two. The net effect would be to increase the number of TPs by 1 and decrease the number of FNs by 1. This would result in an MCC value of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`. If sequence P formed a new OTU on its own, then the number of TPs would decrease by one and the number of FNs would increase by one resulting in an MCC value of `r mcc(tp=tp-2, tn=tn, fp=fp, fn=fn+2)`. The best option is for sequence P to leave its OTU with sequence K and join the OTU containing sequences L and O. The updated counts are `r tp` TPs, `r tn` TNs, `r fp` FPs, `r fn` FNs.

---

To finish the first round of processing each sequence, sequence Q is processed. Sequence Q is similar to both sequences E and F. Because sequences E and F are in the same OTU, the situation is similar to processing sequence I.

```{r Q}
names <- c(	"B,D,I", "E,F", "C,G", "A,H,J", "M,N", "L,O,P", "K", "Q", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["Q","Q"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["E,F", "Q"] <- mcc(tp=tp+2, tn=tn, fp=fp, fn=fn-2)
C["E,F", "Q"] <- 0.2

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.03, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.5, self.shiftx =-0.02, self.shifty=-0.04, xpd=T, box.cex=0.6, arr.length=0.3)

tp <- tp+2
fn <- fn-2
```

By joining the OTU containing sequences E and F the number of TPs increases by two and the number of FNs decreases by one. The updated counts are `r tp` TPs, `r tn` TNs, `r fp` FPs, and `r fn` FNs, which result in an improved MCC value of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`.

---

Having processed each sequence, the first iteration of the algorithm is complete. The MCC value has changed from 0.00 to `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`. Because the MCC value changed, it is necessary to re-evaluate each sequence again and re-evaluate the final MCC value to determine whether it has changed. In this case, evaluation of sequences A through J result in the same clustering pattern. When the algorithm reaches sequence K it finds that the sequence is similar to sequence P, which is in an OTU with L and O; however sequence K is not similar to L or O.

```{r K2}
names <- c(	"B,D,I", "E,F,Q", "C,G", "A,H,J", "M,N", "L,O,P", "K", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["K","K"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["L,O,P", "K"] <- mcc(tp=tp+1, tn=tn-2, fp=fp+2, fn=fn-1)
C["L,O,P", "K"] <- 0.4

par(mar=c(1,2,1,2), xpd=T)

plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, cex.txt = 0.6, box.size = 0.03, box.type = "circle", shadow.size=0, arr.type="triangle", dtext=0.5, self.shiftx =-0.02, self.shifty=-0.04, xpd=T, box.cex=0.6, arr.length=0.3)

```

Although, sequence K is similar to sequence P, it is not similar to sequences L or O. Were sequence K to join their OTU, it would increase the number of TPs by one and decrease the number of FNs by one because of its similarity to sequence P, but it would increase the number of FPs by two and decrease the number of TNs by two because K is not similar to L or O. The end result would be a MCC value of `r mcc(tp=tp+1, tn=tn-2, fp=fp+2, fn=fn-1)`, which is less than the MCC value of keeping sequence K on its own (i.e. `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`).

---

Continuing the process for the remaining sequences, none of the sequences will move between OTUs and the MCC value does not change. At this point, the clustering has converged to the optimum MCC value of `r mcc(tp=tp, tn=tn, fp=fp, fn=fn)`. Repeating this process using 100 different seeds for the random number generator required a median of 3 iterations (range from 2 to 4) to converge.

### 2016-09-28 code/build_seeding_figure.R
Correct error in where the sample names were taken from

@@ -42,8 +42,8 @@ facet_labels <- as_labeller(c('avg_mcc' = "Mean Matthew's\nCorrelation Coefficie
 ggplot(gathered, aes(dataset, variable, col=dataset, shape=method)) +
 	geom_point(position = position_dodge(0.1), size=2) +
 	facet_grid(parameter~., scales = 'free_y', switch = 'y', labeller = facet_labels) +
-	scale_x_discrete(breaks=levels(agg_test$dataset),
-		labels=pretty_datasets[levels(agg_test$dataset)]) +
+	scale_x_discrete(breaks=levels(gathered$dataset),
+		labels=pretty_datasets[levels(gathered$dataset)]) +
 	scale_color_manual(breaks=datasets, labels=pretty_datasets, values=
 		c('black', wes_palette("Darjeeling")), guide=F)+
 	scale_shape_manual(name="Sequences initialized as", values=c(19,21), labels=pretty_methods) +

### 2016-09-28 submission/Westcott_Opticlust_mSystems_2016.Rmd
Work on figure legends

@@
 The sequence pairs that are within the user-specified threshold are used to indicate those sequences that should appear within the same OTU and those that have a distance larger than the threshold should not appear within the same OTU.
-The algorithm starts by placing each sequence within its own OTU and calculating the number of true positives, true negatives, false positives, and false negatives based on the pairwise distances.
+The algorithm starts by placing each sequence either within its own OTU or into a single OTU.
+It then calculates the number of true positives, true negatives, false positives, and false negatives based on the pairwise distances.
 These counts are used to calculate the metric used for optimization.
@@
 Each iteration consists of interrogating every sequence in the dataset.
-The algorithm continues until the optimization metric stabilizizes.
+The algorithm continues until the optimization metric stabilizes.
 
@@
 
-**Figure 1. The OptiClust algorithm.**
-A. Flowchart describing algorithm with an example (?)
-B. Plot of MCC/F1Score/Accuracy change with iterations
-C. Plot of MCC/F1Score/Accuracy for the different datasets
+**Figure 1. OptiClust performance.**
+Plot of MCC (A) and execution times (B) for the different starting conditions
+(C) A Number of steps required to converge with different thresholds
 
-**Figure 2.** MCC values for comparison of *de novo* clustering algorithms when applied to five natural and two synthetic datasets.
+**Figure 2.** MCC values (A) and their coefficient of variation (B) and the number of observed OTUs (C) for comparison of *de novo* clustering algorithms when applied to four natural and two synthetic datasets.
 For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.
@@
 
-**Figure 3.** F1-score values for comparison of *de novo* clustering algorithms when applied to five natural and two synthetic datasets.
-For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.
-The median of 10 re-orderings of the data is presented for each method and dataset.
-
-**Figure 4.** Accuracy values for comparison of *de novo* clustering algorithms when applied to five natural and two synthetic datasets.
-For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.
-The median of 10 re-orderings of the data is presented for each method and dataset.
-
-**Figure 5.** Demonstration of how execution time (A) and memory usage (B) scale with the number of unique sequences for each clustering algorithm.
+**Figure 3.** Demonstration of how execution time (A) and memory usage (B) scale with the number of unique sequences for each clustering algorithm.
 For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.
 
-**Figure 6.** Comparison of cluster.split and cluster for average neighbor, VSEARCH-based abundance-based greedy clustering, and OptiClust.
+**Figure 4.** Comparison of cluster.split and cluster for average neighbor, VSEARCH-based abundance-based greedy clustering, and OptiClust.
+
+**Supplemental text.** Worked example of how OptiClust algorithm clusters sequences into OTUs.
 

### 2016-09-29 code/build_algorithm_figure.R
Finalize algorithm evaluation figure

@@ -1,0 +1,119 @@
+library(dplyr)
+library(ggplot2)
+library(tidyr)
+library(wesanderson)
+library(cowplot)
+
+
+my_theme <- theme_classic() +
+	theme(
+		panel.grid.major = element_blank(),
+		panel.grid.minor = element_blank(),
+		panel.border = element_rect(color = "black", fill=NA, size=1),
+		panel.background = element_rect(fill=NA),
+#		strip.background = element_blank(),
+#		strip.text.y = element_text(size = 11)
+		axis.text.y = element_text(size=8),
+		axis.text.x = element_blank(),
+		axis.title = element_text(size=8),
+		legend.position="none"
+	)
+
+my_legend <- theme(
+		legend.title=element_text(size=0),
+		legend.text = element_text(size=5),
+		legend.key.size = unit(0.55, "line"),
+		legend.background = element_blank()
+	)
+
+
+cluster_data <- read.table(file="data/processed/cluster_data.summary", header=TRUE, stringsAsFactors=FALSE)
+
+mcc_methods <- c("mcc", "mcc_agg")
+pretty_methods <- c("Separate OTUs", "Single OTU")
+
+datasets <- c('soil', 'marine', 'mice', 'human', 'even', 'staggered')
+pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
+names(pretty_datasets) <- datasets
+
+agg_comparison <- cluster_data %>%
+			filter(frac==1.0 & method %in% mcc_methods) %>%
+			group_by(dataset, method) %>%
+			summarize(avg_mcc=mean(mcc, na.rm=T),
+								min_mcc=min(mcc, na.rm=T),
+								max_mcc=max(mcc, na.rm=T),
+								avg_secs=mean(cluster_secs, na.rm=T),
+								min_secs=min(cluster_secs, na.rm=T),
+								max_secs=max(cluster_secs, na.rm=T))
+agg_comparison$dataset <- factor(agg_comparison$dataset, datasets)
+
+
+steps_data <- read.table(file="data/processed/cluster_steps.summary", header=TRUE, stringsAsFactors=FALSE)
+
+completion <- c("close_enough", "convergence")
+pretty_completion <- c("Converge within 0.01%", "Complete convergence")
+names(pretty_completion) <- completion
+
+
+steps_comparison <- steps_data %>%
[…]

### 2016-09-29 code/build_seeding_figure.R

@@ -1,53 +1,0 @@
-library(dplyr)
-library(ggplot2)
-library(tidyr)
-library(wesanderson)
-
-cluster_data <- read.table(file="data/processed/cluster_data.summary", header=TRUE, stringsAsFactors=FALSE)
-
-mcc_methods <- c("mcc", "mcc_agg")
-pretty_methods <- c("Separate OTUs", "Single OTU")
-
-datasets <- c('soil', 'marine', 'mice', 'human', 'even', 'staggered')
-pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
-names(pretty_datasets) <- datasets
-
-cluster <- cluster_data %>%
-			filter(frac==1.0 & method %in% mcc_methods) %>%
-			group_by(dataset, method) %>%
-			summarize(avg_mcc=mean(mcc, na.rm=T),
-								avg_secs=mean(cluster_secs, na.rm=T))
-cluster$dataset <- factor(cluster$dataset, datasets)
-
-my_theme <- theme_classic() +
-	theme(
-		panel.grid.major = element_blank(),
-		panel.grid.minor = element_blank(),
-		panel.border = element_rect(color = "black", fill=NA, size=1),
-		panel.background = element_rect(fill=NA),
-		strip.background = element_blank(),
-		strip.text.y = element_text(size = 11),
-		axis.text.x=element_text(size=10),
-		legend.position=c(.75, .7),
-		legend.title=element_text(size=9),
-		legend.text = element_text(size=9),
-		legend.key.size = unit(0.75, "line")
-	)
-
-
-gathered <- gather(cluster, parameter, variable, avg_mcc, avg_secs)
-
-facet_labels <- as_labeller(c('avg_mcc' = "Mean Matthew's\nCorrelation Coefficient", 'avg_secs' = "Execution Time\n(seconds)"))
-
-ggplot(gathered, aes(dataset, variable, col=dataset, shape=method)) +
-	geom_point(position = position_dodge(0.1), size=2) +
-	facet_grid(parameter~., scales = 'free_y', switch = 'y', labeller = facet_labels) +
-	scale_x_discrete(breaks=levels(gathered$dataset),
-		labels=pretty_datasets[levels(gathered$dataset)]) +
-	scale_color_manual(breaks=datasets, labels=pretty_datasets, values=
-		c('black', wes_palette("Darjeeling")), guide=F)+
[…]

### 2016-09-29 code/build_algorithm_figure.R
Make plot consist of MCC, Sobs, and time to completion

@@ -1,119 +1,0 @@
-library(dplyr)
-library(ggplot2)
-library(tidyr)
-library(wesanderson)
-library(cowplot)
-
-
-my_theme <- theme_classic() +
-	theme(
-		panel.grid.major = element_blank(),
-		panel.grid.minor = element_blank(),
-		panel.border = element_rect(color = "black", fill=NA, size=1),
-		panel.background = element_rect(fill=NA),
-#		strip.background = element_blank(),
-#		strip.text.y = element_text(size = 11)
-		axis.text.y = element_text(size=8),
-		axis.text.x = element_blank(),
-		axis.title = element_text(size=8),
-		legend.position="none"
-	)
-
-my_legend <- theme(
-		legend.title=element_text(size=0),
-		legend.text = element_text(size=5),
-		legend.key.size = unit(0.55, "line"),
-		legend.background = element_blank()
-	)
-
-
-cluster_data <- read.table(file="data/processed/cluster_data.summary", header=TRUE, stringsAsFactors=FALSE)
-
-mcc_methods <- c("mcc", "mcc_agg")
-pretty_methods <- c("Separate OTUs", "Single OTU")
-
-datasets <- c('soil', 'marine', 'mice', 'human', 'even', 'staggered')
-pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
-names(pretty_datasets) <- datasets
-
-agg_comparison <- cluster_data %>%
-			filter(frac==1.0 & method %in% mcc_methods) %>%
-			group_by(dataset, method) %>%
-			summarize(avg_mcc=mean(mcc, na.rm=T),
-								min_mcc=min(mcc, na.rm=T),
-								max_mcc=max(mcc, na.rm=T),
-								avg_secs=mean(cluster_secs, na.rm=T),
-								min_secs=min(cluster_secs, na.rm=T),
-								max_secs=max(cluster_secs, na.rm=T))
-agg_comparison$dataset <- factor(agg_comparison$dataset, datasets)
-
-
-steps_data <- read.table(file="data/processed/cluster_steps.summary", header=TRUE, stringsAsFactors=FALSE)
-
-completion <- c("close_enough", "convergence")
-pretty_completion <- c("Converge within 0.01%", "Complete convergence")
-names(pretty_completion) <- completion
-
-
-steps_comparison <- steps_data %>%
[…]

### 2016-09-29 code/build_seeding_figure.R

@@ -1,0 +1,73 @@
+library(dplyr)
+library(ggplot2)
+library(tidyr)
+library(wesanderson)
+library(cowplot)
+
+
+my_theme <- theme_classic() +
+	theme(
+		panel.grid.major = element_blank(),
+		panel.grid.minor = element_blank(),
+		panel.border = element_rect(color = "black", fill=NA, size=1),
+		panel.background = element_rect(fill=NA),
+		strip.background = element_blank(),
+		axis.text.y = element_text(size=7),
+		axis.text.x = element_text(size=7),
+		axis.title = element_text(size=7),
+		legend.position="none"
+	)
+
+
+
+
+cluster_data <- read.table(file="data/processed/cluster_data.summary", header=TRUE, stringsAsFactors=FALSE)
+
+mcc_methods <- c("mcc", "mcc_agg")
+pretty_methods <- c("Separate OTUs", "Single OTU")
+names(pretty_methods) <- mcc_methods
+
+datasets <- c('soil', 'marine', 'mice', 'human', 'even', 'staggered')
+pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
+names(pretty_datasets) <- datasets
+
+variables <- c("mcc", "num_otus", "cluster_secs")
+pretty_variables <- c("Matthew's\nCorrelation Coefficient", "Number of OTUs", "Time to Cluster\n(seconds)")
+names(pretty_variables) <- variables
+
+agg_comparison <- cluster_data %>%
+			filter(frac==1.0 & method %in% mcc_methods) %>%
+			gather(variable, value, mcc, num_otus, cluster_secs) %>%
+			group_by(dataset, method, variable) %>%
+			summarize(avg=mean(value, na.rm=T),
+								min=min(value, na.rm=T),
+								max=max(value, na.rm=T)
+								)
+agg_comparison$dataset <- factor(agg_comparison$dataset, datasets)
+agg_comparison$method <- factor(agg_comparison$method, mcc_methods)
+agg_comparison$variable <- factor(agg_comparison$variable, variables)
+
+facet_labels <- as_labeller(pretty_variables)
+
+figure <- ggplot(agg_comparison, aes(dataset, avg, col=dataset, shape=method))+
+	geom_point(position = position_dodge(0.3), size=2) +
[…]

### 2016-09-30 code/aggregate_steps.R
Have code bring in timing and sobs data

@@ -1,3 +1,5 @@
+library(dplyr)
+
 #this function concatenates together the dataset, fraction, replicate number,
 #and method to generate the filename stub
 cat_cols_files <- function(x){
@@ -16,13 +18,19 @@ build_steps_files <- function(x){
 
 
 #return the total number of steps to reach point where change is less than #0.01% or to convergence (0.00% change)
-get_required_steps <- function(file){
+get_required_steps <- function(file, time){
 	steps <- read.table(file=file, header=T)
 
 	mcc_a <- steps[-nrow(steps),"mcc"]
 	mcc_b <- steps[-1, "mcc"]
 
-	c(min(steps$iter[(mcc_b - mcc_a) / mcc_a < 0.0001]), max(steps$iter))
+	threshold <- which.min((mcc_b - mcc_a) / mcc_a > 0.0001) + 1
+	extra_time <- sum(steps$time[(threshold+1):nrow(steps)])
+	full_time <- time[file,"cluster_secs"]
+	partial_time <- full_time-extra_time
+
+	c(steps$iter[threshold], steps$num_otus[threshold], partial_time,
+		 max(steps$iter), steps$num_otus[nrow(steps)], full_time)
 }
 
 # Six datasets - two synthetic and four biological
@@ -48,8 +56,15 @@ method_vector <- rep(methods, each=n_reps, times=n_fracs*n_datasets)
 
 results <- data.frame(dataset=dataset_vector, frac=frac_vector, rep=rep_vector, method=method_vector, stringsAsFactors=F)
 
+cluster_data <- read.table(file="data/processed/cluster_data.summary", header=T,
+													stringsAsFactors=FALSE)
+timings <- cluster_data %>%
+							filter(method %in% methods, frac==1) %>%
+							select(dataset, frac, rep, method, cluster_secs)
+rownames(timings) <- apply(timings[1:4], 1, build_steps_files)
+
 steps_names <- apply(results, 1, build_steps_files)
-all_stats <- t(sapply(steps_names, get_required_steps))
+all_stats <- t(sapply(steps_names, get_required_steps, time=timings))
 results$close_enough <- all_stats[,1] #within 0.01%
 results$convergence <- all_stats[,2] #complete convergence
 

### 2016-09-30 code/build_performance_figure.R
Edit labels

@@ -52,7 +52,7 @@ my_theme <- theme_classic() +
 plot_performance <- function(x, var, label, ylimits){
 	mcc <- ggplot(x, aes_string("method", var, col="dataset", shape="dataset")) +
 		geom_point(position = position_dodge(0.5), size=2) +
-		geom_vline(xintercept=0.5 + 1:(length(levels(x$method))-1)) +
+		geom_vline(xintercept=0.5 + (1:(length(levels(x$method))-1))) +
 		coord_cartesian(ylim=ylimits) +
 		ylab(label) +
 		xlab(NULL) +
@@ -71,15 +71,15 @@ plot_performance <- function(x, var, label, ylimits){
 	return(mcc)
 }
 
-mcc <- plot_performance(subset, 'avg_mcc', "Mean Matthew's\nCorrelation Coefficient", c(0,1))
-cov <- plot_performance(subset, 'cov_mcc', "Coefficient of Variation for\nMatthew's Correlation Coefficient", c(0,0.04))
+mcc <- plot_performance(subset, 'avg_mcc', "Matthew's\nCorrelation Coefficient", c(0,1))
+cov <- plot_performance(subset, 'cov_mcc', "Coefficient of\nVariation for MCC", c(0,0.04))
 sobs <- plot_performance(subset, 'avg_sobs', "Number of OTUs", c(0,1.05e5))
 
 my_legend <- theme(
-	legend.title=element_blank(),
-	legend.position = c(0.12, 0.7),
-	legend.text = element_text(size = 8),
-	legend.key.height=unit(0.7,"line"),
+	legend.position = c(0.12, 0.6),
+	legend.title=element_text(size=0),
+	legend.text = element_text(size=5),
+	legend.key.size = unit(0.55, "line"),
 	legend.key = element_rect(fill = NA, linetype=0),
 	legend.margin = unit(-2,"line")
 )
@@ -87,6 +87,6 @@ my_legend <- theme(
 ggdraw() +
 	draw_plot(mcc,0,0.7,1,0.3) +
 	draw_plot(cov + my_legend, 0,0.40,1,0.3) +
-	draw_plot(sobs + 	theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5, size=8)), 0,0.02,1,0.38) +
-	draw_plot_label(c("A", "B", "C"), x=c(0,0,0), y=c(1.00,0.72,0.42), size=18) +
[…]

### 2016-10-21 code/build_performance_figure.R
Create summary file that contains data for the performance figure

@@ -58,6 +58,7 @@ tidy$method <- factor(tidy$method, method_ordering$method)
 tidy$dataset <- factor(tidy$dataset, datasets)
 tidy$metric <- factor(tidy$metric, metrics)
 
+write.table(tidy, file="data/processed/performance_data.tsv", row.names=F, sep='\t', quote=F)
 
 
 my_theme <- theme_classic() +

### 2016-10-22 submission/Westcott_OptiClust_mSystems_2016.Rmd
Finish first full draft of results section

---
title: "**OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units**"
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
csl: msystems.csl
fontsize: 11pt
geometry: margin=1.0in
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
library(dplyr)
library(tidyr)
library(knitr)

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
	print(x)
	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
		}

	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)

basic_methods <- c('an', 'fn', 'nn', 'otuclust', 'sumaclust', 'swarm',
										'uagc', 'udgc', 'vagc_1', 'vdgc_1', 'mcc')

```

\begin{center}
\vspace{25mm}
Sarah L. Westcott and Patrick D. Schloss${^\dagger}$

\vspace{30mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
\end{center}

\newpage
\linenumbers

## Abstract

\newpage

## Introduction

Amplicon-based sequencing has provided incredible insights into Earth's microbial biodiversity.

Numerous methods have been proposed

Needed an objective criteria to evaluate the various methods methods

Average neighbor is consistently the best method, but others including USEARCH and VSEARCH resulted in comparable results

Difficulty with amount of memory and time required to complete average neighbor algorithm is a significant hurdle to completion of the method.

Instead of retrospectively evaluating methods for their ability to correctly assign sequences we sought to develop a method that would prospectively assign sequences to OTUs to optimize classification metrics.

Clustering quality within the algorithm is assessed by counting the number of true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN) based on the pairwise distances. Sequence pairs that are within the user-specified threshold and are clustered together represent TPs and those in different OTUs are FNs. Those sequence pairs that have a distance larger than the threshold and are not clustered in the same OTU are TNs and those in the same OTU are FPs. These counts are used to calculate the optimization metric.

## Results

***OptiClust algorithm.***

The OptiClust algorithm uses the metric that should be used to assess clustering quality, a list of all sequence names in the dataset, and the pairs of sequences that are within a desired threshold of each other (e.g. 0.03). A detailed description of the algorithm is provided for a toy dataset in the Supplementary Material. Briefly, the algorithm starts by placing each sequence either within its own OTU or into a single OTU. The algorithm proceeds by interrogating each sequence and re-calculating the metric for the cases where the sequence stays in its current OTU, is moved to each of the other OTUs, or is moved into a new OTU. The location that results in the best clustering quality indicates whether the sequence should remain in its current OTU or be moved to a different or new OTU. Each iteration consists of interrogating every sequence in the dataset. Although numerous options are available within the mothur-based implementation of the algorithm (e.g. sensitivity, specificity, accuracy, F1 score, etc.), the default metric is MCC because it includes all four parameters from the confusion matrix. The algorithm continues until the optimization metric stabilizes or until it reaches a defined stopping criteria.

***OptiClust-generated OTUs are more robust than those from other methods.***

```{r performance_analysis}
perf <- read.table(file="data/processed/cluster_data.summary", header=T) %>%
				filter(frac == 1.0, method %in% basic_methods) %>%
				select(dataset, method, cluster_secs, cluster_kb, mcc, num_otus) %>%
				group_by(dataset, method) %>%
				summarize(
					mean_secs = mean(cluster_secs, na.rm=T),
					mean_kb = mean(cluster_kb, na.rm=T),
					mean_mcc = mean(mcc, na.rm=T),
					mean_num_otus = mean(num_otus, na.rm=T)
				)
perf_summary <- perf %>%
								group_by(method) %>%
								summarize(
									mean_mcc = mean(mean_mcc, na.rm=T),
									mean_secs = mean(mean_secs, na.rm=T)
								) %>%
								mutate(
									rel_mcc = mean_mcc/mean_mcc[method=='mcc'],
									rel_secs = mean_secs/mean_secs[method=='mcc']
								)

mcc_rel_an <-100*(1-perf_summary %>% filter(method=="an") %>% select(rel_mcc))
mcc_rel_vdgc <- 100*(1-perf_summary %>% filter(method=="vdgc_1") %>% select(rel_mcc))

secs_rel_an <- perf_summary %>% filter(method=="an") %>% select(rel_secs)
secs_rel_vdgc <- perf_summary %>% filter(method=="vdgc_1") %>% select(rel_secs)

mcc_sobs_cor <- cor.test(perf$mean_num_otus, perf$mean_mcc, method="spearman")
```

To evaluate the OptiClust algorithm and compare its performance to other algorithms, we utilized six datasets including two synthetic communities and four previously published large datasets generated from soil, marine, human, and murine samples (Table 1). When we seeded the OptiClust algorithm with each sequence in a separate OTU and ran the algorithm until complete convergence, the MCC values were `r round(mcc_rel_an, 1)` and `r round(mcc_rel_vdgc, 1)`% higher than the OTUs using average neighbor and distance-based greedy clustering (DGC) with VSEARCH, respectively (Figure 1). The number of OTUs formed by the various methods was negatively correlated with their MCC value (\rho=`r round(mcc_sobs_cor$estimate, 2)`; p=`r round(mcc_sobs_cor$p.value, 2)`). The OptiClust algorithm was considerably faster than the hierarchical algorithms and somewhat slower than the heuristic-based algorithms. Across the six datasets, the OptiClust algorithm was `r round(secs_rel_an, 1)`-times faster than average neighbor and `r round(1/secs_rel_vdgc, 1)`-times slower than DGC with VSEARCH. The human dataset was a challenge for a number of the algorithms. OTUCLUST and SumaClust were unable to cluster the human dataset in less than 50 hours and the average neighbor algorithm required more than 48 GB of RAM. The USEARCH-based methods were unable to cluster the human data using the 32-bit free version of the software that limits the amount of RAM to approximately 3.5 GB. These data demonstrate that OptiClust generates significantly more robust OTU assignments than existing methods across a diverse collection of datasets with performance that is comparable to popular methods.

***OptiClust stopping criteria.***

```{r stopping_analysis}
steps <- read.table(file="data/processed/mcc_steps.summary", header=T) %>%
					filter(frac == 1.0, dataset!="even", dataset!="staggered") %>%
					select(-rep, -frac, -method, -contains("_secs_cluster")) %>%
					group_by(dataset) %>%
					summarize(med_full_secs = median(full_secs),
							per_diff_mcc = 100 * mean((full_mcc-close_mcc)/full_mcc),
							diff_numotus = median(close_num_otus - full_num_otus),
							per_diff_numotus = 100 * mean((close_num_otus - full_num_otus)/full_num_otus),
							ratio_secs = mean(full_secs/close_secs),
							med_close_steps = median(close_steps),
							med_full_steps = median(full_steps)
					)
```

By default, the mothur-based implementation of the algorithm stops when the optimization metric changes by less than 0.0001; however, this can be altered by the user. This implementation also allows the user to stop the algorithm if a maximum number of iterations is exceeded. By default mothur uses a maximum value of 100 iterations. The justification for allowing incomplete convergence was based on the observation that numerous iterations are performed that extend the time required to complete the clustering with minimal improvement in clustering. We evaluated the results of clustering to partial convergence (i.e. a change in the MCC value that was less than 0.0001) or until complete convergence of the MCC value (i.e. until it did not change between iterations) when seeding the algorithm with each sequence in a separate OTU (Figure 1). The small difference in MCC values between the output from partial and complete convergence resulted in a difference in the median number of OTUs that ranged between `r paste(format(round(range(steps$diff_numotus), 1), nsmall=1L), collapse=" and ")` OTUs. This represented a difference of less than `r format(round(max(steps$per_diff_numotus), 2), nsmall=1L)`%. Among the four natural datasets, between `r min(steps$med_close_steps)` and `r max(steps$med_close_steps)` were needed to achieve partial convergence and between `r min(steps$med_full_steps)` and `r max(steps$med_full_steps)` iterations were needed to reach full convergence. The additional steps required between `r paste(format(round(range(steps$ratio_secs), 1), nsmall=1L), collapse=" and ")` times longer to complete the algorithm. These results suggest that achieving full convergence of the optimization metric adds computational effort; however, considering full convergence took between `r
paste(round(range(steps$med_full_secs/60), 0), collapse=" and ")` minutes the extra effort was relatively small. Although the mothur's default setting is partial convergence, the remainder of our analysis used complete convergence to be more conservative.

***Effect of seeding OTUs on OptiClust performance.***

```{r aggregate_analysis}
agg <- read.table(file="data/processed/cluster_data.summary", header=T) %>%
			filter(frac==1.0, method %in% (c("mcc", "mcc_agg"))) %>%
			select(dataset, method, cluster_secs, mcc, num_otus) %>%
			group_by(dataset, method) %>%
			summarize(
				mean_secs = mean(cluster_secs, na.rm=T),
				mean_mcc = mean(mcc, na.rm=T),
				mean_num_otus = mean(num_otus, na.rm=T)
			) %>%
			group_by(dataset) %>%
			summarize(
				fold_secs = mean_secs[method=="mcc_agg"]/mean_secs[method=="mcc"],
				per_diff_mcc = 100*(1-mean_mcc[method=="mcc_agg"]/mean_mcc[method=="mcc"]),
				per_diff_num_otus =100*(1-mean_num_otus[method=="mcc_agg"]/mean_num_otus[method=="mcc"])
			)
```

As implemented within mothur, the OptiClust algorithm either starts with each sequence in a separate OTU or with all of the sequences in a single OTU. We repeated the complete convergence analysis, but seeded the algorithm with all sequences in a single OTU. We found that the MCC values for clusters generated seeding OptiClust with the sequences as a single OTU were between `r paste(round(range(agg$per_diff_mcc), 1), collapse=" and ")`% lower than when seeding the algorithm with sequences in separate OTUs (Figure 1). Interestingly, with the exception of the `r agg[which.min(agg$per_diff_num_otus),"dataset"]` dataset (`r format(abs(round(agg[which.min(agg$per_diff_num_otus),"per_diff_num_otus"], 1)), nsmall=1L)`% more OTUs), the number of OTUs was as much as `r format(round(agg[which.max(agg$per_diff_num_otus),"per_diff_num_otus"], 1), nsmall=1L)`% lower (`r agg[which.max(agg$per_diff_num_otus),"dataset"]`) than when the algorithm was seeded with sequence in separate OTUs. Finally, the amount of time required to cluster the data when the algorithm was seeded with a single OTU was between `r paste(format(round(range(agg$fold_secs), 1), nsmall=1L), collapse=" and ")`-times longer than if sequences were seeded as separate OTUs. This analysis demonstrates that seeding the algorithm with sequences as separate OTUs results in the best OTU assignments in the shortest amount of time.

***OptiClust-generated OTUs are as stable as those from other algorithms.***

```{r stability_analysis}
stability <- read.table(file="data/processed/cluster_data.summary",header=T) %>%
						filter(frac == 1.0, method %in% basic_methods) %>%
						select(dataset, method, mcc, num_otus) %>%
						group_by(dataset, method) %>%
						summarize(
							cov_mcc = sd(mcc) / mean(mcc) * 100,
							cov_num_otus = sd(num_otus) / mean(num_otus) * 100
						) %>%
						group_by(method) %>%
						summarize(
							med_cov_mcc = median(cov_mcc, na.rm=T),
							med_cov_num_otus = median(cov_num_otus, na.rm=T)
						)
```

One concern that many have with *de novo* clustering algorithms is that their output is sensitive to the initial order of the sequences. An additional concern with the OptiClust algorithm is that it may stabilize at a local optimum. To evaluate these concerns we compared the results obtained using ten randomizations of the order that sequences were given to the algorithm. The median the coefficient of variation across the six datasets for MCC values obtained from the replicate clusterings using OptiClust was `r format(round(stability %>% filter(method=="mcc") %>% select(med_cov_mcc), 1), nsmall=1L)`% (Figure 1). We also measured the coefficient of variation for the number of OTUs across the six datasets for each method. The median coefficient of variation for the number of OTUs generated using OptiClust was `r format(round(stability %>% filter(method=="mcc") %>% select(med_cov_num_otus), 1), nsmall=1L)`%. Confirming our previous results, all of the methods we tested were stable to stochastic processes. Of the method that involved randomization, the coefficient of variation for MCC values considerably smaller than the other methods and the coefficient of variation for the number of OTUs was comparable to the other methods. The variation observed in clustering quality suggests that the algorithm does not appear to converge to a locally optimum MCC value. More importantly, the random variation does yield output of a similarly high quality.

***Time and memory required to complete Optimization-based clustering scales efficiently.***

```{r scaling_analysis}
datasets <- c('human', 'mice', 'marine', 'soil')

speed_data <- read.table(file="data/processed/mcc_steps.summary", header=T)

speed_models <- speed_data %>%
		filter(dataset %in% datasets) %>%
		select(dataset, frac, full_secs_cluster, full_steps) %>%
		group_by(dataset) %>%
		do(
			model = nls((full_secs_cluster/full_steps) ~ b * frac ^ z,
														start = list(b = 1000, z = 2), data= .)
		)

speed_coef <- c(
		speed_models[speed_models$dataset=="human",2][[1]][[1]]$m$getPars()["z"],
		speed_models[speed_models$dataset=="mice",2][[1]][[1]]$m$getPars()["z"],
		speed_models[speed_models$dataset=="marine",2][[1]][[1]]$m$getPars()["z"],
		speed_models[speed_models$dataset=="soil",2][[1]][[1]]$m$getPars()["z"]
	)
names(speed_coef) <- datasets

memory_data <- read.table(file="data/processed/cluster_data.summary", header=T)

memory_models <- memory_data %>%
		filter(dataset %in% datasets, method == "mcc") %>%
		mutate(cluster_gb=cluster_kb/1048000) %>%
		select(dataset, frac, cluster_gb) %>%
		group_by(dataset) %>%
		do(model = nls((cluster_gb) ~ b * frac ^ z, start = list(b = 1e6, z = 3), data= .))

memory_coef <- c(
		memory_models[memory_models$dataset=="human",2][[1]][[1]]$m$getPars()["z"],
		memory_models[memory_models$dataset=="mice",2][[1]][[1]]$m$getPars()["z"],
		memory_models[memory_models$dataset=="marine",2][[1]][[1]]$m$getPars()["z"],
		memory_models[memory_models$dataset=="soil",2][[1]][[1]]$m$getPars()["z"]
	)
names(memory_coef) <- datasets

```

Although not as important as the quality of clustering, the amount of time and memory required to assign sequences to OTUs is a legitimate concern. To evaluate how the speed and memory usage scaled with the number of sequences in the dataset, we measured the time required and maximum RAM usage to cluster 20, 40, 60, 80, and 100% of the unique sequences from each of the natural datasets using the OptiClust algorithm (Figure 2). Within each iteration of the algorithm, each sequence is compared to every other sequence and each comparison requires a recalculation of the confusion matrix. This would result in a worst case algorithmic complexity on the order of N^3, where N is the number of unique sequences. Because the algorithm only needs to keep track of the sequence pairs that are within the threshold of each other, it is likely that the implementation of the algorithm is more efficient. To empirically determine the algorithmic complexity, we fit a power law function to the data in Figure 2A. We observed power coefficients between `r paste(format(round(range(speed_coef), 1), nsmall=1L), collapse=" and ")` for the `r paste(names(speed_coef[c(which.min(speed_coef), which.max(speed_coef))]), collapse=" and ")` datasets, respectively. The algorithm requires storing a matrix that contains the pairs of sequences that are close to each other as well as a matrix that indicates which sequences are clustered together. The memory required to store these matrices is on the order of N^2, where N is the number of unique sequences. In fact, when we fit a power law function to the data in Figure 2B, the power coefficients were  `r format(round(median(memory_coef), 1), nsmall=1L)`. This analysis suggests that doubling the number of sequences in a dataset would increase the time required to cluster the data by 4 to 8-fold and increase the RAM required by 4-fold. It is possible that future improvements to the implementation of the algorithm could improve this performance.

```{r split_analysis}
datasets <- c('soil', 'marine', 'mice', 'human')

full_data <- read.table(file="data/processed/cluster_data.summary", header=T, stringsAsFactors=FALSE)

full_data$method <- gsub("^mcc$", "mcc_split1_8", full_data$method)
full_data$method <- gsub("^vdgc_1$", "vdgc_split1_8", full_data$method)
full_data$method <- gsub("^an$", "an_split1_8", full_data$method)

split_data <- full_data %>%
				mutate(method = gsub("^mcc$", "mcc_split1_8", method)) %>%
				mutate(method = gsub("^vdgc_1$", "vdgc_split1_8", method)) %>%
				mutate(method = gsub("^an$", "an_split1_8", method)) %>%
				filter(
						grepl("split", method),
						frac==1.0,
						dataset==datasets
				) %>%
				select(dataset, method, mcc, num_otus) %>%
				separate(method, into=c("method", "split", "processors"), sep="_") %>%
				mutate(tax_level=as.numeric(gsub("split", "", split))) %>%
				select(-split, -processors) %>%
				gather(metric, value, mcc, num_otus) %>%
				group_by(dataset, method, metric, tax_level) %>%
				summarize(
					mean=if(sum(!is.na(value))>0){mean(value, na.rm=TRUE)}	else { NA }
				)

rel_mcc_data <- split_data %>%
			filter(metric == "mcc") %>%
			group_by(dataset, method, tax_level) %>%
			summarize(mean = mean(mean, na.rm=T)) %>%
			group_by(dataset, method) %>%
			mutate(rel_mean=mean/mean[tax_level==1]) %>%
			group_by(method, tax_level) %>%
			summarize(min_mcc = min(rel_mean, na.rm=T))

rel_otu_data <- split_data %>%
			filter(metric == "num_otus") %>%
			group_by(dataset, method, tax_level) %>%
			summarize(mean = mean(mean, na.rm=T)) %>%
			group_by(dataset, method) %>%
			mutate(rel_mean=mean/mean[tax_level==1]) %>%
			group_by(method, tax_level) %>%
			summarize(max_num_otus = max(rel_mean, na.rm=T))

```

***Cluster splitting heuristic generates OTUs that are as good as non-split approach (Figure 6).***

We previously described a heuristic to accelerate OTU assignments where sequences were classified to taxonomic groups and within each taxon sequences were assigned to OTUs using the average neighbor clustering algorithm. This can accelerate the clustering and reduce the memory requirements because the number of unique sequences is effectively reduced by splitting sequences across taxonomic groups. Furthermore, because sequences in different taxonomic groups are assumed to belong to different OTUs they are independent, which permits parallelization and additional reduction in computation time. Reduction in clustering quality are encountered in this approach if there are errors in classification or if two sequences within the desired threshold belong to different taxonomic groups. It is expected that these errors would increase as the taxonomic level goes from kingdom to genus. To characterize the clustering quality, we calculated the MCC values using OptiClust, average neighbor, and DGC with VSEARCH when splitting at each taxonomic level (Figure 3). For each method, the MCC values decreased as the taxonomic resolution increased; however, the decrease in MCC as not as large as the difference between clustering methods. As the resolution of the taxonomic levels increased, the clustering quality remained high, relative to clusters formed from the entire dataset (i.e. kingdom-level). The MCC values when splitting the datasets at the class and genus levels were within `r format(round(min(100*(rel_mcc_data %>% filter(tax_level == 3))$min_mcc), 1), nsmall=1L)` and `r format(round(min(100*(rel_mcc_data %>% filter(tax_level == 6))$min_mcc), 1), nsmall=1L)`%, respectively, of the MCC values obtained from the entire dataset. These decreases in MCC value resulted in the formation of as many as `r format(round(min(100*((rel_otu_data %>% filter(tax_level == 3))$max_num_otus-1), na.rm=T), 1), nsmall=1L)` and `r format(round(min(100*((rel_otu_data %>% filter(tax_level == 6))$max_num_otus-1), na.rm=T), 1), nsmall=1L)`% more OTUs, respectively, than were observed from the entire dataset. For the datasets included in the current analysis, the use of the cluster splitting heuristic is not worth the loss in clustering quality. However, as datasets become larger, it may be necessary to use the heuristic to clustering the data into OTUs. For example, we were unable to cluster the full human data using less than 48 GB of RAM; however, when we split the dataset at the family level, we were able to cluster the data with the limited resources.

## Discussion
Restate most important contributions
* Optimized clustering based on an objective criteria. Results are meaningfully better than existing methods
* Added benefits include smaller CPU and RAM footprint
* Result is efficient analysis of large datasets without sacrificing clustering quality as has been experienced in heuristic methods.

Choice of objective criteria
* Value of using a metric that is based on all four parameters
* Preference is for Matthew's Correlation Coefficient because... other options include F1 Score and accuracy

Preference for OptiClust over cluster splitting approach
* Risks are mis-classification and artificially splitting similar sequences between OTUs because they classify to different taxa
* Still potential to merge the methods for more efficient processing.

Data quality - analysis uses unique sequences and we should not expect an infinite number of unique sequences unless there is a large amount of random sequencing error.

Based on our model, we propose the following...

***Optimization of clustering using composite metrics significantly improves clustering quality (Figure 2, 3, 4).*** There are multiple metrics available to assess clustering quality. The mothur-based implementation of OptiClust allows the user to use the MCC, F1 score, accuracy, sensitivity, specificity, and the sum of true positives and negatives. The MCC, F1 score, and accuracy are preferred because each of them incorporates all four values from the confusion matrix while others only utilize two values. It is relatively straightforward to implement other metrics. **Comparison of MCC, accuracy, and F1 score to each other for each dataset using the full set of sequences.** Each metric outperformed the observed values for the other clustering algorithms indicating that regardless of the metric one uses to evaluate clustering quality, the OptiClust algorithm generates better OTU assignments than any of the other methods. The number of OTUs generated by an algorithm is often used as a metric for clustering quality. We did not observe a significant correlation between clustering quality as measured by MCC, F1 score, or accuracy and the number of OTUs generated by each algorithm. **Interestingly, the values of the MCC, F1 score, and accuracy generated when using each metric to optimize clustering were similar across implementations.** Based on these results and its previous use in the OTU assignment literature, the remainder of our analysis uses the MCC metric for optimization.

## Materials and Methods

***Sequence data and processing steps.***
To evaluate the OptiClust and the other algorithms we created two synthetic sequence collections and four sequence collections generated from previously published studies. The V4 region of the 16S rRNA gene was used from all datasets because it is a popular region that can be fully sequenced with two-fold coverage using the commonly used MiSeq sequencer from Illumina [@Kozich2013]. The method for generating the simulated datasets followed the approach used by Kopylova et al. [-@Kopylova2016] and Schloss [@Schloss2016]. Briefly, we randomly selected 10,000 uniques V4 fragments from 16S rRNA gene sequences that were unique from the SILVA non-redundant database [@Pruesse2007]. A community with an even relative abundance profile was generated by specifying that each sequence had a frequency of 100 reads. A community with a staggered relative abundance profile was generated by specifying that the abundance of each sequence was a randomly drawn integer sampled from a uniform distribution between 1 and 200. Sequence collections collected from human feces [@Baxter2016], murine feces [@Schloss2012], soil [@Johnston2016], and seawater [@Henson2016] were used to characterize the algorithms' performance with natural communities. These sequence collections were all generated using paired 150 or 250 nt reads of the V4 region. We re-processed all of the reads using a common analysis pipeline that included quality score-based error correction [@Kozich2013], alignment against a SILVA reference database [@Pruesse2007; @Schloss2010], screening for chimeras using UCHIME [@Edgar2011], and classification using a naive Bayesian classifier with the RDP training set [@Wang2007].

***Implementation of clustering algorithms.***
In addition to the OptiClust algorithm we evaluated ten different *de novo* clustering algorithms. These included three hierarchical algorithms, average neighbor (AN), nearest neighbor (NN), and furthest neighbor (FN), which are implemented in mothur [v.1.39.0; @Schloss2009]. Seven heuristic methods were also used including abundance-based greedy clustering (UAGC) and distance-based greedy clustering (UDGC) as implemented in USEARCH [v.6.1; @Edgar2010], abundance-based greedy clustering (VAGC) and distance-based greedy clustering (VDGC) as implemented in VSEARCH [v.X.X.X; @Rognes2015], OTUClust [v.X.X.X; @XXXXXX], SumaClust [v.X.X.X; @XXXXXX], and Swarm [v.2.1.1; @Mah2014]. With the exception of Swarm each of these methods uses distance-based thresholds to report OTU assignments. To judge the quality of the Swarm-generated OTU assignments we calculated the MCC value using thresholds incremented by 1% between 0 and 5% and selected the threshold that provided the optimal MCC value [@Westcott2015]. We also assessed the ability of a previously describe heuristic to cluster sequences using the UDGC and OptiClust algorithms. In this heuristic sequences are split into bacterial families based on sequence classification and clustered within the taxonomic family [@Westcott20XX]. Finally, for our benchmarking analysis, we evaluate the memory and time requirements when using 8 processors with the UDGC and VDGC algorithms and with the taxonomic splitting heuristic.

***Benchmarking.*** We evaluated the quality of the sequence clustering, reproducibility of the clustering, the speed of clustering, and the amount of memory required to complete the clustering. To assess the quality of the clusters generated by each method, we counted the cells within a confusion matrix that indicated how well the clusterings represented the distances between the pair of sequences [@Schloss2011Assessing]. Pairs of sequences that were in the same OTU and had a distance less than 3% were true positives (TPs), those that were in different OTUs and had a distance greater than 3% were true negatives (TNs), those that were in the same OTU and had a distance greater than 3% were false positives (FPs), and those that were in different OTUs and had a distance less than 3% were false negatives (FNs). To synthesize the matrix into a single metric we used the Matthew's Correlation Coefficient, F1 score, and accuracy using the `sens.spec` command in mothur using the following equations.

$$
MCC = \frac{TP \times TN-FP \times FN}{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)} }
$$

$$
F1_score = \frac{2 \times TP}{2 \times TP+FP+FN}
$$

$$
Accuracy = \frac{TP + TN}{(TP+FP+TN+FN)}
$$

To assess the reproducibility of the algorithms we randomized the starting order of each sequence collection ten times and ran each algorithm on each randomized collection. We then measured the MCC, F1 score, and accuracy for each randomization and quantified their coefficient of variation (CV; the ratio of the standard deviation to the mean).

To assess how the the memory and time requirements scaled with the number of sequences included in each sequence collection, we randomly subsampled 20, 40, 60, or 80% of the unique sequences in each collection. We obtained 10 subsamples at each depth for each dataset and ran each collection (N= 50 = 5 sequencing depths x 10 replicates) through each of the algorithms. We used the timeout script to quantify the maximum RAM used and the amount of time required to process each sequence collection (https://github.com/pshved/timeout). We limited each algorithm to 48 GB of RAM, 50 hours, and unless otherwise specified, a single processor.

***Data and code availability.*** The workflow utilized commands in GNU make (v.3.81), GNU bash (v.4.1.2), mothur [v.1.39.0; @Schloss2009], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015]. A reproducible version of this manuscript and analysis is available at https://github.com/SchlossLab/Westcott_OptiClust_mSystems_2015.

\newpage

## Figures

**Figure 1. OptiClust performance.**
Plot of MCC (A) and execution times (B) for the different starting conditions
(C) A Number of steps required to converge with different thresholds

**Figure 2.** MCC values (A) and their coefficient of variation (B) and the number of observed OTUs (C) for comparison of *de novo* clustering algorithms when applied to four natural and two synthetic datasets. For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time. The median of 10 re-orderings of the data is presented for each method and dataset.

**Figure 3.** Demonstration of how execution time (A) and memory usage (B) scale with the number of unique sequences for each clustering algorithm. For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.

**Figure 4.** Comparison of cluster.split and cluster for average neighbor, VSEARCH-based abundance-based greedy clustering, and OptiClust.

**Supplemental text.** Worked example of how OptiClust algorithm clusters sequences into OTUs.

\newpage

**Table 1. Description of datasets used to evaluate the OptiClust algorithm and compare its performance to other algorithms.**

## References

### 2016-10-22 code/build_performance_figure.R

@@ -58,9 +58,6 @@ tidy$method <- factor(tidy$method, method_ordering$method)
 tidy$dataset <- factor(tidy$dataset, datasets)
 tidy$metric <- factor(tidy$metric, metrics)
 
-write.table(tidy, file="data/processed/performance_data.tsv", row.names=F, sep='\t', quote=F)
-
-
 my_theme <- theme_classic() +
 	theme(
 		axis.text.x = element_blank(),#(angle = 90, hjust = 1, vjust=0.5, size=7),

### 2016-10-22 submission/Westcott_OptiClust_mSystems_2016.Rmd
Get manuscript to render to a PDF

@@
 When we seeded the OptiClust algorithm with each sequence in a separate OTU and ran the algorithm until complete convergence, the MCC values were `r round(mcc_rel_an, 1)` and `r round(mcc_rel_vdgc, 1)`% higher than the OTUs using average neighbor and distance-based greedy clustering (DGC) with VSEARCH, respectively (Figure 1).
-The number of OTUs formed by the various methods was negatively correlated with their MCC value (\rho=`r round(mcc_sobs_cor$estimate, 2)`; p=`r round(mcc_sobs_cor$p.value, 2)`).
+The number of OTUs formed by the various methods was negatively correlated with their MCC value ($\rho$=`r round(mcc_sobs_cor$estimate, 2)`; p=`r round(mcc_sobs_cor$p.value, 2)`).
 The OptiClust algorithm was considerably faster than the hierarchical algorithms and somewhat slower than the heuristic-based algorithms.
@@
 The additional steps required between `r paste(format(round(range(steps$ratio_secs), 1), nsmall=1L), collapse=" and ")` times longer to complete the algorithm.
-These results suggest that achieving full convergence of the optimization metric adds computational effort; however, considering full convergence took between `r
-paste(round(range(steps$med_full_secs/60), 0), collapse=" and ")` minutes the extra effort was relatively small.
+These results suggest that achieving full convergence of the optimization metric adds computational effort; however, considering full convergence took between `r paste(round(range(steps$med_full_secs/60), 0), collapse=" and ")` minutes the extra effort was relatively small.
 Although the mothur's default setting is partial convergence, the remainder of our analysis used complete convergence to be more conservative.

### 2016-10-24 submission/Westcott_OptiClust_mSystems_2016.Rmd
First full draft of Discussion section

@@
 ```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
-library(dplyr)
+library(dplyr, warn.conflicts=FALSE)
 library(tidyr)
@@
 These counts are used to calculate the optimization metric.
+
+***Optimization of clustering using composite metrics significantly improves clustering quality (Figure 2, 3, 4).*** There are multiple metrics available to assess clustering quality.
+The mothur-based implementation of OptiClust allows the user to use the MCC, F1 score, accuracy, sensitivity, specificity, and the sum of true positives and negatives.
+The MCC, F1 score, and accuracy are preferred because each of them incorporates all four values from the confusion matrix while others only utilize two values.
+It is relatively straightforward to implement other metrics.
+**Comparison of MCC, accuracy, and F1 score to each other for each dataset using the full set of sequences.** Each metric outperformed the observed values for the other clustering algorithms indicating that regardless of the metric one uses to evaluate clustering quality, the OptiClust algorithm generates better OTU assignments than any of the other methods.
+The number of OTUs generated by an algorithm is often used as a metric for clustering quality.
+We did not observe a significant correlation between clustering quality as measured by MCC, F1 score, or accuracy and the number of OTUs generated by each algorithm.
+**Interestingly, the values of the MCC, F1 score, and accuracy generated when using each metric to optimize clustering were similar across implementations.** Based on these results and its previous use in the OTU assignment literature, the remainder of our analysis uses the MCC metric for optimization.
 
@@
 ## Discussion
-Restate most important contributions
-* Optimized clustering based on an objective criteria. Results are meaningfully better than existing methods
-* Added benefits include smaller CPU and RAM footprint
-* Result is efficient analysis of large datasets without sacrificing clustering quality as has been experienced in heuristic methods.
+Myriad methods have been proposed for assigning 16S rRNA gene sequences to OTUs that each claim improved performance based on speed, memory usage, representation of taxonomic information, and number of OTUs.
+Each of these metrics is subjective and do not actually indicate the quality of the clustering.
+This led us to propose using the MCC as a metric for assessing the quality of clustering, post hoc.
+Here, we described a new clustering method that seeks to optimize clustering based on an objective criterion that measures clustering quality, a priori.
+In the OptiClust algorithm clustering is driven by  optimizing a metric that assesses whether any two sequences should be grouped into the same OTU.
+The result are clusters that are significantly more robust and is efficient in the time and memory required to cluster the sequences into OTUs.
+This makes it more tractable to analyze large datasets without sacrificing clustering quality as was previously necessary using heuristic methods.
 
-Choice of objective criteria
-* Value of using a metric that is based on all four parameters
-* Preference is for Matthew's Correlation Coefficient because... other options include F1 Score and accuracy
+The cluster optimization procedure is dependent on the metric that is chosen for optimization.
+We employed the MCC because it includes the four values from a confusion matrix.
+Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled.
+Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1 score, which is based on three values from the confusion matrix.
+Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure.
+For example, optimizing for sensitivity could lead to a large number of FPs.
+Since we would like to minimize both FPs and FNs and not just the total number of false assignments, we decided to optimize utilizing the MCC.
+The MCC measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix.
+It is possible that other metrics could be developed and employed for optimization of the clustering.
 
-Preference for OptiClust over cluster splitting approach
-* Risks are mis-classification and artificially splitting similar sequences between OTUs because they classify to different taxa
-* Still potential to merge the methods for more efficient processing.
+The OptiClust algorithm is relatively simple.
+For each sequence it effectively ask whether the MCC value will increase if the sequence is moved to a different OTU including creating a new OTU.
+If the value does not change, it remains in the current OTU.
+The algorithm repeats until the MCC value stabilizes.
+Assuming that the algorithm is seeded with each sequence in a separate OTU, it does not appear that the algorithm converges to a local optimum.
+Furthermore, execution of the algorithm with different random number generator seeds produces OTU assignments of consistently high quality.
+Future improvements to the implementation of the algorithm could provide optimization to further improve its speed and susceptibility to find a local optimum.
+Users are encourage to repeat the OTU assignment several times to confirm that they have found the best OTU assignments.
 
-Data quality - analysis uses unique sequences and we should not expect an infinite number of unique sequences unless there is a large amount of random sequencing error.
+Our previous MCC-based analysis of numerous clustering algorithms indicated that the average neighbor algorithm consistently produced the best OTU assignments with the DGC-based method using USEARCH also producing robust OTU assignments.
+The challenge in using the average neighbor algorithm is that it requires a large amount of RAM and is computationally demanding.
+This led to the development of a splitting approach that divides the clustering across distinct taxonomic groups.
+The improved performance provided by the OptiClust algorithm likely makes such splitting unnecessary for most current datasets.
+We have demonstrated that although the OTU assignments made at the genus level are still better than that of other methods, the quality is not as good as that found without splitting.
+The loss of quality is likely due to misclassification because of limitations in the clustering algorithms and reference databases.
+The practical significance of such small differences in clustering quality remain to be determined; however, based on the current analysis, it does appear that the number of OTUs is artificially inflated.
+Regardless, the best clustering quality should be pursued given the available computer resources.
 
-Based on our model, we propose the following...
+The time required to execute the OptiClust algorithm scaled proportionally to the number of unique sequences raised to the second or third power and the amount of RAM required scaled to the second power.
+The power for the time requirement is affected by the similarity of the sequences in the dataset with datasets containing more similar sequences having a higher power.
+Also, the number of unique sequences is the basis for both the amount of time and memory required to complete the algorithm.
+Both the similarity of sequences and number of unique sequences can be driven by the sequencing error since any errors will increase the number of unique sequences and these sequences will be closely related to the perfect sequence.
+This underscores the importance of reducing the noise in the sequence data.
+If sequencing errors are not remediated and are relatively randomly distributed, then it is likely that the algorithm will require an unnecessary amount of time and RAM to complete.
 
-***Optimization of clustering using composite metrics significantly improves clustering quality (Figure 2, 3, 4).*** There are multiple metrics available to assess clustering quality.
-The mothur-based implementation of OptiClust allows the user to use the MCC, F1 score, accuracy, sensitivity, specificity, and the sum of true positives and negatives.
-The MCC, F1 score, and accuracy are preferred because each of them incorporates all four values from the confusion matrix while others only utilize two values.
-It is relatively straightforward to implement other metrics.
-**Comparison of MCC, accuracy, and F1 score to each other for each dataset using the full set of sequences.** Each metric outperformed the observed values for the other clustering algorithms indicating that regardless of the metric one uses to evaluate clustering quality, the OptiClust algorithm generates better OTU assignments than any of the other methods.
-The number of OTUs generated by an algorithm is often used as a metric for clustering quality.
-We did not observe a significant correlation between clustering quality as measured by MCC, F1 score, or accuracy and the number of OTUs generated by each algorithm.
-**Interestingly, the values of the MCC, F1 score, and accuracy generated when using each metric to optimize clustering were similar across implementations.** Based on these results and its previous use in the OTU assignment literature, the remainder of our analysis uses the MCC metric for optimization.
+The rapid expansion in sequencing capacity has demonstrated that database-independent assignment of sequences into OTUs is a critical step in the sequence analysis pipeline.
+Of course, the quality of the assignments is dependent on the quality of the underlying data.
+The ability to objectively optimize the clustering quality will significantly enhance the quality of those OTU assignments while making the time and memory requirements more efficient.
+The development of the OptiClust algorithm represents a significant advance that likely has numerous other applications.
 

### 2016-10-24 submission/Westcott_OptiClust_mSystems_2016.Rmd
Redo final paragraph of Discussion

@@
 
-The rapid expansion in sequencing capacity has demonstrated that database-independent assignment of sequences into OTUs is a critical step in the sequence analysis pipeline.
-Of course, the quality of the assignments is dependent on the quality of the underlying data.
-The ability to objectively optimize the clustering quality will significantly enhance the quality of those OTU assignments while making the time and memory requirements more efficient.
-The development of the OptiClust algorithm represents a significant advance that likely has numerous other applications.
+The rapid expansion in sequencing capacity has demanded that the algorithms used to assign 16S rRNA gene sequences to OTUs be efficient while maintaining robust assignments.
+Although database-based approaches have been proposed to facilitate this analysis, they are limited by their limited coverage of bacterial taxonomy and by the inconsistent process used to name taxa.
+Thus, the ability to assign sequences to OTUs with an a priori focus on clustering quality will significantly enhance downstream analysis.
+The development of the OptiClust algorithm represents a significant advance that is likely to have numerous other applications.
 

### 2016-10-24 submission/Westcott_OptiClust_mSystems_2016.Rmd
Edit Methods section

@@
 These sequence collections were all generated using paired 150 or 250 nt reads of the V4 region.
-We re-processed all of the reads using a common analysis pipeline that included quality score-based error correction [@Kozich2013], alignment against a SILVA reference database [@Pruesse2007; @Schloss2010], screening for chimeras using UCHIME [@Edgar2011], and classification using a naive Bayesian classifier with the RDP training set [@Wang2007].
+We re-processed all of the reads using a common analysis pipeline that included quality score-based error correction [@Kozich2013], alignment against a SILVA reference database [@Pruesse2007; @Schloss2010], screening for chimeras using UCHIME [@Edgar2011], and classification using a naive Bayesian classifier with the RDP training set requiring an 80% confidence score [@Wang2007].
 
@@
 These included three hierarchical algorithms, average neighbor (AN), nearest neighbor (NN), and furthest neighbor (FN), which are implemented in mothur [v.1.39.0; @Schloss2009].
-Seven heuristic methods were also used including abundance-based greedy clustering (UAGC) and distance-based greedy clustering (UDGC) as implemented in USEARCH [v.6.1; @Edgar2010], abundance-based greedy clustering (VAGC) and distance-based greedy clustering (VDGC) as implemented in VSEARCH [v.X.X.X; @Rognes2015], OTUClust [v.X.X.X; @XXXXXX], SumaClust [v.X.X.X; @XXXXXX], and Swarm [v.2.1.1; @Mah2014].
+Seven heuristic methods were also used including abundance-based greedy clustering (AGC) and (distance-based greedy clustering) DGC as implemented in USEARCH [v.6.1; @Edgar2010] and VSEARCH [v.X.X.X; @Rognes2015], OTUClust [v.X.X.X; @XXXXXX], SumaClust [v.X.X.X; @XXXXXX], and Swarm [v.2.1.1; @Mah2014].
 With the exception of Swarm each of these methods uses distance-based thresholds to report OTU assignments.
-To judge the quality of the Swarm-generated OTU assignments we calculated the MCC value using thresholds incremented by 1% between 0 and 5% and selected the threshold that provided the optimal MCC value [@Westcott2015].
-We also assessed the ability of a previously describe heuristic to cluster sequences using the UDGC and OptiClust algorithms.
-In this heuristic sequences are split into bacterial families based on sequence classification and clustered within the taxonomic family [@Westcott20XX].
-Finally, for our benchmarking analysis, we evaluate the memory and time requirements when using 8 processors with the UDGC and VDGC algorithms and with the taxonomic splitting heuristic.
 
@@
 Pairs of sequences that were in the same OTU and had a distance less than 3% were true positives (TPs), those that were in different OTUs and had a distance greater than 3% were true negatives (TNs), those that were in the same OTU and had a distance greater than 3% were false positives (FPs), and those that were in different OTUs and had a distance less than 3% were false negatives (FNs).
-To synthesize the matrix into a single metric we used the Matthew's Correlation Coefficient, F1 score, and accuracy using the `sens.spec` command in mothur using the following equations.
+To synthesize the matrix into a single metric we used the Matthews Correlation Coefficient using the `sens.spec` command in mothur using the following equations.
 
@@
 
-$$
-F1_score = \frac{2 \times TP}{2 \times TP+FP+FN}
-$$
-
-$$
-Accuracy = \frac{TP + TN}{(TP+FP+TN+FN)}
-$$
-
 To assess the reproducibility of the algorithms we randomized the starting order of each sequence collection ten times and ran each algorithm on each randomized collection.
-We then measured the MCC, F1 score, and accuracy for each randomization and quantified their coefficient of variation (CV; the ratio of the standard deviation to the mean).
+We then measured the MCC for each randomization and quantified their percent coefficient of variation (% CV; 100 times the ratio of the standard deviation to the mean).
 
@@
 We used the timeout script to quantify the maximum RAM used and the amount of time required to process each sequence collection (https://github.com/pshved/timeout).
-We limited each algorithm to 48 GB of RAM, 50 hours, and unless otherwise specified, a single processor.
+We limited each algorithm to 48 GB of RAM and 50 hours using a single processor.
 
 ***Data and code availability.*** The workflow utilized commands in GNU make (v.3.81), GNU bash (v.4.1.2), mothur [v.1.39.0; @Schloss2009], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015].
+Within R we utilized the wesanderson [v.X.X.X; @XXXXXX], dplyr [v.X.X.X; @XXXXXX], tidyr [v.X.X.X; @XXXXXX], cowplot [v.X.X.X; @XXXXXX], and ggplot2 [v.X.X.X; @XXXXXX] packages.
 A reproducible version of this manuscript and analysis is available at https://github.com/SchlossLab/Westcott_OptiClust_mSystems_2015.

### 2016-10-25 submission/Westcott_OptiClust_mSystems_2016.Rmd
Update figure legends

@@
 
-**Figure 1. OptiClust performance.**
-Plot of MCC (A) and execution times (B) for the different starting conditions
-(C) A Number of steps required to converge with different thresholds
-
-**Figure 2.** MCC values (A) and their coefficient of variation (B) and the number of observed OTUs (C) for comparison of *de novo* clustering algorithms when applied to four natural and two synthetic datasets.
+**Figure 1.
+Comparison of de novo clustering algorithms.**  Plot of MCC (A), number of OTUs (B), and execution times (C) for the comparison of *de novo* clustering algorithms when applied to four natural and two synthetic datasets.
+The first three columns of each figure contain the results of clustering the datasets (i) seeding the algorithm with one sequence per OTU and allowing the algorithm to proceed until the MCC value no longer changed; (ii) seeding the algorithm with one sequence per OTU and allowing the algorithm to proceed until the MCC changed by less than 0.0001; (iii) seeding the algorithm with all of the sequences in one OTU and allowing the algorithm to proceed until the MCC value no longer changed.
 For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.
 The median of 10 re-orderings of the data is presented for each method and dataset.
+The confidence interval is indicated by the error bars, which are typically smaller than the plotting symbol.
 
-**Figure 3.** Demonstration of how execution time (A) and memory usage (B) scale with the number of unique sequences for each clustering algorithm.
-For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.
+**Figure 2.
+OptiClust performance** The average execution time (A) and memory usage (B) required to cluster the four natural datasets.
+The confidence intervals indicate the range between the minimum and maximum values.
+The y-axis is scaled by the square root to demonstrate the relationship between the time and memory requirements relative to the number of unique sequences squared.
 
-**Figure 4.** Comparison of cluster.split and cluster for average neighbor, VSEARCH-based abundance-based greedy clustering, and OptiClust.
+**Figure 3.
+Effects of taxonomically splitting the datasets on clustering quality.** The datasets were split at each taxonomic level based on their classification using a naive Bayesian classifier and clustered using average neighbor, VSEARCH-based AGC, and OptiClust.
 

### 2016-10-25 submission/Westcott_OptiClust_mSystems_2016.Rmd
More editing

@@
 The human dataset was a challenge for a number of the algorithms.
-OTUCLUST and SumaClust were unable to cluster the human dataset in less than 50 hours and the average neighbor algorithm required more than 48 GB of RAM.
+OTUCLUST and SumaClust were unable to cluster the human dataset in less than 50 hours and the average neighbor algorithm required more than 45 GB of RAM.
 The USEARCH-based methods were unable to cluster the human data using the 32-bit free version of the software that limits the amount of RAM to approximately 3.5 GB.
@@
 However, as datasets become larger, it may be necessary to use the heuristic to clustering the data into OTUs.
-For example, we were unable to cluster the full human data using less than 48 GB of RAM; however, when we split the dataset at the family level, we were able to cluster the data with the limited resources.
+For example, we were unable to cluster the full human data using less than 45 GB of RAM; however, when we split the dataset at the family level, we were able to cluster the data with the limited resources.
 
@@
 We used the timeout script to quantify the maximum RAM used and the amount of time required to process each sequence collection (https://github.com/pshved/timeout).
-We limited each algorithm to 48 GB of RAM and 50 hours using a single processor.
+We limited each algorithm to 45 GB of RAM and 50 hours using a single processor.
 
@@
 The first three columns of each figure contain the results of clustering the datasets (i) seeding the algorithm with one sequence per OTU and allowing the algorithm to proceed until the MCC value no longer changed; (ii) seeding the algorithm with one sequence per OTU and allowing the algorithm to proceed until the MCC changed by less than 0.0001; (iii) seeding the algorithm with all of the sequences in one OTU and allowing the algorithm to proceed until the MCC value no longer changed.
-For the purposes of this analysis, clustering was limited to 48 GB of RAM and 50 hours of execution time.
+The human dataset could not be clustered by the average neighbor, Sumaclust, USEARCH, or OTUCLUST with less than 45 GB of RAM or 50 hours of execution time.
 The median of 10 re-orderings of the data is presented for each method and dataset.

### 2016-10-25 submission/Westcott_OptiClust_mSystems_2016.Rmd
More editing

@@
 **Figure 3.
-Effects of taxonomically splitting the datasets on clustering quality.** The datasets were split at each taxonomic level based on their classification using a naive Bayesian classifier and clustered using average neighbor, VSEARCH-based AGC, and OptiClust.
+Effects of taxonomically splitting the datasets on clustering quality.** The datasets were split at each taxonomic level based on their classification using a naive Bayesian classifier and clustered using average neighbor, VSEARCH-based DGC, and OptiClust.
 

### 2016-10-25 submission/Westcott_OptiClust_mSystems_2016.Rmd
First full draft of Introduction

@@
 Amplicon-based sequencing has provided incredible insights into Earth's microbial biodiversity.
+It has become common for studies to include sequencing millions of 16S rRNA gene sequences across hundreds of samples.
+This is three to four orders of magnitude greater sequencing depth than was previously achieved using Sanger sequencing.
+The increased sequencing depth has highlighted the limitations of existing bioinformatics tools for analyzing these sequences.
+Furthermore, these sequencing efforts have revealed novel taxonomic diversity that are not adequately represented in reference databases.
+Because of this, a critical component to the analysis of microbial biodiversity has been the assignment of 16S rRNA gene sequences and other marker gene sequences to operational taxonomic units (OTUs) without reliance on taxonomic classification.
 
-Numerous methods have been proposed
+A growing number of algorithms have been developed to assign sequences into OTUs and these can be clustered into three general categories.
+The first category of algorithms are classification-based and have been termed closed-reference or phylotyping.
+Sequences are compared to a reference collection and clustered based on the reference sequence that they are similar to.
+This approach is fast; however, struggles when a sequence is similar to multiple reference sequences that may have different taxonomies.
+The second category of algorithm have been called *de novo* because they assign sequences to OTUs without the use of a reference.
+These include hierarchical algorithms such as nearest, furthest, and average neighbor and methods that employ heuristics such as abundance or distance-based greedy clustering as implemented in USEARCH or VSEARCH, Sumaclust, OTUCLUST, and Swarm.
+These methods tend to be more computationally intense and it has proven difficult to know which method generates the best assignments.
+A third category of algorithm is a hybrid approach or open-reference clustering.
+Here sequences are assigned to OTUs using closed-reference clustering and sequences that are not within a threshold of a reference sequence are then clustered using a *de novo* approach.
+This category blends the strengths and weaknesses of the other method and adds the complication that closed-reference and *de novo* clustering use different OTU definitions.
 
-Needed an objective criteria to evaluate the various methods methods
-
-Average neighbor is consistently the best method, but others including USEARCH and VSEARCH resulted in comparable results
-
-Difficulty with amount of memory and time required to complete average neighbor algorithm is a significant hurdle to completion of the method.
-
-Instead of retrospectively evaluating methods for their ability to correctly assign sequences we sought to develop a method that would prospectively assign sequences to OTUs to optimize classification metrics.
-
-Clustering quality within the algorithm is assessed by counting the number of true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN) based on the pairwise distances.
+Several metrics have emerged for assessing the quality of OTU assignment algorithms.
+These have included the time and memory required to run the algorithm, agreement between OTU assignments and the sequences' taxonomy, sensitivity of an algorithm to stochastic processes, and the number of OTUs generated by the algorithm.
+Unfortunately, these methods fail to directly quantify the quality of the OTU assignments.
+An algorithm may complete with minimal time and memory requirements or generate an idealized number of OTUs, but the composition of the OTUs could be incorrect.
+These metrics also tend to be subjective.
+For instance, a method may appear to be recapitulate the taxonomy of a synthetic community with known taxonomic structure, but do a poor job when applied to real communities with poorly defined taxonomic structure or for sequences that are prone to misclassification.
+As an alternative, we developed an approach to objectively benchmark the clustering quality of OTU assignments.
+This approach counts the number of true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN) based on the pairwise distances.
 Sequence pairs that are within the user-specified threshold and are clustered together represent TPs and those in different OTUs are FNs.
 Those sequence pairs that have a distance larger than the threshold and are not clustered in the same OTU are TNs and those in the same OTU are FPs.
-These counts are used to calculate the optimization metric.
+These values can be synthesized into a single correlation coefficient, the Matthews' Correlation Coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix.
 
-***Optimization of clustering using composite metrics significantly improves clustering quality (Figure 2, 3, 4).*** There are multiple metrics available to assess clustering quality.
-The mothur-based implementation of OptiClust allows the user to use the MCC, F1 score, accuracy, sensitivity, specificity, and the sum of true positives and negatives.
-The MCC, F1 score, and accuracy are preferred because each of them incorporates all four values from the confusion matrix while others only utilize two values.
-It is relatively straightforward to implement other metrics.
-**Comparison of MCC, accuracy, and F1 score to each other for each dataset using the full set of sequences.** Each metric outperformed the observed values for the other clustering algorithms indicating that regardless of the metric one uses to evaluate clustering quality, the OptiClust algorithm generates better OTU assignments than any of the other methods.
-The number of OTUs generated by an algorithm is often used as a metric for clustering quality.
-We did not observe a significant correlation between clustering quality as measured by MCC, F1 score, or accuracy and the number of OTUs generated by each algorithm.
-**Interestingly, the values of the MCC, F1 score, and accuracy generated when using each metric to optimize clustering were similar across implementations.** Based on these results and its previous use in the OTU assignment literature, the remainder of our analysis uses the MCC metric for optimization.
+We previously used this approach to assess the OTU assignments of a variety of algorithms using synthetic and real datasets.
+Consistently, the average neighbor algorithm was identified as among the best or the best algorithm.
+The distance-based greedy clustering as implemented in VSEARCH has also performed well.
+The computational resources required to complete the average neighbor algorithm can be significant for large datasets and so there is a need for an algorithm that efficiently produces consistently high quality OTU assignments.
+Previously, we assessed the quality of the clusters after the completion of the algorithm or post hoc.
+In the current study we developed and benchmarked a new de novo clustering algorithm.
+Instead of assessing the methods post hoc, the new method uses a priori calculation of the MCC to direct the algorithm.
+The result is the OptiClust algorithm, which produces significantly better sequence assignments while making efficient use of computational resources.
 
@@
 Since we would like to minimize both FPs and FNs and not just the total number of false assignments, we decided to optimize utilizing the MCC.
-The MCC measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix.
 It is possible that other metrics could be developed and employed for optimization of the clustering.

### 2016-10-25 code/summarize_datasets.R
Create a metadata file to summarize each dataset

@@ -1,0 +1,21 @@
+dataset <- c('soil', 'marine', 'human', 'mice', 'staggered', 'even')
+
+count_file_names <- paste0('data/', dataset, '/', dataset, '.count_table')
+
+get_stats <- function(count_file_name) {
+	count_data <- read.table(file=count_file_name, header=T)
+
+	n_unique_seqs <- nrow(count_data)
+	n_total_seqs <- sum(count_data$total)
+
+	n_samples <- ncol(count_data) - 2
+	if(n_samples == 0) n_samples <- NA
+
+	c(n_samples=n_samples, n_unique_seqs=n_unique_seqs, n_total_seqs=n_total_seqs)
+}
+
+summary_table <- t(sapply(count_file_names, get_stats))
+rownames(summary_table) <- dataset
+
+write.table(summary_table, file="data/processed/datasets.summary", quote=F, sep='\t')
+

### 2016-10-25 results/table/table_1.Rmd
Build Table 1 which contains metadata summary

---
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: table_header.tex
tables: true
geometry: margin=0.75in
---

**Table 1. Description of datasets used to evaluate the OptiClust algorithm and compare its performance to other algorithms.** Each dataset contains sequences from the V4 region of the 16S rRNA gene. The even and staggered datasets were generated by extracting the V4 region from full length reference sequences and the datasets from the natural communities were generated by sequencing the V4 region using a Illumina MiSeq with either paired 150 or 250 nt reads.

```{r results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
library('xtable')

options(xtable.sanitize.text.function=identity)

dataset <- c(
							soil = "Soil",
							marine = "Marine",
							mice = "Mice",
							human = "Human",
							even = "Even",
							staggered = "Staggered"
						)

sequencing <- c(
							soil = "150",
							marine = "250",
							mice = "250",
							human = "250",
							even = "NA",
							staggered = "NA"
						)

reference <-  c(
							soil = "XX",
							marine = "XX",
							mice = "XX",
							human = "XX",
							even = "XX,YY",
							staggered = "XX,YY"
						)

data_summary <- read.table(file="data/processed/datasets.summary", header=T, na.string="", stringsAsFactors=F)
data_summary <- data_summary[names(reference),]

stopifnot(names(dataset) == rownames(data_summary))
stopifnot(names(sequencing) == rownames(data_summary))
stopifnot(names(reference) == rownames(data_summary))

dataset_ref <- paste0(dataset, ' (', reference, ')')

data_table <- cbind(dataset_ref, sequencing, data_summary$n_samples,
 										format(data_summary$n_total_seqs, big.mark=','),
 										format(data_summary$n_unique_seqs, big.mark=','))

colnames(data_table) <- c("Dataset (Ref.)",
													"\\bigcell{c}{Read \\\\ Length }",
													"\\bigcell{c}{Samples \\\\ (N) }",
													"\\bigcell{c}{Total \\\\ Seqs. (N)}",
													"\\bigcell{c}{Unique \\\\ Seqs. (N) }")

colnames(data_table) <- paste0("\\textbf{", colnames(data_table), "}")

composite <- xtable(data_table, label="")

align(composite) <- "llcccc"

print(composite,
		include.rownames=FALSE,
		include.colnames=TRUE,
		hline.after = c(-1,0, nrow(composite)),
		comment=FALSE,
		table.placement="!ht"
		)
```

### 2016-12-22 submission/Westcott_OptiClust_mSystems_2016.Rmd
Build Table 1 which contains metadata summary
Editing of manuscript

@@
 Amplicon-based sequencing has provided incredible insights into Earth's microbial biodiversity.
-It has become common for studies to include sequencing millions of 16S rRNA gene sequences across hundreds of samples.
-This is three to four orders of magnitude greater sequencing depth than was previously achieved using Sanger sequencing.
-The increased sequencing depth has highlighted the limitations of existing bioinformatics tools for analyzing these sequences.
-Furthermore, these sequencing efforts have revealed novel taxonomic diversity that are not adequately represented in reference databases.
-Because of this, a critical component to the analysis of microbial biodiversity has been the assignment of 16S rRNA gene sequences and other marker gene sequences to operational taxonomic units (OTUs) without reliance on taxonomic classification.
+It has become common for studies to include sequencing millions of 16S rRNA gene sequences across hundreds of samples [ref].
+This is three to four orders of magnitude greater sequencing depth than was previously achieved using Sanger sequencing [ref].
+The increased sequencing depth has revealed novel taxonomic diversity that is not adequately represented in reference databases [ref].
+However, the advance has forced re-engineering of methods to overcome the rate and memory limiting steps in computational pipelines that process raw sequences through the generation of tables containing the number of sequences in different taxa for each sample [ref].
+A critical component to these pipelines has been the assignment of amplicon sequences to taxonomic units that are ether defined based on similarity to a reference or operationally based on the similarity of the sequences to each other within the dataset.
 
-A growing number of algorithms have been developed to assign sequences into OTUs and these can be clustered into three general categories.
-The first category of algorithms are classification-based and have been termed closed-reference or phylotyping.
-Sequences are compared to a reference collection and clustered based on the reference sequence that they are similar to.
-This approach is fast; however, struggles when a sequence is similar to multiple reference sequences that may have different taxonomies.
-The second category of algorithm have been called *de novo* because they assign sequences to OTUs without the use of a reference.
-These include hierarchical algorithms such as nearest, furthest, and average neighbor and methods that employ heuristics such as abundance or distance-based greedy clustering as implemented in USEARCH or VSEARCH, Sumaclust, OTUCLUST, and Swarm.
-These methods tend to be more computationally intense and it has proven difficult to know which method generates the best assignments.
-A third category of algorithm is a hybrid approach or open-reference clustering.
+A growing number of algorithms have been developed to cluster sequences into OTUs.
+These algorithms can be classified into three general categories.
+The first category of algorithms has been termed closed-reference or phylotyping [ref].
+Sequences are compared to a reference collection and clustered based on the reference sequences that they are similar to.
+This approach is fast; however, the method struggles when a sequence is similar to multiple reference sequences that may have different taxonomies and when it is not similar to sequences in the reference.
+The second category of algorithms has been called *de novo* because they assign sequences to OTUs without the use of a reference [ref].
+These include hierarchical algorithms such as nearest, furthest, and average neighbor and algorithms that employ heuristics such as abundance or distance-based greedy clustering as implemented in USEARCH or VSEARCH, Sumaclust, OTUCLUST, and Swarm [ref].
+*De novo* methods tend to be more computationally intense and it has proven difficult to know which method generates the best assignments.
+A third category of algorithm is open-reference clustering, which is a hybrid approach [ref].
 Here sequences are assigned to OTUs using closed-reference clustering and sequences that are not within a threshold of a reference sequence are then clustered using a *de novo* approach.
 This category blends the strengths and weaknesses of the other method and adds the complication that closed-reference and *de novo* clustering use different OTU definitions.
+These algorithms take different approaches to handling large datasets to minimize the time and memory requirements while attempting to assign sequences to meaningful OTUs.
 
 Several metrics have emerged for assessing the quality of OTU assignment algorithms.
-These have included the time and memory required to run the algorithm, agreement between OTU assignments and the sequences' taxonomy, sensitivity of an algorithm to stochastic processes, and the number of OTUs generated by the algorithm.
+These have included the time and memory required to run the algorithm [ref], agreement between OTU assignments and the sequences' taxonomy [ref], sensitivity of an algorithm to stochastic processes [ref], and the number of OTUs generated by the algorithm [ref].
 Unfortunately, these methods fail to directly quantify the quality of the OTU assignments.
@@
 For instance, a method may appear to be recapitulate the taxonomy of a synthetic community with known taxonomic structure, but do a poor job when applied to real communities with poorly defined taxonomic structure or for sequences that are prone to misclassification.
-As an alternative, we developed an approach to objectively benchmark the clustering quality of OTU assignments.
+As an alternative, we developed an approach to objectively benchmark the clustering quality of OTU assignments [ref].
 This approach counts the number of true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN) based on the pairwise distances.
@@
 Those sequence pairs that have a distance larger than the threshold and are not clustered in the same OTU are TNs and those in the same OTU are FPs.
-These values can be synthesized into a single correlation coefficient, the Matthews' Correlation Coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix.
-
-We previously used this approach to assess the OTU assignments of a variety of algorithms using synthetic and real datasets.
+These values can be synthesized into a single correlation coefficient, the Matthews' Correlation Coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix [ref].
 Consistently, the average neighbor algorithm was identified as among the best or the best algorithm.
@@
 The computational resources required to complete the average neighbor algorithm can be significant for large datasets and so there is a need for an algorithm that efficiently produces consistently high quality OTU assignments.
-Previously, we assessed the quality of the clusters after the completion of the algorithm or post hoc.
-In the current study we developed and benchmarked a new de novo clustering algorithm.
-Instead of assessing the methods post hoc, the new method uses a priori calculation of the MCC to direct the algorithm.
+
+These previous efforts have assessed the quality of the clusters after the completion of the algorithm.
+In the current study we developed and benchmarked a new *de novo* clustering algorithm that uses real time calculation of the MCC to direct the progress of the clustering.
 The result is the OptiClust algorithm, which produces significantly better sequence assignments while making efficient use of computational resources.
@@
 
-The OptiClust algorithm uses the metric that should be used to assess clustering quality, a list of all sequence names in the dataset, and the pairs of sequences that are within a desired threshold of each other (e.g.
-0.03).
+The OptiClust algorithm uses the pairs of sequences that are within a desired threshold of each other (e.g.
+0.03), a list of all sequence names in the dataset, and the metric that should be used to assess clustering quality.
 A detailed description of the algorithm is provided for a toy dataset in the Supplementary Material.
@@
 mcc_sobs_cor <- cor.test(perf$mean_num_otus, perf$mean_mcc, method="spearman")
+
+stopifnot(round(1/secs_rel_vdgc, 1) == 1)
 ```
@@
 To evaluate the OptiClust algorithm and compare its performance to other algorithms, we utilized six datasets including two synthetic communities and four previously published large datasets generated from soil, marine, human, and murine samples (Table 1).
-When we seeded the OptiClust algorithm with each sequence in a separate OTU and ran the algorithm until complete convergence, the MCC values were `r round(mcc_rel_an, 1)` and `r round(mcc_rel_vdgc, 1)`% higher than the OTUs using average neighbor and distance-based greedy clustering (DGC) with VSEARCH, respectively (Figure 1).
+When we seeded the OptiClust algorithm with each sequence in a separate OTU and ran the algorithm until complete convergence, the MCC values averaged `r round(mcc_rel_an, 1)` and `r round(mcc_rel_vdgc, 1)`% higher than the OTUs using average neighbor and distance-based greedy clustering (DGC) with VSEARCH, respectively (Figure 1).
 The number of OTUs formed by the various methods was negatively correlated with their MCC value ($\rho$=`r round(mcc_sobs_cor$estimate, 2)`; p=`r round(mcc_sobs_cor$p.value, 2)`).
 The OptiClust algorithm was considerably faster than the hierarchical algorithms and somewhat slower than the heuristic-based algorithms.
-Across the six datasets, the OptiClust algorithm was `r round(secs_rel_an, 1)`-times faster than average neighbor and `r round(1/secs_rel_vdgc, 1)`-times slower than DGC with VSEARCH.
+Across the six datasets, the OptiClust algorithm was `r round(secs_rel_an, 1)`-times faster than average neighbor and just as fast as DGC with VSEARCH.
 The human dataset was a challenge for a number of the algorithms.
@@
 The USEARCH-based methods were unable to cluster the human data using the 32-bit free version of the software that limits the amount of RAM to approximately 3.5 GB.
-These data demonstrate that OptiClust generates significantly more robust OTU assignments than existing methods across a diverse collection of datasets with performance that is comparable to popular methods.
+These data demonstrate that OptiClust generated significantly more robust OTU assignments than existing methods across a diverse collection of datasets with performance that was comparable to popular methods.
 
@@
 ```{r aggregate_analysis}
-agg <- read.table(file="data/processed/cluster_data.summary", header=T) %>%
+agg <- read.table(file="data/processed/cluster_data.summary", header=T, stringsAsFactors=F) %>%
 			filter(frac==1.0, method %in% (c("mcc", "mcc_agg"))) %>%
@@
 
-As implemented within mothur, the OptiClust algorithm either starts with each sequence in a separate OTU or with all of the sequences in a single OTU.
-We repeated the complete convergence analysis, but seeded the algorithm with all sequences in a single OTU.
+By default the mothur implementation of the OptiClust algorithm starts with each sequence in a separate OTU.
+An alternative approach is to start with all of the sequences in a single OTU.
 We found that the MCC values for clusters generated seeding OptiClust with the sequences as a single OTU were between `r paste(round(range(agg$per_diff_mcc), 1), collapse=" and ")`% lower than when seeding the algorithm with sequences in separate OTUs (Figure 1).
@@
 Finally, the amount of time required to cluster the data when the algorithm was seeded with a single OTU was between `r paste(format(round(range(agg$fold_secs), 1), nsmall=1L), collapse=" and ")`-times longer than if sequences were seeded as separate OTUs.
-This analysis demonstrates that seeding the algorithm with sequences as separate OTUs results in the best OTU assignments in the shortest amount of time.
+This analysis demonstrates that seeding the algorithm with sequences as separate OTUs resulted in the best OTU assignments in the shortest amount of time.
 
@@
 Confirming our previous results, all of the methods we tested were stable to stochastic processes.
-Of the method that involved randomization, the coefficient of variation for MCC values considerably smaller than the other methods and the coefficient of variation for the number of OTUs was comparable to the other methods.
-The variation observed in clustering quality suggests that the algorithm does not appear to converge to a locally optimum MCC value.
+Of the methods that involved randomization, the coefficient of variation for MCC values considerably smaller with OptiClust than the other methods and the coefficient of variation for the number of OTUs was comparable to the other methods.
+The variation observed in clustering quality suggested that the algorithm does not appear to converge to a locally optimum MCC value.
 More importantly, the random variation does yield output of a similarly high quality.
@@
 
-We previously described a heuristic to accelerate OTU assignments where sequences were classified to taxonomic groups and within each taxon sequences were assigned to OTUs using the average neighbor clustering algorithm.
-This can accelerate the clustering and reduce the memory requirements because the number of unique sequences is effectively reduced by splitting sequences across taxonomic groups.
+We previously described a heuristic to accelerate OTU assignments where sequences were first classified to taxonomic groups and within each taxon sequences were assigned to OTUs using the average neighbor clustering algorithm [ref].
+This accelerated the clustering and reduce the memory requirements because the number of unique sequences is effectively reduced by splitting sequences across taxonomic groups.
 Furthermore, because sequences in different taxonomic groups are assumed to belong to different OTUs they are independent, which permits parallelization and additional reduction in computation time.
@@
 These decreases in MCC value resulted in the formation of as many as `r format(round(min(100*((rel_otu_data %>% filter(tax_level == 3))$max_num_otus-1), na.rm=T), 1), nsmall=1L)` and `r format(round(min(100*((rel_otu_data %>% filter(tax_level == 6))$max_num_otus-1), na.rm=T), 1), nsmall=1L)`% more OTUs, respectively, than were observed from the entire dataset.
-For the datasets included in the current analysis, the use of the cluster splitting heuristic is not worth the loss in clustering quality.
+For the datasets included in the current analysis, the use of the cluster splitting heuristic was probably not worth the loss in clustering quality.
 However, as datasets become larger, it may be necessary to use the heuristic to clustering the data into OTUs.
-For example, we were unable to cluster the full human data using less than 45 GB of RAM; however, when we split the dataset at the family level, we were able to cluster the data with the limited resources.
 
@@
 This led us to propose using the MCC as a metric for assessing the quality of clustering, post hoc.
-Here, we described a new clustering method that seeks to optimize clustering based on an objective criterion that measures clustering quality, a priori.
+Here, we described a new clustering method that seeks to optimize clustering based on an objective criterion that measures clustering quality in real time.
 In the OptiClust algorithm clustering is driven by optimizing a metric that assesses whether any two sequences should be grouped into the same OTU.
-The result are clusters that are significantly more robust and is efficient in the time and memory required to cluster the sequences into OTUs.
+The result is clusters that are significantly more robust and is efficient in the time and memory required to cluster the sequences into OTUs.
 This makes it more tractable to analyze large datasets without sacrificing clustering quality as was previously necessary using heuristic methods.
@@
 We employed the MCC because it includes the four values from a confusion matrix.
-Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled.
+Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled [ref].
 Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1 score, which is based on three values from the confusion matrix.
-Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure.
+Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure [ref].
 For example, optimizing for sensitivity could lead to a large number of FPs.
@@
 
-Our previous MCC-based analysis of numerous clustering algorithms indicated that the average neighbor algorithm consistently produced the best OTU assignments with the DGC-based method using USEARCH also producing robust OTU assignments.
+Our previous MCC-based analysis of clustering algorithms indicated that the average neighbor algorithm consistently produced the best OTU assignments with the DGC-based method using USEARCH also producing robust OTU assignments.
 The challenge in using the average neighbor algorithm is that it requires a large amount of RAM and is computationally demanding.
-This led to the development of a splitting approach that divides the clustering across distinct taxonomic groups.
+This led to the development of a splitting approach that divides the clustering across distinct taxonomic groups [ref].
 The improved performance provided by the OptiClust algorithm likely makes such splitting unnecessary for most current datasets.
@@
 
-The time required to execute the OptiClust algorithm scaled proportionally to the number of unique sequences raised to the second or third power and the amount of RAM required scaled to the second power.
+The time and memory required to execute the OptiClust algorithm scaled proportionally to the number of unique sequences raised to the second power.
 The power for the time requirement is affected by the similarity of the sequences in the dataset with datasets containing more similar sequences having a higher power.
@@
 Both the similarity of sequences and number of unique sequences can be driven by the sequencing error since any errors will increase the number of unique sequences and these sequences will be closely related to the perfect sequence.
-This underscores the importance of reducing the noise in the sequence data.
+This underscores the importance of reducing the noise in the sequence data [ref].
 If sequencing errors are not remediated and are relatively randomly distributed, then it is likely that the algorithm will require an unnecessary amount of time and RAM to complete.
@@
 Although database-based approaches have been proposed to facilitate this analysis, they are limited by their limited coverage of bacterial taxonomy and by the inconsistent process used to name taxa.
-Thus, the ability to assign sequences to OTUs with an a priori focus on clustering quality will significantly enhance downstream analysis.
+The ability to assign sequences to OTUs using an algorithm that optimizes clustering by directly measuring quality will significantly enhance downstream analysis.
 The development of the OptiClust algorithm represents a significant advance that is likely to have numerous other applications.
@@
 In addition to the OptiClust algorithm we evaluated ten different *de novo* clustering algorithms.
-These included three hierarchical algorithms, average neighbor (AN), nearest neighbor (NN), and furthest neighbor (FN), which are implemented in mothur [v.1.39.0; @Schloss2009].
+These included three hierarchical algorithms, average neighbor, nearest neighbor, and furthest neighbor, which are implemented in mothur [v.1.39.0; @Schloss2009].
 Seven heuristic methods were also used including abundance-based greedy clustering (AGC) and (distance-based greedy clustering) DGC as implemented in USEARCH [v.6.1; @Edgar2010] and VSEARCH [v.X.X.X; @Rognes2015], OTUClust [v.X.X.X; @XXXXXX], SumaClust [v.X.X.X; @XXXXXX], and Swarm [v.2.1.1; @Mah2014].
@@
 
-**Table 1. Description of datasets used to evaluate the OptiClust algorithm and compare its performance to other algorithms.**
-
 ## References

### 2016-12-22 submission/Westcott_OptiClust_mSystems_2016.Rmd
Move code for Table 1 into manuscript Rmd file

@@
 
+**Table 1.
+Description of datasets used to evaluate the OptiClust algorithm and compare its performance to other algorithms.** Each dataset contains sequences from the V4 region of the 16S rRNA gene.
+The even and staggered datasets were generated by extracting the V4 region from full length reference sequences and the datasets from the natural communities were generated by sequencing the V4 region using a Illumina MiSeq with either paired 150 or 250 nt reads.
+
+```{r results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
+library('xtable')
+
+options(xtable.sanitize.text.function=identity)
+
+dataset <- c(
+							soil = "Soil",
+							marine = "Marine",
+							mice = "Mice",
+							human = "Human",
+							even = "Even",
+							staggered = "Staggered"
+						)
+
+sequencing <- c(
+							soil = "150",
+							marine = "250",
+							mice = "250",
+							human = "250",
+							even = "NA",
+							staggered = "NA"
+						)
+
+reference <-  c(
+							soil = "XX",
+							marine = "XX",
+							mice = "XX",
+							human = "XX",
+							even = "XX,YY",
+							staggered = "XX,YY"
+						)
+
+data_summary <- read.table(file="data/processed/datasets.summary", header=T, na.string="", stringsAsFactors=F)
+data_summary <- data_summary[names(reference),]
+
+stopifnot(names(dataset) == rownames(data_summary))
+stopifnot(names(sequencing) == rownames(data_summary))
+stopifnot(names(reference) == rownames(data_summary))
+
+dataset_ref <- paste0(dataset, ' (', reference, ')')
+
+data_table <- cbind(dataset_ref, sequencing, data_summary$n_samples,
+ 										format(data_summary$n_total_seqs, big.mark=','),
+ 										format(data_summary$n_unique_seqs, big.mark=','))
+
+colnames(data_table) <- c("Dataset (Ref.)",
+													"\\bigcell{c}{Read \\\\ Length }",
+													"\\bigcell{c}{Samples \\\\ (N) }",
+													"\\bigcell{c}{Total \\\\ Seqs. (N)}",
+													"\\bigcell{c}{Unique \\\\ Seqs. (N) }")
+
+colnames(data_table) <- paste0("\\textbf{", colnames(data_table), "}")
+
+composite <- xtable(data_table, label="")
+
+align(composite) <- "llcccc"
+
+print(composite,
+		include.rownames=FALSE,
+		include.colnames=TRUE,
+		hline.after = c(-1,0, nrow(composite)),
+		comment=FALSE,
+		table.placement="!ht"
+		)
+```
+
+\newpage
+
 ## Figures

### 2016-12-22 submission/Westcott_OptiClust_mSystems_2016.Rmd
Fix Figure 1 legend to indicate what CI represent
Insert citations

@@
 
-Amplicon-based sequencing has provided incredible insights into Earth's microbial biodiversity.
-It has become common for studies to include sequencing millions of 16S rRNA gene sequences across hundreds of samples [ref].
-This is three to four orders of magnitude greater sequencing depth than was previously achieved using Sanger sequencing [ref].
-The increased sequencing depth has revealed novel taxonomic diversity that is not adequately represented in reference databases [ref].
-However, the advance has forced re-engineering of methods to overcome the rate and memory limiting steps in computational pipelines that process raw sequences through the generation of tables containing the number of sequences in different taxa for each sample [ref].
-A critical component to these pipelines has been the assignment of amplicon sequences to taxonomic units that are ether defined based on similarity to a reference or operationally based on the similarity of the sequences to each other within the dataset.
+Amplicon-based sequencing has provided incredible insights into Earth's microbial biodiversity [@Schloss2016b; @Locey2016].
+It has become common for studies to include sequencing millions of 16S rRNA gene sequences across hundreds of samples [@Rideout2014; @Huttenhower2012].
+This is three to four orders of magnitude greater sequencing depth than was previously achieved using Sanger sequencing [@Eckburg2005; @Elshahed2008].
+The increased sequencing depth has revealed novel taxonomic diversity that is not adequately represented in reference databases [@Schloss2016b; @Rideout2014].
+However, the advance has forced re-engineering of methods to overcome the rate and memory limiting steps in computational pipelines that process raw sequences through the generation of tables containing the number of sequences in different taxa for each sample [e.g.
+@Kozich2013; @Schloss2009a; @Edgar2011; @Wang2007].
+A critical component to these pipelines has been the assignment of amplicon sequences to taxonomic units that are ether defined based on similarity to a reference or operationally based on the similarity of the sequences to each other within the dataset [@Schloss2009b; @Caporaso2010].
 
@@
 These algorithms can be classified into three general categories.
-The first category of algorithms has been termed closed-reference or phylotyping [ref].
+The first category of algorithms has been termed closed-reference or phylotyping [@Schloss2011; @NavasMolina2013].
 Sequences are compared to a reference collection and clustered based on the reference sequences that they are similar to.
 This approach is fast; however, the method struggles when a sequence is similar to multiple reference sequences that may have different taxonomies and when it is not similar to sequences in the reference.
-The second category of algorithms has been called *de novo* because they assign sequences to OTUs without the use of a reference [ref].
-These include hierarchical algorithms such as nearest, furthest, and average neighbor and algorithms that employ heuristics such as abundance or distance-based greedy clustering as implemented in USEARCH or VSEARCH, Sumaclust, OTUCLUST, and Swarm [ref].
+The second category of algorithms has been called *de novo* because they assign sequences to OTUs without the use of a reference [@NavasMolina2013].
+These include hierarchical algorithms such as nearest, furthest, and average neighbor [@Schloss2005] and algorithms that employ heuristics such as abundance or distance-based greedy clustering as implemented in USEARCH [@Edgar2010] or VSEARCH [@Rognes2016], Sumaclust, OTUCLUST [@Albanese2015], and Swarm [@Mah2014].
 *De novo* methods tend to be more computationally intense and it has proven difficult to know which method generates the best assignments.
-A third category of algorithm is open-reference clustering, which is a hybrid approach [ref].
+A third category of algorithm is open-reference clustering, which is a hybrid approach [@NavasMolina2013; @Rideout2014].
 Here sequences are assigned to OTUs using closed-reference clustering and sequences that are not within a threshold of a reference sequence are then clustered using a *de novo* approach.
@@
 Several metrics have emerged for assessing the quality of OTU assignment algorithms.
-These have included the time and memory required to run the algorithm [ref], agreement between OTU assignments and the sequences' taxonomy [ref], sensitivity of an algorithm to stochastic processes [ref], and the number of OTUs generated by the algorithm [ref].
+These have included the time and memory required to run the algorithm [@Sun2009; @Cai2011; @Rideout2014; @Mah2014], agreement between OTU assignments and the sequences' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016], sensitivity of an algorithm to stochastic processes [@He2015], the number of OTUs generated by the algorithm [@Edgar2013; @Kopylova2016], and the ability to regenerate the assignments made by other algorithms [@Schmidt2014; @Rideout2014].
 Unfortunately, these methods fail to directly quantify the quality of the OTU assignments.
@@
 For instance, a method may appear to be recapitulate the taxonomy of a synthetic community with known taxonomic structure, but do a poor job when applied to real communities with poorly defined taxonomic structure or for sequences that are prone to misclassification.
-As an alternative, we developed an approach to objectively benchmark the clustering quality of OTU assignments [ref].
+As an alternative, we developed an approach to objectively benchmark the clustering quality of OTU assignments [@Schloss2016a; @Westcott2015; @Schloss2011].
 This approach counts the number of true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN) based on the pairwise distances.
@@
 Those sequence pairs that have a distance larger than the threshold and are not clustered in the same OTU are TNs and those in the same OTU are FPs.
-These values can be synthesized into a single correlation coefficient, the Matthews' Correlation Coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix [ref].
+These values can be synthesized into a single correlation coefficient, the Matthews' Correlation Coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix [@Matthews1975; @Powers2011].
 Consistently, the average neighbor algorithm was identified as among the best or the best algorithm.
@@
 
-We previously described a heuristic to accelerate OTU assignments where sequences were first classified to taxonomic groups and within each taxon sequences were assigned to OTUs using the average neighbor clustering algorithm [ref].
+We previously described a heuristic to accelerate OTU assignments where sequences were first classified to taxonomic groups and within each taxon sequences were assigned to OTUs using the average neighbor clustering algorithm [@Schloss2011].
 This accelerated the clustering and reduce the memory requirements because the number of unique sequences is effectively reduced by splitting sequences across taxonomic groups.
@@
 We employed the MCC because it includes the four values from a confusion matrix.
-Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled [ref].
+Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled [@Schloss2005; @Schloss2011].
 Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1 score, which is based on three values from the confusion matrix.
-Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure [ref].
+Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure [@Powers2011].
 For example, optimizing for sensitivity could lead to a large number of FPs.
@@
 The challenge in using the average neighbor algorithm is that it requires a large amount of RAM and is computationally demanding.
-This led to the development of a splitting approach that divides the clustering across distinct taxonomic groups [ref].
+This led to the development of a splitting approach that divides the clustering across distinct taxonomic groups [@Schloss2011].
 The improved performance provided by the OptiClust algorithm likely makes such splitting unnecessary for most current datasets.
@@
 Both the similarity of sequences and number of unique sequences can be driven by the sequencing error since any errors will increase the number of unique sequences and these sequences will be closely related to the perfect sequence.
-This underscores the importance of reducing the noise in the sequence data [ref].
+This underscores the importance of reducing the noise in the sequence data [@Kozich2013].
 If sequencing errors are not remediated and are relatively randomly distributed, then it is likely that the algorithm will require an unnecessary amount of time and RAM to complete.
@@
 The method for generating the simulated datasets followed the approach used by Kopylova et al.
-[-@Kopylova2016] and Schloss [@Schloss2016].
+[-@Kopylova2016] and Schloss [-@Schloss2016a].
 Briefly, we randomly selected 10,000 uniques V4 fragments from 16S rRNA gene sequences that were unique from the SILVA non-redundant database [@Pruesse2007].
@@
 In addition to the OptiClust algorithm we evaluated ten different *de novo* clustering algorithms.
-These included three hierarchical algorithms, average neighbor, nearest neighbor, and furthest neighbor, which are implemented in mothur [v.1.39.0; @Schloss2009].
-Seven heuristic methods were also used including abundance-based greedy clustering (AGC) and (distance-based greedy clustering) DGC as implemented in USEARCH [v.6.1; @Edgar2010] and VSEARCH [v.X.X.X; @Rognes2015], OTUClust [v.X.X.X; @XXXXXX], SumaClust [v.X.X.X; @XXXXXX], and Swarm [v.2.1.1; @Mah2014].
+These included three hierarchical algorithms, average neighbor, nearest neighbor, and furthest neighbor, which are implemented in mothur (v.1.39.0) [@Schloss2009b].
+Seven heuristic methods were also used including abundance-based greedy clustering (AGC) and (distance-based greedy clustering) DGC as implemented in USEARCH (v.6.1) [@Edgar2010] and VSEARCH (v.2.3.3) (@Rognes2016], OTUCLUST (v.0.1) [@Albanese2015], SumaClust (v.1.0.20), and Swarm (v.2.1.9) [@Mah2014].
 With the exception of Swarm each of these methods uses distance-based thresholds to report OTU assignments.
@@
 ***Benchmarking.*** We evaluated the quality of the sequence clustering, reproducibility of the clustering, the speed of clustering, and the amount of memory required to complete the clustering.
-To assess the quality of the clusters generated by each method, we counted the cells within a confusion matrix that indicated how well the clusterings represented the distances between the pair of sequences [@Schloss2011Assessing].
+To assess the quality of the clusters generated by each method, we counted the cells within a confusion matrix that indicated how well the clusterings represented the distances between the pair of sequences [@Schloss2011].
 Pairs of sequences that were in the same OTU and had a distance less than 3% were true positives (TPs), those that were in different OTUs and had a distance greater than 3% were true negatives (TNs), those that were in the same OTU and had a distance greater than 3% were false positives (FPs), and those that were in different OTUs and had a distance less than 3% were false negatives (FNs).
@@
 
-***Data and code availability.*** The workflow utilized commands in GNU make (v.3.81), GNU bash (v.4.1.2), mothur [v.1.39.0; @Schloss2009], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015].
-Within R we utilized the wesanderson [v.X.X.X; @XXXXXX], dplyr [v.X.X.X; @XXXXXX], tidyr [v.X.X.X; @XXXXXX], cowplot [v.X.X.X; @XXXXXX], and ggplot2 [v.X.X.X; @XXXXXX] packages.
-A reproducible version of this manuscript and analysis is available at https://github.com/SchlossLab/Westcott_OptiClust_mSystems_2015.
+***Data and code availability.*** The workflow utilized commands in GNU make (v.3.81), GNU bash (v.4.1.2), mothur (v.1.39.0) [@Schloss2009b], and R (v.`r paste(version$major, version$minor, sep='.')`) [@language2015].
+Within R we utilized the wesanderson (v.`r packageDescription("wesanderson")$Version`) [@wesanderson], dplyr (v.`r packageDescription("dplyr")$Version`) [@dplyr], tidyr (v.`r packageDescription("tidyr")$Version`) [@tidyr], cowplot (v.`r packageDescription("cowplot")$Version`) [@cowplot], and ggplot2 (v.`r packageDescription("ggplot2")$Version`) [@ggplot2] packages.
+A reproducible version of this manuscript and analysis is available at https://github.com/SchlossLab/Westcott_OptiClust_mSphere_2017.
 
@@
 reference <-  c(
-							soil = "XX",
-							marine = "XX",
-							mice = "XX",
-							human = "XX",
-							even = "XX,YY",
-							staggered = "XX,YY"
+							soil = "[@Johnston2016]",
+							marine = "[@Henson2016]",
+							mice = "[@Schloss2012]",
+							human = "[@Baxter2016]",
+							even = "[@Kopylova2016; @Schloss2016a]",
+							staggered = "[@Kopylova2016; @Schloss2016a]"
 						)
@@
 The median of 10 re-orderings of the data is presented for each method and dataset.
-The confidence interval is indicated by the error bars, which are typically smaller than the plotting symbol.
+The range of observed values is indicated by the error bars, which are typically smaller than the plotting symbol.
 

### 2016-12-22 submission/Westcott_OptiClust_mSphere_2016.Rmd
Fix formatting

--- 2016-12-22 submission/Westcott_OptiClust_mSystems_2016.Rmd
@@
       in_header: header.tex
-csl: msystems.csl
+csl: msphere.csl
 fontsize: 11pt

### 2016-12-22 submission/Westcott_OptiClust_mSphere_2016.Rmd
Add newer references

@@
 
+## Importance
+
 \newpage
@@
 ***OptiClust algorithm.***
-
 The OptiClust algorithm uses the pairs of sequences that are within a desired threshold of each other (e.g.
@@
 The algorithm continues until the optimization metric stabilizes or until it reaches a defined stopping criteria.
-
-***OptiClust-generated OTUs are more robust than those from other methods.***
 
@@
 
+***OptiClust-generated OTUs are more robust than those from other methods.***
 To evaluate the OptiClust algorithm and compare its performance to other algorithms, we utilized six datasets including two synthetic communities and four previously published large datasets generated from soil, marine, human, and murine samples (Table 1).
@@
 These data demonstrate that OptiClust generated significantly more robust OTU assignments than existing methods across a diverse collection of datasets with performance that was comparable to popular methods.
-
-***OptiClust stopping criteria.***
 
@@
 ```
-
+***OptiClust stopping criteria.***
 By default, the mothur-based implementation of the algorithm stops when the optimization metric changes by less than 0.0001; however, this can be altered by the user.
@@
 Although the mothur's default setting is partial convergence, the remainder of our analysis used complete convergence to be more conservative.
-
-***Effect of seeding OTUs on OptiClust performance.***
 
@@
 
+***Effect of seeding OTUs on OptiClust performance.***
 By default the mothur implementation of the OptiClust algorithm starts with each sequence in a separate OTU.
@@
 This analysis demonstrates that seeding the algorithm with sequences as separate OTUs resulted in the best OTU assignments in the shortest amount of time.
-
-***OptiClust-generated OTUs are as stable as those from other algorithms.***
 
@@
 
+***OptiClust-generated OTUs are as stable as those from other algorithms.***
 One concern that many have with *de novo* clustering algorithms is that their output is sensitive to the initial order of the sequences.
@@
 More importantly, the random variation does yield output of a similarly high quality.
-
-***Time and memory required to complete Optimization-based clustering scales efficiently.***
 
@@
 
+***Time and memory required to complete Optimization-based clustering scales efficiently.***
 Although not as important as the quality of clustering, the amount of time and memory required to assign sequences to OTUs is a legitimate concern.
@@
 
-***Cluster splitting heuristic generates OTUs that are as good as non-split approach (Figure 6).***
-
+***Cluster splitting heuristic generates OTUs that are as good as non-split approach.***
 We previously described a heuristic to accelerate OTU assignments where sequences were first classified to taxonomic groups and within each taxon sequences were assigned to OTUs using the average neighbor clustering algorithm [@Schloss2011].

### 2016-12-22 submission/Westcott_OptiClust_mSphere_2016.Rmd
Add Abstract and Importance sections

@@
 ## Abstract
+Assignment of 16S rRNA gene sequences to operational taxonomic units (OTUs) is a computational bottleneck in the process of analyzing microbial communities.
+Although this has been an active area of research, it has been difficult to overcome the time and memory demands while improving the quality of the OTU assignments.
+Here we developed a new OTU assignment algorithm that iteratively reassigns sequences to new OTUs to optimize the Matthews correlation coefficient (MCC), a measure of the quality of OTU assignments.
+To assess the new algorithm, OptiClust, we compared it to ten other algorithms using 16S rRNA gene sequences from two simulated and four natural communities.
+Using the OptiClust algorithm, the MCC values averaged 15.2 and 16.5% higher than the OTUs generated when we used the average neighbor and distance-based greedy clustering with VSEARCH, respectively.
+Furthermore, on average, OptiClust was 94.6-times faster than the average neighbor algorithm and just as fast as distance-based greedy clustering with VSEARCH.
+An empirical analysis of the efficiency of the algorithms showed that the time and memory required to perform the algorithm scaled quadratically with the number of unique sequences in the dataset.
+The significant improvement in the quality of the OTU assignments over previously existing methods will significantly enhance downstream analysis by limiting the splitting of similar sequences into separate OTUs and merging of dissimilar sequences into the same OTU.
+The development of the OptiClust algorithm represents a significant advance that is likely to have numerous other applications.
 
 ## Importance
+The analysis of microbial communities from diverse environments using 16S rRNA gene sequencing has expanded our knowledge of the biogeography of microorganisms.
+An important step in this analysis is the assignment of sequences into taxonomic groups based on their similarity to sequences in a database or based on their similarity to each other, irrespective of a database.
+In this study, we present a new algorithm for the latter approach.
+The algorithm, OptiClust, seeks to optimize a metric of assignment quality by shuffling sequences between taxonomic groups.
+We found that OptiClust produces more robust assignments and does so in a rapid and memory efficient manner.
+This advance will allow for a more robust analysis of microbial communities and the factors that shape them.
 

### 2016-12-23 code/build_scaling_figure.R
Fix y-axis labels to start at zero and be evenly spaced

@@ -6,6 +6,7 @@ library(tidyr, quietly=T, warn.conflicts=F)
 library(ggplot2, quietly=T, warn.conflicts=F)
 library(wesanderson, quietly=T, warn.conflicts=F)
 library(cowplot, quietly=T, warn.conflicts=F)
+library(scales)
 
 basic_methods <- c('mcc')
 pretty_methods <- c('OptiClust')
@@ -36,11 +37,6 @@ speed_models <- speed_data %>%
 		group_by(dataset) %>%
 		do(model = nls((full_secs_cluster/full_steps) ~ b * frac ^ z, start = list(b = 1000, z = 2), data= .))
 
-# speed_models[[1,2]]$m$getPars()["z"]	#2.131886
-# speed_models[[2,2]]$m$getPars()["z"]	#2.975044
-# speed_models[[3,2]]$m$getPars()["z"]	#2.687899
-# speed_models[[4,2]]$m$getPars()["z"]	#2.379366
-
 
 
 
@@ -62,10 +58,9 @@ memory_models <- memory_data %>%
 		group_by(dataset) %>%
 		do(model = nls((cluster_gb) ~ b * frac ^ z, start = list(b = 1e6, z = 3), data= .))
 
-# memory_models[[1,2]]$m$getPars()["z"]	#1.998879
-# memory_models[[2,2]]$m$getPars()["z"]	#1.97656
-# memory_models[[3,2]]$m$getPars()["z"]	#1.995457
-# memory_models[[4,2]]$m$getPars()["z"]	#1.957556
+
+#######################################################
+
 
 my_theme <- theme_classic() +
 	theme(
@@ -90,12 +85,19 @@ my_legend <- theme(
 		legend.margin = margin(t=0,4,4,4)
 	)
 
+mysqrt_trans <- function() {
+  trans_new("mysqrt",
+            transform = base::sqrt,
+            inverse = function(x) ifelse(x<0, 0, x^2),
+            domain = c(0, Inf))
+}
+
 speed_plot <- speed_summary %>%
 		ggplot() +
 		geom_line(aes(frac, avg, color=dataset)) +
 		geom_point(aes(frac, avg, color=dataset)) +
 		geom_errorbar(aes(frac, ymin=min, ymax=max, col=dataset), width=0.02) +
-		scale_y_sqrt(breaks=c(0,2,4,8,16,32,64,128,256)) +
[…]

### 2016-12-23 code/build_performance_figure.R
Fix image width

@@ -126,4 +126,4 @@ ggdraw() +
 	draw_plot(sobs + theme(axis.title.y=element_text(margin=margin(r=11,l=0,t=0,b=0))) + my_legend,x=0,y=0.4,1,0.3) +
 	draw_plot(secs,x=0,y=0.0,1,0.4) +
 	draw_plot_label(c("A", "B", "C"), x=c(0,0,0), y=c(1.00,0.71,0.41), size=12) +
-	ggsave('results/figures/performance.tiff', width=6, height=5.5, unit='in')
+	ggsave('results/figures/performance.tiff', width=6.875, height=5.5, unit='in')

### 2016-12-23 submission/Westcott_OptiClust_mSphere_2016.Rmd
Add running title

@@
 ```
+
+**Running title:** OptiClust: Optimized Clustering
 
@@
 Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1 score, which is based on three values from the confusion matrix.
-Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure [@Powers2011].
+Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure.
 For example, optimizing for sensitivity could lead to a large number of FPs.
@@
 A reproducible version of this manuscript and analysis is available at https://github.com/SchlossLab/Westcott_OptiClust_mSphere_2017.
+
+## Acknowledgements
+This work was supported through funding from the National Institutes of Health to PDS ().
+SLW designed, implemented, and evaluated the algorithm.
+PDS designed and evaluated the algorithm.
+Both authors wrote and edited the manuscript.
 

### 2016-12-23 submission/Westcott_OptiClust_mSphere_2016.Rmd
Embed images into manuscript file for initial submission

@@
       in_header: header.tex
+    fig_caption: false
 csl: msphere.csl
@@
 
+\begin{center}
+
 **Running title:** OptiClust: Optimized Clustering
-
-\begin{center}
 \vspace{25mm}
@@
 ## Acknowledgements
-This work was supported through funding from the National Institutes of Health to PDS ().
+This work was supported through funding from the National Institutes of Health to PDS (P30DK034933).
 SLW designed, implemented, and evaluated the algorithm.
@@
 ```{r results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
-library('xtable')
-
-options(xtable.sanitize.text.function=identity)
 
@@
 
-dataset_ref <- paste0(dataset, ' (', reference, ')')
+dataset_ref <- paste0(dataset, ' ', reference)
 
@@
 
-colnames(data_table) <- c("Dataset (Ref.)",
-													"\\bigcell{c}{Read \\\\ Length }",
-													"\\bigcell{c}{Samples \\\\ (N) }",
-													"\\bigcell{c}{Total \\\\ Seqs. (N)}",
-													"\\bigcell{c}{Unique \\\\ Seqs. (N) }")
+colnames(data_table) <- c("**Dataset (Ref.)**",
+													"**Read Length**",
+													"**Samples**",
+													"**Total Seqs.**",
+													"**Unique Seqs.**")
 
-colnames(data_table) <- paste0("\\textbf{", colnames(data_table), "}")
+kable(data_table, row.names=F, align="lcccc")
 
-composite <- xtable(data_table, label="")
-
-align(composite) <- "llcccc"
-
-print(composite,
-		include.rownames=FALSE,
-		include.colnames=TRUE,
-		hline.after = c(-1,0, nrow(composite)),
-		comment=FALSE,
-		table.placement="!ht"
-		)
 ```
@@
 
-## Figures
+\includegraphics[width=6.0in]{../results/figures/performance.png}
 
@@
 
+\newpage
+
+\includegraphics[width=3.5in]{../results/figures/speed_memory.png}
+
 **Figure 2.
@@
 
+\newpage
+
+\includegraphics[width=3.5in]{../results/figures/split_mcc.png}
+
 **Figure 3.
 Effects of taxonomically splitting the datasets on clustering quality.** The datasets were split at each taxonomic level based on their classification using a naive Bayesian classifier and clustered using average neighbor, VSEARCH-based DGC, and OptiClust.
+
+\newpage
 

### 2016-12-23 code/build_performance_figure.R

@@ -126,4 +126,5 @@ ggdraw() +
 	draw_plot(sobs + theme(axis.title.y=element_text(margin=margin(r=11,l=0,t=0,b=0))) + my_legend,x=0,y=0.4,1,0.3) +
 	draw_plot(secs,x=0,y=0.0,1,0.4) +
 	draw_plot_label(c("A", "B", "C"), x=c(0,0,0), y=c(1.00,0.71,0.41), size=12) +
-	ggsave('results/figures/performance.tiff', width=6.875, height=5.5, unit='in')
+	ggsave('results/figures/performance.tiff', width=6.875, height=5.5, unit='in') +
+	ggsave('results/figures/performance.png', width=6.875, height=5.5, unit='in')

### 2016-12-23 code/build_scaling_figure.R

@@ -143,4 +143,5 @@ ggdraw() +
 	draw_plot(speed_plot,  x=0, y=0.535, 1.0, 0.465) +
 	draw_plot(memory_plot, x=0, y=0.000, 1.0, 0.535) +
 	draw_plot_label(c("A", "B"), x=c(0,0), y=c(1.00,0.54), size=12) +
-	ggsave('results/figures/speed_memory.tiff', width=3.5, height=4.0, unit='in')
+	ggsave('results/figures/speed_memory.tiff', width=3.5, height=4.0, unit='in') +
+	ggsave('results/figures/speed_memory.png', width=3.5, height=4.0, unit='in')

### 2016-12-23 code/build_split_figure.R

@@ -104,4 +104,5 @@ figure <- split_data %>% filter(metric == "mcc") %>%
 ggdraw() +
 	draw_plot(figure,  x=0, y=0, 1.0, 1) +
 	draw_plot_label(pretty_datasets, x=rep(0.15,4), y=c(0.825,0.6,0.375,0.15), size=10, hjust=0) +
-	ggsave("results/figures/split_mcc.tiff", width=3.5, height=5.0, unit="in")
+	ggsave("results/figures/split_mcc.tiff", width=3.5, height=5.0, unit="in") +
+	ggsave("results/figures/split_mcc.png", width=3.5, height=5.0, unit="in")

### 2016-12-23 submission/Westcott_OptiClust_mSphere_2016.Rmd
Fix missing reference

@@
 Those sequence pairs that have a distance larger than the threshold and are not clustered in the same OTU are TNs and those in the same OTU are FPs.
-These values can be synthesized into a single correlation coefficient, the Matthews' Correlation Coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix [@Matthews1975; @Powers2011].
+These values can be synthesized into a single correlation coefficient, the Matthews' Correlation Coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix [@Matthews1975].
 Consistently, the average neighbor algorithm was identified as among the best or the best algorithm.

### 2017-01-31 code/summarize_datasets.R
Add number of otus and distances

@@ -1,8 +1,20 @@
 dataset <- c('soil', 'marine', 'human', 'mice', 'staggered', 'even')
 
-count_file_names <- paste0('data/', dataset, '/', dataset, '.count_table')
+distances <- read.table(file="data/processed/distance_counts.tsv")
+distances$V2 <- gsub("data/(.*)/(.*).1_0.01.sm.dist", "\\1", distances$V2)
+rownames(distances) <- distances$V2
 
-get_stats <- function(count_file_name) {
+sobs <- read.table(file="data/processed/sobs_counts.tsv")
+sobs <- sobs[grep("1_0.*.mcc.list", x=sobs$V1),]
+sobs$V1 <- gsub("data/(.*)/.*\\.1_0.*", "\\1", sobs$V1)
+sobs_mean <- sobs %>%
+							select(c(1,3)) %>%
+							group_by(V1) %>%
+							summarize(ave=mean(V3))
+
+get_stats <- function(dataset) {
+
+	count_file_name <- paste0('data/', dataset, '/', dataset, '.count_table')
 	count_data <- read.table(file=count_file_name, header=T)
 
 	n_unique_seqs <- nrow(count_data)
@@ -11,11 +23,14 @@ get_stats <- function(count_file_name) {
 	n_samples <- ncol(count_data) - 2
 	if(n_samples == 0) n_samples <- NA
 
-	c(n_samples=n_samples, n_unique_seqs=n_unique_seqs, n_total_seqs=n_total_seqs)
+	n_distances <- distances[dataset, "V1"]
+	n_sobs <- sobs_mean %>% filter(V1==dataset) %>% select(ave) %>% .[[1]]
+	
+	c(n_samples=n_samples, n_unique_seqs=n_unique_seqs, n_total_seqs=n_total_seqs,
+		n_distances=n_distances, n_sobs=n_sobs)
 }
 
-summary_table <- t(sapply(count_file_names, get_stats))
+summary_table <- t(sapply(dataset, get_stats))
 rownames(summary_table) <- dataset
 
 write.table(summary_table, file="data/processed/datasets.summary", quote=F, sep='\t')
-

### 2017-01-31 code/summarize_datasets.R
Add number of otus and distances

@@ -1,3 +1,5 @@
+library(dplyr)
+
 dataset <- c('soil', 'marine', 'human', 'mice', 'staggered', 'even')
 
 distances <- read.table(file="data/processed/distance_counts.tsv")

### 2017-02-02 code/aggregate_stats.R
Add new methods

@@ -152,11 +152,12 @@ n_fracs <- length(fractions)
 replicates <- 1:10
 n_reps <- length(replicates)
 
-# these are the 22 methods that we tested for clustering
-methods <- c("accuracy", "an", "an_split2_8", "an_split3_8", "an_split4_8", "an_split5_8", "an_split6_8",
-			"f1score", "fn", "mcc", "mcc_split2_8", "mcc_split3_8", "mcc_split4_8", "mcc_split5_8", "mcc_split6_8",
-			"nn", "otuclust", "sumaclust", "swarm", "uagc", "udgc", "vagc_1", "vagc_8", "vdgc_1", "vdgc_8",
-			"vdgc_split2_8", "vdgc_split3_8", "vdgc_split4_8", "vdgc_split5_8", "vdgc_split6_8", "mcc_agg")
+# these are the 37 methods that we tested for clustering
+methods <- c("accuracy", "an", "an_split2_8", "an_split3_8", "an_split4_8", "an_split5_8",
+			"an_split6_8", "f1score", "fn", "mcc", "mcc_split2_8", "mcc_split3_8", "mcc_split4_8",
+			"mcc_split5_8", "mcc_split6_8", "nn", "otuclust", "sumaclust", "swarm", "uagc", "udgc",
+			"vagc_1", "vagc_8", "vdgc_1", "vdgc_8", "vdgc_split2_8", "vdgc_split3_8", "vdgc_split4_8",
+			"vdgc_split5_8", "vdgc_split6_8", "mcc_agg", "fpfn", "npv", "opti_fn", "sens", "tp", "tptn")
 n_methods <- length(methods)
 
 

### 2017-02-02 code/build_optimization_figure.R
Build optimization figure to match performance figure

@@ -1,0 +1,63 @@
+library(dplyr)
+library(ggplot2)
+library(wesanderson)
+
+get_steps_data <- function(dataset, rep, fraction="1_0"){
+
+	steps_file_name <- paste0("data/", dataset, "/", dataset, ".", fraction, ".", rep, ".mcc.steps")
+	read.table(file=steps_file_name, stringsAsFactors=FALSE, header=T) %>%
+						select(iter, mcc) %>%
+						filter(iter != 0) %>%
+						mutate(dataset=dataset, replicate=rep)
+
+}
+
+
+datasets <- c('soil', 'marine', 'mice', 'human', 'even', 'staggered')
+pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
+names(pretty_datasets) <- datasets
+
+libraries <- rep(datasets, each=10)
+reps <- rep(c(paste0('0', 1:9), '10'), length.out=length(libraries))
+
+
+all_step_data <- data.frame()
+
+for(i in 1:length(libraries)){
+	new_step_data <- get_steps_data(dataset=libraries[i], rep=reps[i])
+	all_step_data <- rbind(all_step_data, new_step_data)
+}
+
+#write.table(all_step_data, file="test.tsv", row.names=FALSE)
+#all_step_data <- read.table(file="test.tsv", header=T, stringsAsFactors=F)
+#all_step_data$dataset <- factor(all_step_data$dataset, levels=datasets)
+
+my_theme <- theme_classic() +
+	theme(
+		panel.grid.major = element_blank(),
+		panel.grid.minor = element_blank(),
+		panel.border = element_rect(color = "black", fill=NA, size=1),
+		panel.background = element_rect(fill=NA),
+		axis.text.y=element_text(size=5),
+		axis.title.y=element_text(size=7),
+		axis.text.x=element_text(size=5),
+		axis.title.x=element_text(size=7),
+		legend.text = element_text(size=5),
+		legend.margin = margin(t=0,0,0,0),
+		legend.key.height=unit(0.5,"line"),
+		legend.key.width=unit(1,"line")
+	)
+
+
+
+all_step_data %>%
+	ggplot(aes(x=iter, y=mcc, color=dataset, group=interaction(dataset, replicate))) +
+	geom_line() +
+	scale_color_manual(breaks=levels(all_step_data$dataset), labels=pretty_datasets[levels(all_step_data$dataset)],
[…]

### 2017-02-02 code/build_performance_figure.R

@@ -22,7 +22,7 @@ pretty_metrics <- c("Matthew's\nCorrelation Coefficient", "Number of OTUs", "Tim
 names(pretty_metrics) <- metrics
 
 
-full_data <- read.table(file="data/processed/cluster_data.summary", header=T)
+full_data <- read.table(file="data/processed/cluster_data.summary", header=T, stringsAsFactors=F)
 
 full_tidy <- full_data %>%
 						filter(method %in% basic_methods, frac==1.0) %>%
@@ -33,7 +33,7 @@ full_tidy <- full_data %>%
 											min=min(value, na.rm=T),
 											max=max(value, na.rm=T))
 
-close_data <- read.table(file="data/processed/mcc_steps.summary", header=T)
+close_data <- read.table(file="data/processed/mcc_steps.summary", header=T, stringsAsFactors=F)
 close_data <- close_data %>%  select(-starts_with("full_"), -close_steps, -close_secs_cluster)
 close_data$method <- 'mcc_close'
 colnames(close_data) <- c('dataset', 'frac', 'rep', 'method', 'mcc', 'num_otus', 'cluster_secs')
@@ -70,8 +70,11 @@ my_theme <- theme_classic() +
 		legend.position = "none"
 	)
 
+
+alpha <- 0.3
 mcc <- tidy %>% filter(metric == "mcc") %>%
 	ggplot(aes(method, avg, col=dataset, shape=dataset))+
+	geom_hline(yintercept = seq(0,1,0.25), colour = "gray", size=0.15) +
 	geom_point(position = position_dodge(0.7), size=2) +
 	geom_errorbar(position = position_dodge(0.7), aes(ymin=min, ymax=max), width=0.2) +
 	expand_limits(y=c(0,1)) +
@@ -84,6 +87,8 @@ mcc <- tidy %>% filter(metric == "mcc") %>%
 
 sobs <- tidy %>% filter(metric == "num_otus") %>%
 	ggplot(aes(method, avg, col=dataset, shape=dataset))+
+	geom_hline(yintercept = seq(0,1e5,2.5e4), colour = "gray", size=0.15) +
+
 	geom_point(position = position_dodge(0.7), size=2) +
 	geom_errorbar(position = position_dodge(0.7), aes(ymin=min, ymax=max), width=0.2) +
 	expand_limits(y=c(0,1)) +
[…]

### 2017-02-02 code/build_optimization_figure.R
Fix legend

@@ -30,7 +30,7 @@ for(i in 1:length(libraries)){
 
 #write.table(all_step_data, file="test.tsv", row.names=FALSE)
 #all_step_data <- read.table(file="test.tsv", header=T, stringsAsFactors=F)
-#all_step_data$dataset <- factor(all_step_data$dataset, levels=datasets)
+all_step_data$dataset <- factor(all_step_data$dataset, levels=datasets)
 
 my_theme <- theme_classic() +
 	theme(

### 2017-02-02 code/build_optimization_figure.R
Change proportions

@@ -59,5 +59,5 @@ all_step_data %>%
 	expand_limits(y=c(0.7,0.9), x=c(0,17)) +
 	scale_x_continuous(breaks=c(0,5,10,15,20), labels=c(0,5,10,15,20)) +
 	my_theme +
-	ggsave("results/figures/optimization.tiff", width=3.5, height=3.5, unit='in') +
-	ggsave("results/figures/optimization.png", width=3.5, height=3.5, unit='in')
+	ggsave("results/figures/optimization.tiff", width=5, height=3.5, unit='in') +
+	ggsave("results/figures/optimization.png", width=5, height=3.5, unit='in')

### 2017-02-02 code/build_optimization_figure.R
Change proportions

@@ -59,5 +59,5 @@ all_step_data %>%
 	expand_limits(y=c(0.7,0.9), x=c(0,17)) +
 	scale_x_continuous(breaks=c(0,5,10,15,20), labels=c(0,5,10,15,20)) +
 	my_theme +
-	ggsave("results/figures/optimization.tiff", width=5, height=3.5, unit='in') +
-	ggsave("results/figures/optimization.png", width=5, height=3.5, unit='in')
+	ggsave("results/figures/optimization.tiff", width=3.5, height=2.5, unit='in') +
+	ggsave("results/figures/optimization.png", width=3.5, height=2.5, unit='in')

### 2017-02-02 code/build_performance_other_figure.R
Add optimization performance for other metrics

@@ -1,0 +1,125 @@
+#devtools::install_github("hadley/ggplot2")
+#devtools::install_github("hadley/scales")
+#devtools::install_github("wilkelab/cowplot")
+library(dplyr, warn.conflicts=F)
+library(tidyr, warn.conflicts=F)
+library(ggplot2, warn.conflicts=F)
+library(cowplot, warn.conflicts=F)
+library(wesanderson, warn.conflicts=F)
+
+basic_methods <- c('accuracy', 'f1score', 'fpfn', 'mcc', 'npv', 'opti_fn', 'sens', 'tp', 'tptn')
+
+pretty_methods <- c('Accuracy\n(max.)', 'F1-score\n(max.)', 'False Positives +\nFalse Negatives\n(min.)',
+										"Matthew's Correlation\nCoefficient (max.)",
+										"Negative\nPredictive\nValue (max.)", "False Negatives\n(min.)", "Sensitivity\n(max.)",
+ 										"True Positives\n(max.)", "True Positives +\nTrue Negatives (max.)")
+
+names(pretty_methods) <- basic_methods
+
+datasets <- c('soil', 'marine', 'mice', 'human', 'even', 'staggered')
+pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
+names(pretty_datasets) <- datasets
+
+metrics <- c("mcc", "num_otus", "cluster_secs")
+pretty_metrics <- c("Matthew's\nCorrelation Coefficient", "Number of OTUs", "Time to Cluster\n(seconds)")
+names(pretty_metrics) <- metrics
+
+
+full_data <- read.table(file="data/processed/cluster_data.summary", header=T, stringsAsFactors=F)
+
+full_tidy <- full_data %>%
+						filter(method %in% basic_methods, frac==1.0) %>%
+						select(dataset, method, mcc, num_otus, cluster_secs) %>%
+						gather(metric, value, mcc, num_otus, cluster_secs) %>%
+						group_by(dataset, method, metric) %>%
+						summarize(avg=mean(value, na.rm=T),
+											min=min(value, na.rm=T),
+											max=max(value, na.rm=T))
+
+method_ordering <- full_tidy %>%
+			filter(metric=='mcc') %>%
+			group_by(method) %>%
+			summarize(avg_mcc=mean(avg, na.rm=T)) %>%
+			arrange(desc(avg_mcc))
+
[…]

### 2017-02-02 code/build_performance_other_figure.R
Adjust layout of x-axis labels

@@ -9,10 +9,11 @@ library(wesanderson, warn.conflicts=F)
 
 basic_methods <- c('accuracy', 'f1score', 'fpfn', 'mcc', 'npv', 'opti_fn', 'sens', 'tp', 'tptn')
 
-pretty_methods <- c('Accuracy\n(max.)', 'F1-score\n(max.)', 'False Positives +\nFalse Negatives\n(min.)',
+pretty_methods <- c('Accuracy (max.)', 'F1-score (max.)',
+										'False Positives +\nFalse Negatives (min.)',
 										"Matthew's Correlation\nCoefficient (max.)",
-										"Negative\nPredictive\nValue (max.)", "False Negatives\n(min.)", "Sensitivity\n(max.)",
- 										"True Positives\n(max.)", "True Positives +\nTrue Negatives (max.)")
+										"Negative Predictive\nValue (max.)", "False Negatives (min.)", "Sensitivity (max.)",
+ 										"True Positives (max.)", "True Positives +\nTrue Negatives (max.)")
 
 names(pretty_methods) <- basic_methods
 

### 2017-02-02 submission/response_to_reviewers.md
First attempt to respond to initial round of revisions

---
title: ""
output:
  pdf_document:
    includes:
      in_header: header.tex
fontsize: 11pt
geometry: margin=1.0in
---

## Reviewer 1

### Major comments
**Optimizing a criterion is a good idea. It would be useful to clarify if each of the other algorithms aims to optimize some other criterion and if so which one. There is some text about this in the discussion, but this information might work better earlier in the manuscript. It would also be helpful to see how each algorithm performs on these other metrics (e.g., minimizing FPs). You can argue that MCC is a better criterion to use, but it is also good to show if the methods that aim to optimize another criterion actually do so and how MCC-based OptiClust does with regards to these metrics. In addition, some further discussion about the relative costs of FPs and FNs in typical metagenomics applications and downstream studies would provide some context for your argument to use MCC.**

We have added text to the Introduction to describe that the furthest and nearest neighbor algorithms seek to minimize the number of FPs and FNs, respectively (***LXXX***). We have also added text to the Discussion (***LXXX***) pointing out that More FPs would increase the number of OTUs while more FNs would collapse OTUs together. It is difficult to know which is worse since both alpha and beta diversity metrics are tied to the number and composition of OTUs. As we point out, this is why it is important to jointly minimize the number of FPs and FNs.

**For the simulations, it would be nice to show the true values of some of the output measures (e.g., number of clusters) versus the estimated values.**

Unfortunately, this is a major problem with evaluating clustering algorithms. Except under highly contrived conditions, we do not know what the true values are for output measures such as the number of clusters. This problem is outlined in ***LXXX-LXXX***. This is why we have developed the concept of the MCC over our last several papers on the subject of OTU clustering.

**A systematic evaluation of the influence of additional variables on performance (i.e., accuracy metrics, RAM, and speed) would provide better insight into expected performance on other data sets. The analyzed data sets do cover a broad range of environments with different types of communities. Even with those data, it would be nice to see performance metrics plotted versus phylogenetic diversity, percentage of sequences in the reference database (for pre-assignment analyses), and sequencing variables (e.g., read length, quality scores). In the simulations, you could additionally introduce realistic sequencing errors evaluate how performance changes.**

As the reviewer correctly points out, there is no ideal dataset. This is why we selected four datasets that represent a variety of environments that are currently popular among microbial ecologists. We also added the simulated datasets (i.e. even and staggered) since these have been used in prior studies evaluating clustering methods. Given that the nature of the algorithm is to optimize the MCC coefficient (or any other coefficient) and we have demonstrated that it is effective at doing this, we don't think it is necessary to add additional datasets to reinfoce the point. To the reviewer's suggestion, we have added the number of OTUs that were found using the OptiClust algorithm and the number of pairwise distances that were equal to and smaller than 0.03 to the summary statistics in Table 1. Unfortunately, some of the datasets are too large and the sequencing data too noisy to permit the construction of the phylogenetic trees that would be needed to calculate the phylogenetic diversity for each study. As for comparing to percentage of sequences in a reference database, we feel that this is would be a misleading point as we have previously shown that reference-based methods are inferior to *de novo* clustering methods (***XXXX***) and the identity of the sequences does not affect the performance of the algorithm (i.e. we could have used synthetic DNA sequences and communities and reached the same conclusion). Finally, the the point regarding read length, quality scores, and sequencing errors we do feel that these analyses are beyond the scope of the current paper. OptiClust only uses pairwise distances as inputs to the clustering and data quality is a very important, but different problem. Given the data provided in Table 1 and Figure 2, we have demonstrated the effect of increasing the number of distances on the performance of the algorithm.

**This may be beyond the scope, but it would be very interesting to see the performance of OptiClust with other criteria as the target of optimization. In other words, how much of the performance is due to targeting MCC as the criterion versus the algorithm itself as an alternative to hierarchical clustering. You could investigate this by comparing the OptiClust algorithm minimizing FPs, for example, as compared to hierarchical clustering minimizing FPs. If this analysis is not performed, I recommend discussing that the algorithm for building clusters and the criterion that is being optimized are evaluated together in your analyses and not disentangled.**

We have added two supplementary figures. The first demonstrates the performance of the algorithm when we optimized clustering based on a variety of other metrics (Figure S1). The second, demonstrates how the MCC values improve when optimizing for the MCC value with each successive iteration (Figure S2).

**Providing the analysis pipeline is great. Users who want to apply the method may like to have more description of the implementation and a stable link to the code (even if this is the same as mothur). To enable applications outside OTU clustering, you should also have a code base that is not embedded in mothur.**

The code for the OptiClust algorithm is available as C++ source code within mothur (e.g. opticluster.cpp at https://github.com/mothur/mothur/tree/master/source). We are reluctant to separate the OptiClust source code from mothur to make it a stand alone tool. This makes maintenance of the software cumbersome and is contrary to the already successful development strategy that we have had for mothur. It should be noted that mothur is wrapped within other tools such as QIIME. Also, mothur can be run from the command line or using scripts. Because of these points, we do not feel it is necessary to separate the code out of the main package.

### Minor comments

**Line 69: Typo: “to be recapitulate”**

This has been fixed.

**Line 132: Not clear why the number of OTUs is mentioned here**

This was meant to indicate that the small difference in MCC values (0.0001) had a relatively small effect on the number of OTUs that result from the algorithm. Although we do not feel that the number of OTUs is a particularly relevant metric of clustering quality, we did feel that it was important to show the effect of small differences in MCC on measures of community richness.

**Line 138: With partial convergence, does the output include the last value for change in MCC?**

Yes, the MCC values are provided as part of the output of running the command.

**Lines 140-150: Mention if agglomerative hierarchical methods are also better than divisive (cite literature or own data).**

We have added a sentence to the Introduction (***LXXX***) indicating that all of the *de novo* methods used by microbial ecologists are agglomerative.

**Lines 151-153: Processing the sequences/OTUs deterministically is a good option for alleviating dependence on data order.**

We have added a clause to this sentence to indicate that this really isn't possible since the algorithms must find a way to break ties when a sequence is equally similar to multiple OTUs.

**Line 161: Typo: “values considerably smaller”**

This has been fixed.

**Line 170: Is disk or CPU also useful to measure?**

In our experience users are limited by RAM and want the methods to complete as quickly as possible. This is shown by the failure of the USEARCH methods to load the datasets in the available RAM or other algorithms to finish in 50 hrs with 48 GB of RAM. The availability of disk space is rarely a problem.

**Lines 182-185: Clarify that this is assuming the pairwise similarities are similar to the analyzed data set. Can you say something about what happens as diversity increases or decreases?**

We have added a line at ***LXXX*** and ***LXXX*** to indicate that speed was proportionate to the number of distances smaller than 0.03. It does not necessarily appear to be related to the number of sequences or the number of OTUs.

**Line 187: Remind the reader that this is “open reference”**

A sentence has been added to this section to clarify the differnces between our method and open reference clustering. Briefly, in open reference clustering sequences are classified and any that do not map to a reference are clustered using a de novo approach. In our approach, all sequences are classified (not mapped) to a broad taxonomic level and then all sequences are clustered using a de novo approach.

**Line 189: Typo: “reduce” should be past tense**

This has been fixed.

**Line 197: Are all of these with taxonomic assignment first? Clarify.**

This has been clarified.

**Line 204: Explain why (i.e., what types of errors the extra clusters represent).**

This has been clarified.

**Line 232: Typo: “effectively ask”**

This has been fixed.

**Line 239: Typo: “are encourage”**

This has been fixed.

## Reviewer 2

### Major comments

**This is a well described, self-enclosed, well implemented piece of work that does not require major revision.  However, there are minor issues which should be addressed.**

### Minor comments

**On page 6 and 7 the authors compare the speed of OptiClust to other methods; however, I do not think Figure 1 really conveys this, mostly due to the log-scaled y-axis for the time panel.  Whilst I recognise why the log-scale has been used, the authors may consider a different way to visually present the data that better conveys the message "The OptiClust algorithm was considerably faster than the hierarchical algorithms and somewhat slower than the heuristic-based algorithms"**

We tried several methods for improving the visualization. First, we made all speed values relative to those in the first column. The problem with this was that the range of values was still considerable and obscured the ability to see the differences. Second, we added grid lines that overlapped with the values from the first column and extended to the right along the plot. This ended up making the figure look overly busy. Third, we attempted to group the methods within the sample types (we had grouped the samples within the methods type), but it was difficult to clearly represent the 13 different methods and 6 samples efficiently. Ultimately, we decided to add a thin gridline at the major y-axis values to give the reader something to anchor their perspective as they looked across the plot to compare the different methods.

**The comparison of RAM used (at the top of page 7) may also be better demonstrated visually.**

At ***LXXX***, our goal wasn't to compare RAM usage between methods. We merely wanted to point out that the USEARCH-based approaches could not handle large datasets without purchasing the 64-bit version of the software and that OTUCLUST and SumClust could not store the data within 45 GB of RAM.

**It wasn't clear to me which software implementation was used for "Nearest Neighbour" and "Furthest Neighbour".**

These were run using the implementations found in mothur (***LXXXX***).

**The authors describe the potential implications of optimising a metric other than MCC - I wonder if they have actually done this?**

We have added text to the manuscript in response to this comment and that of the first reviewer.

### 2017-02-02 submission/Westcott_OptiClust_mSphere_2016.Rmd
Fix cover page formatting
First attempt to respond to initial round of revisions

@@
 
-**Running title:** OptiClust: Optimized Clustering
+Running title: OptiClust: Optimized Clustering
+
 \vspace{25mm}
@@
 These include hierarchical algorithms such as nearest, furthest, and average neighbor [@Schloss2005] and algorithms that employ heuristics such as abundance or distance-based greedy clustering as implemented in USEARCH [@Edgar2010] or VSEARCH [@Rognes2016], Sumaclust, OTUCLUST [@Albanese2015], and Swarm [@Mah2014].
-*De novo* methods tend to be more computationally intense and it has proven difficult to know which method generates the best assignments.
-A third category of algorithm is open-reference clustering, which is a hybrid approach [@NavasMolina2013; @Rideout2014].
+*De novo* methods are agglomerative and tend to be more computationally intense.
+It has proven difficult to know which method generates the best assignments.
+A third category of algorithm is open reference clustering, which is a hybrid approach [@NavasMolina2013; @Rideout2014].
 Here sequences are assigned to OTUs using closed-reference clustering and sequences that are not within a threshold of a reference sequence are then clustered using a *de novo* approach.
@@
 These metrics also tend to be subjective.
-For instance, a method may appear to be recapitulate the taxonomy of a synthetic community with known taxonomic structure, but do a poor job when applied to real communities with poorly defined taxonomic structure or for sequences that are prone to misclassification.
+For instance, a method may appear to recapitulate the taxonomy of a synthetic community with known taxonomic structure, but do a poor job when applied to real communities with poorly defined taxonomic structure or for sequences that are prone to misclassification.
 As an alternative, we developed an approach to objectively benchmark the clustering quality of OTU assignments [@Schloss2016a; @Westcott2015; @Schloss2011].
@@
 Consistently, the average neighbor algorithm was identified as among the best or the best algorithm.
+Other hierarchical algorithms such as furthest and nearest neighbor, which seek to minimize the number of FP and FN, respectively, fared significantly worse.
 The distance-based greedy clustering as implemented in VSEARCH has also performed well.
@@
 Each iteration consists of interrogating every sequence in the dataset.
-Although numerous options are available within the mothur-based implementation of the algorithm (e.g.
-sensitivity, specificity, accuracy, F1 score, etc.), the default metric is MCC because it includes all four parameters from the confusion matrix.
-The algorithm continues until the optimization metric stabilizes or until it reaches a defined stopping criteria.
+Although numerous options are available for optimizing the clusters and for assessing the quality of the clusters within the mothur-based implementation of the algorithm (e.g.
+sensitivity, specificity, accuracy, F1 score, etc.), the default metric for optimization and assessment is MCC because it includes all four parameters from the confusion matrix (Figure S1).
+The algorithm continues until the optimization metric stabilizes or until it reaches a defined stopping criteria (Figure S2).
 
@@
 ***OptiClust-generated OTUs are as stable as those from other algorithms.***
-One concern that many have with *de novo* clustering algorithms is that their output is sensitive to the initial order of the sequences.
+One concern that many have with *de novo* clustering algorithms is that their output is sensitive to the initial order of the sequences because each algorithm must break ties where a sequence could be assigned to multiple OTUs.
 An additional concern with the OptiClust algorithm is that it may stabilize at a local optimum.
@@
 Confirming our previous results, all of the methods we tested were stable to stochastic processes.
-Of the methods that involved randomization, the coefficient of variation for MCC values considerably smaller with OptiClust than the other methods and the coefficient of variation for the number of OTUs was comparable to the other methods.
+Of the methods that involved randomization, the coefficient of variation for MCC values was considerably smaller with OptiClust than the other methods and the coefficient of variation for the number of OTUs was comparable to the other methods.
 The variation observed in clustering quality suggested that the algorithm does not appear to converge to a locally optimum MCC value.
@@
 Although not as important as the quality of clustering, the amount of time and memory required to assign sequences to OTUs is a legitimate concern.
-To evaluate how the speed and memory usage scaled with the number of sequences in the dataset, we measured the time required and maximum RAM usage to cluster 20, 40, 60, 80, and 100% of the unique sequences from each of the natural datasets using the OptiClust algorithm (Figure 2).
+We observed that the time required to complete the OptiClust algorithm (Figure 1C) paralleled the number of pairwise distances that were smaller than 0.03.
+To further evaluate how the speed and memory usage scaled with the number of sequences in the dataset, we measured the time required and maximum RAM usage to cluster 20, 40, 60, 80, and 100% of the unique sequences from each of the natural datasets using the OptiClust algorithm (Figure 2).
 Within each iteration of the algorithm, each sequence is compared to every other sequence and each comparison requires a recalculation of the confusion matrix.
@@
 In fact, when we fit a power law function to the data in Figure 2B, the power coefficients were  `r format(round(median(memory_coef), 1), nsmall=1L)`.
-This analysis suggests that doubling the number of sequences in a dataset would increase the time required to cluster the data by 4 to 8-fold and increase the RAM required by 4-fold.
+Using the four natural community datasets, doubling the number of sequences in a dataset would increase the time required to cluster the data by 4 to 8-fold and increase the RAM required by 4-fold.
 It is possible that future improvements to the implementation of the algorithm could improve this performance.
@@
 We previously described a heuristic to accelerate OTU assignments where sequences were first classified to taxonomic groups and within each taxon sequences were assigned to OTUs using the average neighbor clustering algorithm [@Schloss2011].
-This accelerated the clustering and reduce the memory requirements because the number of unique sequences is effectively reduced by splitting sequences across taxonomic groups.
+This method is similar to open reference clustering except that in our approach all sequences are subjected to *de novo* clustering following classification whereas in open reference clustering only those sequences that cannot be classified are subjected to *de novo* clustering.
+Our cluster splitting approach accelerated the clustering and reduced the memory requirements because the number of unique sequences is effectively reduced by splitting sequences across taxonomic groups.
 Furthermore, because sequences in different taxonomic groups are assumed to belong to different OTUs they are independent, which permits parallelization and additional reduction in computation time.
@@
 It is expected that these errors would increase as the taxonomic level goes from kingdom to genus.
-To characterize the clustering quality, we calculated the MCC values using OptiClust, average neighbor, and DGC with VSEARCH when splitting at each taxonomic level (Figure 3).
+To characterize the clustering quality, we classified each sequence at each taxonomic level and calculated the MCC values using OptiClust, average neighbor, and DGC with VSEARCH when splitting at each taxonomic level (Figure 3).
 For each method, the MCC values decreased as the taxonomic resolution increased; however, the decrease in MCC as not as large as the difference between clustering methods.
@@
 These decreases in MCC value resulted in the formation of as many as `r format(round(min(100*((rel_otu_data %>% filter(tax_level == 3))$max_num_otus-1), na.rm=T), 1), nsmall=1L)` and `r format(round(min(100*((rel_otu_data %>% filter(tax_level == 6))$max_num_otus-1), na.rm=T), 1), nsmall=1L)`% more OTUs, respectively, than were observed from the entire dataset.
+These errors were due to the generation of additional false negatives due to splitting similar sequences into different taxonomic groups.
 For the datasets included in the current analysis, the use of the cluster splitting heuristic was probably not worth the loss in clustering quality.
@@
 We employed the MCC because it includes the four values from a confusion matrix.
-Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled [@Schloss2005; @Schloss2011].
+Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled, respectively [@Schloss2005; @Schloss2011].
 Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1 score, which is based on three values from the confusion matrix.
@@
 For example, optimizing for sensitivity could lead to a large number of FPs.
-Since we would like to minimize both FPs and FNs and not just the total number of false assignments, we decided to optimize utilizing the MCC.
-It is possible that other metrics could be developed and employed for optimization of the clustering.
+More FPs increases the number of OTUs while more FNs collapses OTUs together.
+It is difficult to know which is worse since community richness and diversity are linked to the number of OTUs.
+In addition, increasing the number of FNs would overstate the differences between communities while increasing the number of FPs would overstate their similarity.
+Therefore, it is important to jointly minimize the number of FPs and FNs.
+With this in mind, we decided to optimize utilizing the MCC.
+It is possible that other metrics that balance the four parameters could be developed and employed for optimization of the clustering.
 
 The OptiClust algorithm is relatively simple.
-For each sequence it effectively ask whether the MCC value will increase if the sequence is moved to a different OTU including creating a new OTU.
+For each sequence it effectively asks whether the MCC value will increase if the sequence is moved to a different OTU including creating a new OTU.
 If the value does not change, it remains in the current OTU.
@@
 Future improvements to the implementation of the algorithm could provide optimization to further improve its speed and susceptibility to find a local optimum.
-Users are encourage to repeat the OTU assignment several times to confirm that they have found the best OTU assignments.
+Users are encouraged to repeat the OTU assignment several times to confirm that they have found the best OTU assignments.
 
@@
 Description of datasets used to evaluate the OptiClust algorithm and compare its performance to other algorithms.** Each dataset contains sequences from the V4 region of the 16S rRNA gene.
+The number of distances for each dataset are those that were less than or equal to 0.03.
+The number of OTUs were determined using the OptiClust algorithm.
 The even and staggered datasets were generated by extracting the V4 region from full length reference sequences and the datasets from the natural communities were generated by sequencing the V4 region using a Illumina MiSeq with either paired 150 or 250 nt reads.
@@
  										format(data_summary$n_total_seqs, big.mark=','),
- 										format(data_summary$n_unique_seqs, big.mark=','))
+ 										format(data_summary$n_unique_seqs, big.mark=','),
+										format(data_summary$n_distances, big.mark=','),
+										format(round(data_summary$n_sobs), big.mark=','))
 
@@
 													"**Total Seqs.**",
-													"**Unique Seqs.**")
+													"**Unique Seqs.**",
+													"**Distances**",
+													"**OTUs**")
 
-kable(data_table, row.names=F, align="lcccc")
+kable(data_table, row.names=F, align="lcccccc")
 
@@
 
+\includegraphics[width=6.0in]{../results/figures/performance_other.png}
+
+**Figure S1.
+The OptiClust algorithm is able to effectively cluster sequences into OTUs by minimizing or maximizing numerous metrics.** Plot of MCC (A), number of OTUs (B), and execution times (C) for the comparison of output from the OptiClust algorithm when to minimizing or maximizing a variety of parameters when applied to four natural and two synthetic datasets.
+Within mothur, OTU assignments can also be made using other metrics including minimizing false positives and maximizing the specificity, positive predictive value, and true negatives; however, these all resulted in sequences being assigned to separate OTUs, which resulted in no false positives and the maximum number of true negatives.
+The error bars indicate the range of values observed for 10 replicates.
+
+\newpage
+
+\includegraphics[width=3.5in]{../results/figures/optimization.png}
+
+**Figure S2.
+The OptiClust algorithm rapidly converges to optimize the Matthew's Correlation Coefficient.** The six datasets were clustered into OTUs using the OptiClust algorithm seeking to maximize the Matthew's Correlation Coefficient.
+This was repeated 10 times for each dataset.
+
+\newpage
+
 **Supplemental text.** Worked example of how OptiClust algorithm clusters sequences into OTUs.

### 2017-02-02 submission/Westcott_OptiClust_mSphere_2016.Rmd
Add tex versions of the manuscript

@@
   pdf_document:
+    keep_tex: true
     includes:

### 2017-02-02 submission/Westcott_OptiClust_mSphere_2016.tex
Add tex versions of the manuscript

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{\textbf{OptiClust: Improved method for assigning amplicon-based sequence
data to operational taxonomic units}}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}
\usepackage{comment}

\usepackage{multirow}
\usepackage{array}
\newcommand{\bigcell}[2]{\begin{tabular}{@{}#1@{}}#2\end{tabular}}
\usepackage{caption}

\begin{document}
\maketitle

\begin{center}

Running title: OptiClust: Optimized Clustering

\vspace{25mm}
Sarah L. Westcott and Patrick D. Schloss${^\dagger}$

\vspace{30mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
\end{center}

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

Assignment of 16S rRNA gene sequences to operational taxonomic units
(OTUs) is a computational bottleneck in the process of analyzing
microbial communities. Although this has been an active area of
research, it has been difficult to overcome the time and memory demands
while improving the quality of the OTU assignments. Here we developed a
new OTU assignment algorithm that iteratively reassigns sequences to new
OTUs to optimize the Matthews correlation coefficient (MCC), a measure
of the quality of OTU assignments. To assess the new algorithm,
OptiClust, we compared it to ten other algorithms using 16S rRNA gene
sequences from two simulated and four natural communities. Using the
OptiClust algorithm, the MCC values averaged 15.2 and 16.5\% higher than
the OTUs generated when we used the average neighbor and distance-based
greedy clustering with VSEARCH, respectively. Furthermore, on average,
OptiClust was 94.6-times faster than the average neighbor algorithm and
just as fast as distance-based greedy clustering with VSEARCH. An
empirical analysis of the efficiency of the algorithms showed that the
time and memory required to perform the algorithm scaled quadratically
with the number of unique sequences in the dataset. The significant
improvement in the quality of the OTU assignments over previously
existing methods will significantly enhance downstream analysis by
limiting the splitting of similar sequences into separate OTUs and
merging of dissimilar sequences into the same OTU. The development of
the OptiClust algorithm represents a significant advance that is likely
to have numerous other applications.

\subsection{Importance}\label{importance}

The analysis of microbial communities from diverse environments using
16S rRNA gene sequencing has expanded our knowledge of the biogeography
of microorganisms. An important step in this analysis is the assignment
of sequences into taxonomic groups based on their similarity to
sequences in a database or based on their similarity to each other,
irrespective of a database. In this study, we present a new algorithm
for the latter approach. The algorithm, OptiClust, seeks to optimize a
metric of assignment quality by shuffling sequences between taxonomic
groups. We found that OptiClust produces more robust assignments and
does so in a rapid and memory efficient manner. This advance will allow
for a more robust analysis of microbial communities and the factors that
shape them.

\newpage

\subsection{Introduction}\label{introduction}

Amplicon-based sequencing has provided incredible insights into Earth's
microbial biodiversity (1, 2). It has become common for studies to
include sequencing millions of 16S rRNA gene sequences across hundreds
of samples (3, 4). This is three to four orders of magnitude greater
sequencing depth than was previously achieved using Sanger sequencing
(5, 6). The increased sequencing depth has revealed novel taxonomic
diversity that is not adequately represented in reference databases (1,
3). However, the advance has forced re-engineering of methods to
overcome the rate and memory limiting steps in computational pipelines
that process raw sequences through the generation of tables containing
the number of sequences in different taxa for each sample (7--10). A
critical component to these pipelines has been the assignment of
amplicon sequences to taxonomic units that are ether defined based on
similarity to a reference or operationally based on the similarity of
the sequences to each other within the dataset (11, 12).

A growing number of algorithms have been developed to cluster sequences
into OTUs. These algorithms can be classified into three general
categories. The first category of algorithms has been termed
closed-reference or phylotyping (13, 14). Sequences are compared to a
reference collection and clustered based on the reference sequences that
they are similar to. This approach is fast; however, the method
struggles when a sequence is similar to multiple reference sequences
that may have different taxonomies and when it is not similar to
sequences in the reference. The second category of algorithms has been
called \emph{de novo} because they assign sequences to OTUs without the
use of a reference (14). These include hierarchical algorithms such as
nearest, furthest, and average neighbor (15) and algorithms that employ
heuristics such as abundance or distance-based greedy clustering as
implemented in USEARCH (16) or VSEARCH (17), Sumaclust, OTUCLUST (18),
and Swarm (19). \emph{De novo} methods are agglomerative and tend to be
more computationally intense. It has proven difficult to know which
method generates the best assignments. A third category of algorithm is
open reference clustering, which is a hybrid approach (3, 14). Here
sequences are assigned to OTUs using closed-reference clustering and
sequences that are not within a threshold of a reference sequence are
then clustered using a \emph{de novo} approach. This category blends the
strengths and weaknesses of the other method and adds the complication
that closed-reference and \emph{de novo} clustering use different OTU
definitions. These algorithms take different approaches to handling
large datasets to minimize the time and memory requirements while
attempting to assign sequences to meaningful OTUs.

Several metrics have emerged for assessing the quality of OTU assignment
algorithms. These have included the time and memory required to run the
algorithm (3, 19--21), agreement between OTU assignments and the
sequences' taxonomy (19, 21--31), sensitivity of an algorithm to
stochastic processes (32), the number of OTUs generated by the algorithm
(22, 33), and the ability to regenerate the assignments made by other
algorithms (3, 34). Unfortunately, these methods fail to directly
quantify the quality of the OTU assignments. An algorithm may complete
with minimal time and memory requirements or generate an idealized
number of OTUs, but the composition of the OTUs could be incorrect.
These metrics also tend to be subjective. For instance, a method may
appear to recapitulate the taxonomy of a synthetic community with known
taxonomic structure, but do a poor job when applied to real communities
with poorly defined taxonomic structure or for sequences that are prone
to misclassification. As an alternative, we developed an approach to
objectively benchmark the clustering quality of OTU assignments (13, 35,
36). This approach counts the number of true positives (TP), true
negatives (TN), false positives (FP), and false negatives (FN) based on
the pairwise distances. Sequence pairs that are within the
user-specified threshold and are clustered together represent TPs and
those in different OTUs are FNs. Those sequence pairs that have a
distance larger than the threshold and are not clustered in the same OTU
are TNs and those in the same OTU are FPs. These values can be
synthesized into a single correlation coefficient, the Matthews'
Correlation Coefficient (MCC), which measures the correlation between
observed and predicted classifications and is robust to cases where
there is an uneven distribution across the confusion matrix (37).
Consistently, the average neighbor algorithm was identified as among the
best or the best algorithm. Other hierarchical algorithms such as
furthest and nearest neighbor, which seek to minimize the number of FP
and FN, respectively, fared significantly worse. The distance-based
greedy clustering as implemented in VSEARCH has also performed well. The
computational resources required to complete the average neighbor
algorithm can be significant for large datasets and so there is a need
for an algorithm that efficiently produces consistently high quality OTU
assignments.

These previous efforts have assessed the quality of the clusters after
the completion of the algorithm. In the current study we developed and
benchmarked a new \emph{de novo} clustering algorithm that uses real
time calculation of the MCC to direct the progress of the clustering.
The result is the OptiClust algorithm, which produces significantly
better sequence assignments while making efficient use of computational
resources.

\subsection{Results}\label{results}

\textbf{\emph{OptiClust algorithm.}} The OptiClust algorithm uses the
pairs of sequences that are within a desired threshold of each other
(e.g.~0.03), a list of all sequence names in the dataset, and the metric
that should be used to assess clustering quality. A detailed description
of the algorithm is provided for a toy dataset in the Supplementary
Material. Briefly, the algorithm starts by placing each sequence either
within its own OTU or into a single OTU. The algorithm proceeds by
interrogating each sequence and re-calculating the metric for the cases
where the sequence stays in its current OTU, is moved to each of the
other OTUs, or is moved into a new OTU. The location that results in the
best clustering quality indicates whether the sequence should remain in
its current OTU or be moved to a different or new OTU. Each iteration
consists of interrogating every sequence in the dataset. Although
numerous options are available for optimizing the clusters and for
assessing the quality of the clusters within the mothur-based
implementation of the algorithm (e.g.~sensitivity, specificity,
accuracy, F1 score, etc.), the default metric for optimization and
assessment is MCC because it includes all four parameters from the
confusion matrix (Figure S1). The algorithm continues until the
optimization metric stabilizes or until it reaches a defined stopping
criteria (Figure S2).

\textbf{\emph{OptiClust-generated OTUs are more robust than those from
other methods.}} To evaluate the OptiClust algorithm and compare its
performance to other algorithms, we utilized six datasets including two
synthetic communities and four previously published large datasets
generated from soil, marine, human, and murine samples (Table 1). When
we seeded the OptiClust algorithm with each sequence in a separate OTU
and ran the algorithm until complete convergence, the MCC values
averaged 15.2 and 16.5\% higher than the OTUs using average neighbor and
distance-based greedy clustering (DGC) with VSEARCH, respectively
(Figure 1). The number of OTUs formed by the various methods was
negatively correlated with their MCC value (\(\rho\)=-0.47; p=0). The
OptiClust algorithm was considerably faster than the hierarchical
algorithms and somewhat slower than the heuristic-based algorithms.
Across the six datasets, the OptiClust algorithm was 94.6-times faster
than average neighbor and just as fast as DGC with VSEARCH. The human
dataset was a challenge for a number of the algorithms. OTUCLUST and
SumaClust were unable to cluster the human dataset in less than 50 hours
and the average neighbor algorithm required more than 45 GB of RAM. The
USEARCH-based methods were unable to cluster the human data using the
32-bit free version of the software that limits the amount of RAM to
approximately 3.5 GB. These data demonstrate that OptiClust generated
significantly more robust OTU assignments than existing methods across a
diverse collection of datasets with performance that was comparable to
popular methods.

\textbf{\emph{OptiClust stopping criteria.}} By default, the
mothur-based implementation of the algorithm stops when the optimization
metric changes by less than 0.0001; however, this can be altered by the
user. This implementation also allows the user to stop the algorithm if
a maximum number of iterations is exceeded. By default mothur uses a
maximum value of 100 iterations. The justification for allowing
incomplete convergence was based on the observation that numerous
iterations are performed that extend the time required to complete the
clustering with minimal improvement in clustering. We evaluated the
results of clustering to partial convergence (i.e.~a change in the MCC
value that was less than 0.0001) or until complete convergence of the
MCC value (i.e.~until it did not change between iterations) when seeding
the algorithm with each sequence in a separate OTU (Figure 1). The small
difference in MCC values between the output from partial and complete
convergence resulted in a difference in the median number of OTUs that
ranged between 1.5 and 17.0 OTUs. This represented a difference of less
than 0.15\%. Among the four natural datasets, between 3 and 6 were
needed to achieve partial convergence and between 8 and 12.50 iterations
were needed to reach full convergence. The additional steps required
between 1.4 and 1.7 times longer to complete the algorithm. These
results suggest that achieving full convergence of the optimization
metric adds computational effort; however, considering full convergence
took between 2 and 17 minutes the extra effort was relatively small.
Although the mothur's default setting is partial convergence, the
remainder of our analysis used complete convergence to be more
conservative.

\textbf{\emph{Effect of seeding OTUs on OptiClust performance.}} By
default the mothur implementation of the OptiClust algorithm starts with
each sequence in a separate OTU. An alternative approach is to start
with all of the sequences in a single OTU. We found that the MCC values
for clusters generated seeding OptiClust with the sequences as a single
OTU were between 0 and 11.5\% lower than when seeding the algorithm with
sequences in separate OTUs (Figure 1). Interestingly, with the exception
of the human dataset (0.2\% more OTUs), the number of OTUs was as much
as 7.0\% lower (mice) than when the algorithm was seeded with sequence
in separate OTUs. Finally, the amount of time required to cluster the
data when the algorithm was seeded with a single OTU was between 1.5 and
2.9-times longer than if sequences were seeded as separate OTUs. This
analysis demonstrates that seeding the algorithm with sequences as
separate OTUs resulted in the best OTU assignments in the shortest
amount of time.

\textbf{\emph{OptiClust-generated OTUs are as stable as those from other
algorithms.}} One concern that many have with \emph{de novo} clustering
algorithms is that their output is sensitive to the initial order of the
sequences because each algorithm must break ties where a sequence could
be assigned to multiple OTUs. An additional concern with the OptiClust
algorithm is that it may stabilize at a local optimum. To evaluate these
concerns we compared the results obtained using ten randomizations of
the order that sequences were given to the algorithm. The median the
coefficient of variation across the six datasets for MCC values obtained
from the replicate clusterings using OptiClust was 0.1\% (Figure 1). We
also measured the coefficient of variation for the number of OTUs across
the six datasets for each method. The median coefficient of variation
for the number of OTUs generated using OptiClust was 0.1\%. Confirming
our previous results, all of the methods we tested were stable to
stochastic processes. Of the methods that involved randomization, the
coefficient of variation for MCC values was considerably smaller with
OptiClust than the other methods and the coefficient of variation for
the number of OTUs was comparable to the other methods. The variation
observed in clustering quality suggested that the algorithm does not
appear to converge to a locally optimum MCC value. More importantly, the
random variation does yield output of a similarly high quality.

\textbf{\emph{Time and memory required to complete Optimization-based
clustering scales efficiently.}} Although not as important as the
quality of clustering, the amount of time and memory required to assign
sequences to OTUs is a legitimate concern. We observed that the time
required to complete the OptiClust algorithm (Figure 1C) paralleled the
number of pairwise distances that were smaller than 0.03. To further
evaluate how the speed and memory usage scaled with the number of
sequences in the dataset, we measured the time required and maximum RAM
usage to cluster 20, 40, 60, 80, and 100\% of the unique sequences from
each of the natural datasets using the OptiClust algorithm (Figure 2).
Within each iteration of the algorithm, each sequence is compared to
every other sequence and each comparison requires a recalculation of the
confusion matrix. This would result in a worst case algorithmic
complexity on the order of N\^{}3, where N is the number of unique
sequences. Because the algorithm only needs to keep track of the
sequence pairs that are within the threshold of each other, it is likely
that the implementation of the algorithm is more efficient. To
empirically determine the algorithmic complexity, we fit a power law
function to the data in Figure 2A. We observed power coefficients
between 1.7 and 2.5 for the marine and human datasets, respectively. The
algorithm requires storing a matrix that contains the pairs of sequences
that are close to each other as well as a matrix that indicates which
sequences are clustered together. The memory required to store these
matrices is on the order of N\^{}2, where N is the number of unique
sequences. In fact, when we fit a power law function to the data in
Figure 2B, the power coefficients were 1.9. Using the four natural
community datasets, doubling the number of sequences in a dataset would
increase the time required to cluster the data by 4 to 8-fold and
increase the RAM required by 4-fold. It is possible that future
improvements to the implementation of the algorithm could improve this
performance.

\textbf{\emph{Cluster splitting heuristic generates OTUs that are as
good as non-split approach.}} We previously described a heuristic to
accelerate OTU assignments where sequences were first classified to
taxonomic groups and within each taxon sequences were assigned to OTUs
using the average neighbor clustering algorithm (13). This method is
similar to open reference clustering except that in our approach all
sequences are subjected to \emph{de novo} clustering following
classification whereas in open reference clustering only those sequences
that cannot be classified are subjected to \emph{de novo} clustering.
Our cluster splitting approach accelerated the clustering and reduced
the memory requirements because the number of unique sequences is
effectively reduced by splitting sequences across taxonomic groups.
Furthermore, because sequences in different taxonomic groups are assumed
to belong to different OTUs they are independent, which permits
parallelization and additional reduction in computation time. Reduction
in clustering quality are encountered in this approach if there are
errors in classification or if two sequences within the desired
threshold belong to different taxonomic groups. It is expected that
these errors would increase as the taxonomic level goes from kingdom to
genus. To characterize the clustering quality, we classified each
sequence at each taxonomic level and calculated the MCC values using
OptiClust, average neighbor, and DGC with VSEARCH when splitting at each
taxonomic level (Figure 3). For each method, the MCC values decreased as
the taxonomic resolution increased; however, the decrease in MCC as not
as large as the difference between clustering methods. As the resolution
of the taxonomic levels increased, the clustering quality remained high,
relative to clusters formed from the entire dataset
(i.e.~kingdom-level). The MCC values when splitting the datasets at the
class and genus levels were within 98.0 and 93.0\%, respectively, of the
MCC values obtained from the entire dataset. These decreases in MCC
value resulted in the formation of as many as 4.7 and 22.5\% more OTUs,
respectively, than were observed from the entire dataset. These errors
were due to the generation of additional false negatives due to
splitting similar sequences into different taxonomic groups. For the
datasets included in the current analysis, the use of the cluster
splitting heuristic was probably not worth the loss in clustering
quality. However, as datasets become larger, it may be necessary to use
the heuristic to clustering the data into OTUs.

\subsection{Discussion}\label{discussion}

Myriad methods have been proposed for assigning 16S rRNA gene sequences
to OTUs that each claim improved performance based on speed, memory
usage, representation of taxonomic information, and number of OTUs. Each
of these metrics is subjective and do not actually indicate the quality
of the clustering. This led us to propose using the MCC as a metric for
assessing the quality of clustering, post hoc. Here, we described a new
clustering method that seeks to optimize clustering based on an
objective criterion that measures clustering quality in real time. In
the OptiClust algorithm clustering is driven by optimizing a metric that
assesses whether any two sequences should be grouped into the same OTU.
The result is clusters that are significantly more robust and is
efficient in the time and memory required to cluster the sequences into
OTUs. This makes it more tractable to analyze large datasets without
sacrificing clustering quality as was previously necessary using
heuristic methods.

The cluster optimization procedure is dependent on the metric that is
chosen for optimization. We employed the MCC because it includes the
four values from a confusion matrix. Other algorithms such as the
furthest neighbor and nearest neighbor algorithms minimize the number of
FP and FN, respectively; however, these suffer because the number of FN
and FP are not controlled, respectively (13, 15). Alternatively, one
could optimize based on the sensitivity, specificity, or accuracy, which
are each based on two values from the confusion matrix or they could
optimize based on the F1 score, which is based on three values from the
confusion matrix. Because these metrics do not balance all four
parameters equally, it is likely that one parameter will dominate in the
optimization procedure. For example, optimizing for sensitivity could
lead to a large number of FPs. More FPs increases the number of OTUs
while more FNs collapses OTUs together. It is difficult to know which is
worse since community richness and diversity are linked to the number of
OTUs. In addition, increasing the number of FNs would overstate the
differences between communities while increasing the number of FPs would
overstate their similarity. Therefore, it is important to jointly
minimize the number of FPs and FNs. With this in mind, we decided to
optimize utilizing the MCC. It is possible that other metrics that
balance the four parameters could be developed and employed for
optimization of the clustering.

The OptiClust algorithm is relatively simple. For each sequence it
effectively asks whether the MCC value will increase if the sequence is
moved to a different OTU including creating a new OTU. If the value does
not change, it remains in the current OTU. The algorithm repeats until
the MCC value stabilizes. Assuming that the algorithm is seeded with
each sequence in a separate OTU, it does not appear that the algorithm
converges to a local optimum. Furthermore, execution of the algorithm
with different random number generator seeds produces OTU assignments of
consistently high quality. Future improvements to the implementation of
the algorithm could provide optimization to further improve its speed
and susceptibility to find a local optimum. Users are encouraged to
repeat the OTU assignment several times to confirm that they have found
the best OTU assignments.

Our previous MCC-based analysis of clustering algorithms indicated that
the average neighbor algorithm consistently produced the best OTU
assignments with the DGC-based method using USEARCH also producing
robust OTU assignments. The challenge in using the average neighbor
algorithm is that it requires a large amount of RAM and is
computationally demanding. This led to the development of a splitting
approach that divides the clustering across distinct taxonomic groups
(13). The improved performance provided by the OptiClust algorithm
likely makes such splitting unnecessary for most current datasets. We
have demonstrated that although the OTU assignments made at the genus
level are still better than that of other methods, the quality is not as
good as that found without splitting. The loss of quality is likely due
to misclassification because of limitations in the clustering algorithms
and reference databases. The practical significance of such small
differences in clustering quality remain to be determined; however,
based on the current analysis, it does appear that the number of OTUs is
artificially inflated. Regardless, the best clustering quality should be
pursued given the available computer resources.

The time and memory required to execute the OptiClust algorithm scaled
proportionally to the number of unique sequences raised to the second
power. The power for the time requirement is affected by the similarity
of the sequences in the dataset with datasets containing more similar
sequences having a higher power. Also, the number of unique sequences is
the basis for both the amount of time and memory required to complete
the algorithm. Both the similarity of sequences and number of unique
sequences can be driven by the sequencing error since any errors will
increase the number of unique sequences and these sequences will be
closely related to the perfect sequence. This underscores the importance
of reducing the noise in the sequence data (7). If sequencing errors are
not remediated and are relatively randomly distributed, then it is
likely that the algorithm will require an unnecessary amount of time and
RAM to complete.

The rapid expansion in sequencing capacity has demanded that the
algorithms used to assign 16S rRNA gene sequences to OTUs be efficient
while maintaining robust assignments. Although database-based approaches
have been proposed to facilitate this analysis, they are limited by
their limited coverage of bacterial taxonomy and by the inconsistent
process used to name taxa. The ability to assign sequences to OTUs using
an algorithm that optimizes clustering by directly measuring quality
will significantly enhance downstream analysis. The development of the
OptiClust algorithm represents a significant advance that is likely to
have numerous other applications.

\subsection{Materials and Methods}\label{materials-and-methods}

\textbf{\emph{Sequence data and processing steps.}} To evaluate the
OptiClust and the other algorithms we created two synthetic sequence
collections and four sequence collections generated from previously
published studies. The V4 region of the 16S rRNA gene was used from all
datasets because it is a popular region that can be fully sequenced with
two-fold coverage using the commonly used MiSeq sequencer from Illumina
(7). The method for generating the simulated datasets followed the
approach used by Kopylova et al. (33) and Schloss (35). Briefly, we
randomly selected 10,000 uniques V4 fragments from 16S rRNA gene
sequences that were unique from the SILVA non-redundant database (38). A
community with an even relative abundance profile was generated by
specifying that each sequence had a frequency of 100 reads. A community
with a staggered relative abundance profile was generated by specifying
that the abundance of each sequence was a randomly drawn integer sampled
from a uniform distribution between 1 and 200. Sequence collections
collected from human feces (39), murine feces (40), soil (41), and
seawater (42) were used to characterize the algorithms' performance with
natural communities. These sequence collections were all generated using
paired 150 or 250 nt reads of the V4 region. We re-processed all of the
reads using a common analysis pipeline that included quality score-based
error correction (7), alignment against a SILVA reference database (38,
43), screening for chimeras using UCHIME (9), and classification using a
naive Bayesian classifier with the RDP training set requiring an 80\%
confidence score (10).

\textbf{\emph{Implementation of clustering algorithms.}} In addition to
the OptiClust algorithm we evaluated ten different \emph{de novo}
clustering algorithms. These included three hierarchical algorithms,
average neighbor, nearest neighbor, and furthest neighbor, which are
implemented in mothur (v.1.39.0) (11). Seven heuristic methods were also
used including abundance-based greedy clustering (AGC) and
(distance-based greedy clustering) DGC as implemented in USEARCH (v.6.1)
(16) and VSEARCH (v.2.3.3) ((17){]}, OTUCLUST (v.0.1) (18), SumaClust
(v.1.0.20), and Swarm (v.2.1.9) (19). With the exception of Swarm each
of these methods uses distance-based thresholds to report OTU
assignments.

\textbf{\emph{Benchmarking.}} We evaluated the quality of the sequence
clustering, reproducibility of the clustering, the speed of clustering,
and the amount of memory required to complete the clustering. To assess
the quality of the clusters generated by each method, we counted the
cells within a confusion matrix that indicated how well the clusterings
represented the distances between the pair of sequences (13). Pairs of
sequences that were in the same OTU and had a distance less than 3\%
were true positives (TPs), those that were in different OTUs and had a
distance greater than 3\% were true negatives (TNs), those that were in
the same OTU and had a distance greater than 3\% were false positives
(FPs), and those that were in different OTUs and had a distance less
than 3\% were false negatives (FNs). To synthesize the matrix into a
single metric we used the Matthews Correlation Coefficient using the
\texttt{sens.spec} command in mothur using the following equations.

\[
MCC = \frac{TP \times TN-FP \times FN}{\sqrt{(TP+FP)(TP+FN)(TN+FP)(TN+FN)} }
\]

To assess the reproducibility of the algorithms we randomized the
starting order of each sequence collection ten times and ran each
algorithm on each randomized collection. We then measured the MCC for
each randomization and quantified their percent coefficient of variation
(\% CV; 100 times the ratio of the standard deviation to the mean).

To assess how the the memory and time requirements scaled with the
number of sequences included in each sequence collection, we randomly
subsampled 20, 40, 60, or 80\% of the unique sequences in each
collection. We obtained 10 subsamples at each depth for each dataset and
ran each collection (N= 50 = 5 sequencing depths x 10 replicates)
through each of the algorithms. We used the timeout script to quantify
the maximum RAM used and the amount of time required to process each
sequence collection (\url{https://github.com/pshved/timeout}). We
limited each algorithm to 45 GB of RAM and 50 hours using a single
processor.

\textbf{\emph{Data and code availability.}} The workflow utilized
commands in GNU make (v.3.81), GNU bash (v.4.1.2), mothur (v.1.39.0)
(11), and R (v.3.3.2) (44). Within R we utilized the wesanderson
(v.0.3.2) (45), dplyr (v.0.5.0) (46), tidyr (v.0.6.0) (47), cowplot
(v.0.6.3) (48), and ggplot2 (v.2.2.0.9000) (49) packages. A reproducible
version of this manuscript and analysis is available at
\url{https://github.com/SchlossLab/Westcott_OptiClust_mSphere_2017}.

\subsection{Acknowledgements}\label{acknowledgements}

This work was supported through funding from the National Institutes of
Health to PDS (P30DK034933). SLW designed, implemented, and evaluated
the algorithm. PDS designed and evaluated the algorithm. Both authors
wrote and edited the manuscript.

\newpage

\textbf{Table 1. Description of datasets used to evaluate the OptiClust
algorithm and compare its performance to other algorithms.} Each dataset
contains sequences from the V4 region of the 16S rRNA gene. The number
of distances for each dataset are those that were less than or equal to
0.03. The number of OTUs were determined using the OptiClust algorithm.
The even and staggered datasets were generated by extracting the V4
region from full length reference sequences and the datasets from the
natural communities were generated by sequencing the V4 region using a
Illumina MiSeq with either paired 150 or 250 nt reads.

\begin{longtable}[c]{@{}lcccccc@{}}
\toprule
\textbf{Dataset (Ref.)} & \textbf{Read Length} & \textbf{Samples} &
\textbf{Total Seqs.} & \textbf{Unique Seqs.} & \textbf{Distances} &
\textbf{OTUs}\tabularnewline
\midrule
\endhead
Soil (41) & 150 & 18 & 948,243 & 143,677 & 11,775,167 &
40,216\tabularnewline
Marine (42) & 250 & 7 & 1,384,988 & 75,923 & 12,908,857 &
25,787\tabularnewline
Mice (40) & 250 & 360 & 2,825,495 & 32,447 & 6,988,306 &
2,658\tabularnewline
Human (39) & 250 & 489 & 20,951,841 & 121,281 & 38,544,315 &
11,648\tabularnewline
Even (33, 35) & NA & NA & 1,155,800 & 11,558 & 29,694 &
7,651\tabularnewline
Staggered (33, 35) & NA & NA & 1,156,550 & 11,558 & 29,694 &
7,653\tabularnewline
\bottomrule
\end{longtable}

\newpage

\includegraphics[width=6.0in]{../results/figures/performance.png}

\textbf{Figure 1. Comparison of de novo clustering algorithms.} Plot of
MCC (A), number of OTUs (B), and execution times (C) for the comparison
of \emph{de novo} clustering algorithms when applied to four natural and
two synthetic datasets. The first three columns of each figure contain
the results of clustering the datasets (i) seeding the algorithm with
one sequence per OTU and allowing the algorithm to proceed until the MCC
value no longer changed; (ii) seeding the algorithm with one sequence
per OTU and allowing the algorithm to proceed until the MCC changed by
less than 0.0001; (iii) seeding the algorithm with all of the sequences
in one OTU and allowing the algorithm to proceed until the MCC value no
longer changed. The human dataset could not be clustered by the average
neighbor, Sumaclust, USEARCH, or OTUCLUST with less than 45 GB of RAM or
50 hours of execution time. The median of 10 re-orderings of the data is
presented for each method and dataset. The range of observed values is
indicated by the error bars, which are typically smaller than the
plotting symbol.

\newpage

\includegraphics[width=3.5in]{../results/figures/speed_memory.png}

\textbf{Figure 2. OptiClust performance} The average execution time (A)
and memory usage (B) required to cluster the four natural datasets. The
confidence intervals indicate the range between the minimum and maximum
values. The y-axis is scaled by the square root to demonstrate the
relationship between the time and memory requirements relative to the
number of unique sequences squared.

\newpage

\includegraphics[width=3.5in]{../results/figures/split_mcc.png}

\textbf{Figure 3. Effects of taxonomically splitting the datasets on
clustering quality.} The datasets were split at each taxonomic level
based on their classification using a naive Bayesian classifier and
clustered using average neighbor, VSEARCH-based DGC, and OptiClust.

\newpage

\includegraphics[width=6.0in]{../results/figures/performance_other.png}

\textbf{Figure S1. The OptiClust algorithm is able to effectively
cluster sequences into OTUs by minimizing or maximizing numerous
metrics.} Plot of MCC (A), number of OTUs (B), and execution times (C)
for the comparison of output from the OptiClust algorithm when to
minimizing or maximizing a variety of parameters when applied to four
natural and two synthetic datasets. Within mothur, OTU assignments can
also be made using other metrics including minimizing false positives
and maximizing the specificity, positive predictive value, and true
negatives; however, these all resulted in sequences being assigned to
separate OTUs, which resulted in no false positives and the maximum
number of true negatives. The error bars indicate the range of values
observed for 10 replicates.

\newpage

\includegraphics[width=3.5in]{../results/figures/optimization.png}

\textbf{Figure S2. The OptiClust algorithm rapidly converges to optimize
the Matthew's Correlation Coefficient.} The six datasets were clustered
into OTUs using the OptiClust algorithm seeking to maximize the
Matthew's Correlation Coefficient. This was repeated 10 times for each
dataset.

\newpage

\textbf{Supplemental text.} Worked example of how OptiClust algorithm
clusters sequences into OTUs.

\newpage

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-Schloss2016b}{}
1. \textbf{Schloss PD}, \textbf{Girard RA}, \textbf{Martin T},
\textbf{Edwards J}, \textbf{Thrash JC}. 2016. Status of the archaeal and
bacterial census: An update. mBio \textbf{7}:e00201--16.
doi:\href{https://doi.org/10.1128/mbio.00201-16}{10.1128/mbio.00201-16}.

\hypertarget{ref-Locey2016}{}
2. \textbf{Locey KJ}, \textbf{Lennon JT}. 2016. Scaling laws predict
global microbial diversity. Proceedings of the National Academy of
Sciences \textbf{113}:5970--5975.
doi:\href{https://doi.org/10.1073/pnas.1521291113}{10.1073/pnas.1521291113}.

\hypertarget{ref-Rideout2014}{}
3. \textbf{Rideout JR}, \textbf{He Y}, \textbf{Navas-Molina JA},
\textbf{Walters WA}, \textbf{Ursell LK}, \textbf{Gibbons SM},
\textbf{Chase J}, \textbf{McDonald D}, \textbf{Gonzalez A},
\textbf{Robbins-Pianka A}, \textbf{Clemente JC}, \textbf{Gilbert JA},
\textbf{Huse SM}, \textbf{Zhou H-W}, \textbf{Knight R}, \textbf{Caporaso
JG}. 2014. Subsampled open-reference clustering creates consistent,
comprehensive OTU definitions and scales to billions of sequences. PeerJ
\textbf{2}:e545.
doi:\href{https://doi.org/10.7717/peerj.545}{10.7717/peerj.545}.

\hypertarget{ref-Huttenhower2012}{}
4. \textbf{Consortium THMP}. 2012. Structure, function and diversity of
the healthy human microbiome. Nature \textbf{486}:207--214.
doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.

\hypertarget{ref-Eckburg2005}{}
5. \textbf{Eckburg PB}, \textbf{Bik EM}, \textbf{Bernstein CN},
\textbf{Purdom E}, \textbf{Dethlefsen L}, \textbf{Sargent M},
\textbf{Gill SR}, \textbf{Nelson KE}, \textbf{Relman DA}. 2005.
Diversity of the human intestinal microbial flora. Science
\textbf{308}:1635--1638.

\hypertarget{ref-Elshahed2008}{}
6. \textbf{Elshahed MS}, \textbf{Youssef NH}, \textbf{Spain AM},
\textbf{Sheik C}, \textbf{Najar FZ}, \textbf{Sukharnikov LO},
\textbf{Roe BA}, \textbf{Davis JP}, \textbf{Schloss PD}, \textbf{Bailey
VL}, \textbf{Krumholz LR}. 2008. Novelty and uniqueness patterns of rare
members of the soil biosphere. Applied and Environmental Microbiology
\textbf{74}:5422--5428.
doi:\href{https://doi.org/10.1128/aem.00410-08}{10.1128/aem.00410-08}.

\hypertarget{ref-Kozich2013}{}
7. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq Illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-Schloss2009a}{}
8. \textbf{Schloss PD}. 2009. A high-throughput DNA sequence aligner for
microbial ecology studies. PLOS ONE \textbf{4}:e8230.
doi:\href{https://doi.org/10.1371/journal.pone.0008230}{10.1371/journal.pone.0008230}.

\hypertarget{ref-Edgar2011}{}
9. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
\textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves sensitivity
and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.

\hypertarget{ref-Wang2007}{}
10. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
of rRNA sequences into the new bacterial taxonomy. Applied and
Environmental Microbiology \textbf{73}:5261--5267.
doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.

\hypertarget{ref-Schloss2009b}{}
11. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Caporaso2010}{}
12. \textbf{Caporaso JG}, \textbf{Kuczynski J}, \textbf{Stombaugh J},
\textbf{Bittinger K}, \textbf{Bushman FD}, \textbf{Costello EK},
\textbf{Fierer N}, \textbf{Peña AG}, \textbf{Goodrich JK},
\textbf{Gordon JI}, \textbf{Huttley GA}, \textbf{Kelley ST},
\textbf{Knights D}, \textbf{Koenig JE}, \textbf{Ley RE},
\textbf{Lozupone CA}, \textbf{McDonald D}, \textbf{Muegge BD},
\textbf{Pirrung M}, \textbf{Reeder J}, \textbf{Sevinsky JR},
\textbf{Turnbaugh PJ}, \textbf{Walters WA}, \textbf{Widmann J},
\textbf{Yatsunenko T}, \textbf{Zaneveld J}, \textbf{Knight R}. 2010.
QIIME allows analysis of high-throughput community sequencing data.
Nature Methods \textbf{7}:335--336.
doi:\href{https://doi.org/10.1038/nmeth.f.303}{10.1038/nmeth.f.303}.

\hypertarget{ref-Schloss2011}{}
13. \textbf{Schloss PD}, \textbf{Westcott SL}. 2011. Assessing and
improving methods used in operational taxonomic unit-based approaches
for 16S rRNA gene sequence analysis. Applied and Environmental
Microbiology \textbf{77}:3219--3226.
doi:\href{https://doi.org/10.1128/aem.02810-10}{10.1128/aem.02810-10}.

\hypertarget{ref-NavasMolina2013}{}
14. \textbf{Navas-Molina JA}, \textbf{Peralta-Sánchez JM},
\textbf{González A}, \textbf{McMurdie PJ}, \textbf{Vázquez-Baeza Y},
\textbf{Xu Z}, \textbf{Ursell LK}, \textbf{Lauber C}, \textbf{Zhou H},
\textbf{Song SJ}, \textbf{Huntley J}, \textbf{Ackermann GL},
\textbf{Berg-Lyons D}, \textbf{Holmes S}, \textbf{Caporaso JG},
\textbf{Knight R}. 2013. Advancing our understanding of the human
microbiome using QIIME, pp. 371--444. \emph{In} Methods in enzymology.
Elsevier BV.

\hypertarget{ref-Schloss2005}{}
15. \textbf{Schloss PD}, \textbf{Handelsman J}. 2005. Introducing DOTUR,
a computer program for defining operational taxonomic units and
estimating species richness. Applied and Environmental microbiology
\textbf{71}:1501--1506.

\hypertarget{ref-Edgar2010}{}
16. \textbf{Edgar RC}. 2010. Search and clustering orders of magnitude
faster than BLAST. Bioinformatics \textbf{26}:2460--2461.
doi:\href{https://doi.org/10.1093/bioinformatics/btq461}{10.1093/bioinformatics/btq461}.

\hypertarget{ref-Rognes2016}{}
17. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.

\hypertarget{ref-Albanese2015}{}
18. \textbf{Albanese D}, \textbf{Fontana P}, \textbf{Filippo CD},
\textbf{Cavalieri D}, \textbf{Donati C}. 2015. MICCA: A complete and
accurate software for taxonomic profiling of metagenomic data.
Scientific Reports \textbf{5}:9743.
doi:\href{https://doi.org/10.1038/srep09743}{10.1038/srep09743}.

\hypertarget{ref-Mah2014}{}
19. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
\textbf{Vargas C de}, \textbf{Dunthorn M}. 2014. Swarm: Robust and fast
clustering method for amplicon-based studies. PeerJ \textbf{2}:e593.
doi:\href{https://doi.org/10.7717/peerj.593}{10.7717/peerj.593}.

\hypertarget{ref-Sun2009}{}
20. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Liu L}, \textbf{Yu F},
\textbf{Farrell ML}, \textbf{McKendree W}, \textbf{Farmerie W}. 2009.
ESPRIT: Estimating species richness using large collections of 16S rRNA
pyrosequences. Nucleic Acids Research \textbf{37}:e76--e76.
doi:\href{https://doi.org/10.1093/nar/gkp285}{10.1093/nar/gkp285}.

\hypertarget{ref-Cai2011}{}
21. \textbf{Cai Y}, \textbf{Sun Y}. 2011. ESPRIT-tree: Hierarchical
clustering analysis of millions of 16S rRNA pyrosequences in quasilinear
computational time. Nucleic Acids Research \textbf{39}:e95--e95.
doi:\href{https://doi.org/10.1093/nar/gkr349}{10.1093/nar/gkr349}.

\hypertarget{ref-Edgar2013}{}
22. \textbf{Edgar RC}. 2013. UPARSE: Highly accurate OTU sequences from
microbial amplicon reads. Nature Methods \textbf{10}:996--998.
doi:\href{https://doi.org/10.1038/nmeth.2604}{10.1038/nmeth.2604}.

\hypertarget{ref-Mah2015}{}
23. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
\textbf{Vargas C de}, \textbf{Dunthorn M}. 2015. Swarm v2:
Highly-scalable and high-resolution amplicon clustering. PeerJ
\textbf{3}:e1420.
doi:\href{https://doi.org/10.7717/peerj.1420}{10.7717/peerj.1420}.

\hypertarget{ref-Barriuso2011}{}
24. \textbf{Barriuso J}, \textbf{Valverde JR}, \textbf{Mellado RP}.
2011. Estimation of bacterial diversity using next generation sequencing
of 16S rDNA: A comparison of different workflows. BMC Bioinformatics
\textbf{12}:473.
doi:\href{https://doi.org/10.1186/1471-2105-12-473}{10.1186/1471-2105-12-473}.

\hypertarget{ref-Bonder2012}{}
25. \textbf{Bonder MJ}, \textbf{Abeln S}, \textbf{Zaura E},
\textbf{Brandt BW}. 2012. Comparing clustering and pre-processing in
taxonomy analysis. Bioinformatics \textbf{28}:2891--2897.
doi:\href{https://doi.org/10.1093/bioinformatics/bts552}{10.1093/bioinformatics/bts552}.

\hypertarget{ref-Chen2013}{}
26. \textbf{Chen W}, \textbf{Zhang CK}, \textbf{Cheng Y}, \textbf{Zhang
S}, \textbf{Zhao H}. 2013. A comparison of methods for clustering 16S
rRNA sequences into OTUs. PLOS ONE \textbf{8}:e70837.
doi:\href{https://doi.org/10.1371/journal.pone.0070837}{10.1371/journal.pone.0070837}.

\hypertarget{ref-Huse2010}{}
27. \textbf{Huse SM}, \textbf{Welch DM}, \textbf{Morrison HG},
\textbf{Sogin ML}. 2010. Ironing out the wrinkles in the rare biosphere
through improved OTU clustering. Environmental Microbiology
\textbf{12}:1889--1898.
doi:\href{https://doi.org/10.1111/j.1462-2920.2010.02193.x}{10.1111/j.1462-2920.2010.02193.x}.

\hypertarget{ref-May2014}{}
28. \textbf{May A}, \textbf{Abeln S}, \textbf{Crielaard W},
\textbf{Heringa J}, \textbf{Brandt BW}. 2014. Unraveling the outcome of
16S rDNA-based taxonomy analysis through mock data and simulations.
Bioinformatics \textbf{30}:1530--1538.
doi:\href{https://doi.org/10.1093/bioinformatics/btu085}{10.1093/bioinformatics/btu085}.

\hypertarget{ref-Sun2011}{}
29. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Huse SM}, \textbf{Knight R},
\textbf{Farmerie WG}, \textbf{Wang X}, \textbf{Mai V}. 2011. A
large-scale benchmark study of existing algorithms for
taxonomy-independent microbial community analysis. Briefings in
Bioinformatics \textbf{13}:107--121.
doi:\href{https://doi.org/10.1093/bib/bbr009}{10.1093/bib/bbr009}.

\hypertarget{ref-White2010}{}
30. \textbf{White JR}, \textbf{Navlakha S}, \textbf{Nagarajan N},
\textbf{Ghodsi M-R}, \textbf{Kingsford C}, \textbf{Pop M}. 2010.
Alignment and clustering of phylogenetic markers - implications for
microbial diversity studies. BMC Bioinformatics \textbf{11}:152.
doi:\href{https://doi.org/10.1186/1471-2105-11-152}{10.1186/1471-2105-11-152}.

\hypertarget{ref-AlGhalith2016}{}
31. \textbf{Al-Ghalith GA}, \textbf{Montassier E}, \textbf{Ward HN},
\textbf{Knights D}. 2016. NINJA-OPS: Fast accurate marker gene alignment
using concatenated ribosomes. PLOS Computational Biology
\textbf{12}:e1004658.
doi:\href{https://doi.org/10.1371/journal.pcbi.1004658}{10.1371/journal.pcbi.1004658}.

\hypertarget{ref-He2015}{}
32. \textbf{He Y}, \textbf{Caporaso JG}, \textbf{Jiang X-T},
\textbf{Sheng H-F}, \textbf{Huse SM}, \textbf{Rideout JR}, \textbf{Edgar
RC}, \textbf{Kopylova E}, \textbf{Walters WA}, \textbf{Knight R},
\textbf{Zhou H-W}. 2015. Stability of operational taxonomic units: An
important but neglected property for analyzing microbial diversity.
Microbiome \textbf{3}.
doi:\href{https://doi.org/10.1186/s40168-015-0081-x}{10.1186/s40168-015-0081-x}.

\hypertarget{ref-Kopylova2016}{}
33. \textbf{Kopylova E}, \textbf{Navas-Molina JA}, \textbf{Mercier C},
\textbf{Xu ZZ}, \textbf{Mahé F}, \textbf{He Y}, \textbf{Zhou H-W},
\textbf{Rognes T}, \textbf{Caporaso JG}, \textbf{Knight R}. 2016.
Open-source sequence clustering methods improve the state of the art.
mSystems \textbf{1}:e00003--15.
doi:\href{https://doi.org/10.1128/msystems.00003-15}{10.1128/msystems.00003-15}.

\hypertarget{ref-Schmidt2014}{}
34. \textbf{Schmidt TSB}, \textbf{Rodrigues JFM}, \textbf{Mering C von}.
2014. Limits to robustness and reproducibility in the demarcation of
operational taxonomic units. Environ Microbiol \textbf{17}:1689--1706.
doi:\href{https://doi.org/10.1111/1462-2920.12610}{10.1111/1462-2920.12610}.

\hypertarget{ref-Schloss2016a}{}
35. \textbf{Schloss PD}. 2016. Application of a database-independent
approach to assess the quality of operational taxonomic unit picking
methods. mSystems \textbf{1}:e00027--16.
doi:\href{https://doi.org/10.1128/msystems.00027-16}{10.1128/msystems.00027-16}.

\hypertarget{ref-Westcott2015}{}
36. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
methods outperform reference-based methods for assigning 16S rRNA gene
sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.

\hypertarget{ref-Matthews1975}{}
37. \textbf{Matthews B}. 1975. Comparison of the predicted and observed
secondary structure of t4 phage lysozyme. Biochimica et Biophysica Acta
(BBA) - Protein Structure \textbf{405}:442--451.
doi:\href{https://doi.org/10.1016/0005-2795(75)90109-9}{10.1016/0005-2795(75)90109-9}.

\hypertarget{ref-Pruesse2007}{}
38. \textbf{Pruesse E}, \textbf{Quast C}, \textbf{Knittel K},
\textbf{Fuchs BM}, \textbf{Ludwig W}, \textbf{Peplies J},
\textbf{Glockner FO}. 2007. SILVA: A comprehensive online resource for
quality checked and aligned ribosomal RNA sequence data compatible with
ARB. Nucleic Acids Research \textbf{35}:7188--7196.
doi:\href{https://doi.org/10.1093/nar/gkm864}{10.1093/nar/gkm864}.

\hypertarget{ref-Baxter2016}{}
39. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-Schloss2012}{}
40. \textbf{Schloss PD}, \textbf{Schubert AM}, \textbf{Zackular JP},
\textbf{Iverson KD}, \textbf{Young VB}, \textbf{Petrosino JF}. 2012.
Stabilization of the murine gut microbiome following weaning. Gut
Microbes \textbf{3}:383--393.
doi:\href{https://doi.org/10.4161/gmic.21008}{10.4161/gmic.21008}.

\hypertarget{ref-Johnston2016}{}
41. \textbf{Johnston ER}, \textbf{Rodriguez-R LM}, \textbf{Luo C},
\textbf{Yuan MM}, \textbf{Wu L}, \textbf{He Z}, \textbf{Schuur EAG},
\textbf{Luo Y}, \textbf{Tiedje JM}, \textbf{Zhou J},
\textbf{Konstantinidis KT}. 2016. Metagenomics reveals pervasive
bacterial populations and reduced community diversity across the alaska
tundra ecosystem. Front Microbiol \textbf{7}.
doi:\href{https://doi.org/10.3389/fmicb.2016.00579}{10.3389/fmicb.2016.00579}.

\hypertarget{ref-Henson2016}{}
42. \textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst JL},
\textbf{Lanclos VC}, \textbf{Webber AT}, \textbf{Thrash JC}. 2016.
Artificial seawater media facilitate cultivating members of the
microbial majority from the gulf of mexico. mSphere
\textbf{1}:e00028--16.
doi:\href{https://doi.org/10.1128/msphere.00028-16}{10.1128/msphere.00028-16}.

\hypertarget{ref-Schloss2010}{}
43. \textbf{Schloss PD}. 2010. The effects of alignment quality,
distance calculation method, sequence filtering, and region on the
analysis of 16S rRNA gene-based studies. PLOS Comput Biol
\textbf{6}:e1000844.
doi:\href{https://doi.org/10.1371/journal.pcbi.1000844}{10.1371/journal.pcbi.1000844}.

\hypertarget{ref-language2015}{}
44. \textbf{R Core Team}. 2015. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

\hypertarget{ref-wesanderson}{}
45. \textbf{Ram K}, \textbf{Wickham H}. 2015. wesanderson: A wes
anderson palette generator.

\hypertarget{ref-dplyr}{}
46. \textbf{Wickham H}, \textbf{Francois R}. 2016. dplyr: A grammar of
data manipulation.

\hypertarget{ref-tidyr}{}
47. \textbf{Wickham H}. 2016. tidyr: Easily tidy data with `spread()`
and `gather()` functions.

\hypertarget{ref-cowplot}{}
48. \textbf{Wilke CO}. cowplot: Streamlined plot theme and plot
annotations for 'ggplot2'.

\hypertarget{ref-ggplot2}{}
49. \textbf{Wickham H}. 2009. ggplot2: Elegant graphics for data
analysis. Springer-Verlag New York.

\end{document}

### 2017-02-02 submission/Westcott_OptiClust_mSphere_2016_orig.tex
Add tex versions of the manuscript

--- 2017-02-02 submission/Westcott_OptiClust_mSphere_2016.tex
@@
 implemented in USEARCH (16) or VSEARCH (17), Sumaclust, OTUCLUST (18),
-and Swarm (19). \emph{De novo} methods are agglomerative and tend to be
-more computationally intense. It has proven difficult to know which
-method generates the best assignments. A third category of algorithm is
-open reference clustering, which is a hybrid approach (3, 14). Here
-sequences are assigned to OTUs using closed-reference clustering and
-sequences that are not within a threshold of a reference sequence are
-then clustered using a \emph{de novo} approach. This category blends the
-strengths and weaknesses of the other method and adds the complication
-that closed-reference and \emph{de novo} clustering use different OTU
+and Swarm (19). \emph{De novo} methods tend to be more computationally
+intense and it has proven difficult to know which method generates the
+best assignments. A third category of algorithm is open-reference
+clustering, which is a hybrid approach (3, 14). Here sequences are
+assigned to OTUs using closed-reference clustering and sequences that
+are not within a threshold of a reference sequence are then clustered
+using a \emph{de novo} approach. This category blends the strengths and
+weaknesses of the other method and adds the complication that
+closed-reference and \emph{de novo} clustering use different OTU
 definitions. These algorithms take different approaches to handling
@@
 These metrics also tend to be subjective. For instance, a method may
-appear to recapitulate the taxonomy of a synthetic community with known
-taxonomic structure, but do a poor job when applied to real communities
-with poorly defined taxonomic structure or for sequences that are prone
-to misclassification. As an alternative, we developed an approach to
-objectively benchmark the clustering quality of OTU assignments (13, 35,
-36). This approach counts the number of true positives (TP), true
-negatives (TN), false positives (FP), and false negatives (FN) based on
-the pairwise distances. Sequence pairs that are within the
-user-specified threshold and are clustered together represent TPs and
-those in different OTUs are FNs. Those sequence pairs that have a
-distance larger than the threshold and are not clustered in the same OTU
-are TNs and those in the same OTU are FPs. These values can be
+appear to be recapitulate the taxonomy of a synthetic community with
+known taxonomic structure, but do a poor job when applied to real
+communities with poorly defined taxonomic structure or for sequences
+that are prone to misclassification. As an alternative, we developed an
+approach to objectively benchmark the clustering quality of OTU
+assignments (13, 35, 36). This approach counts the number of true
+positives (TP), true negatives (TN), false positives (FP), and false
+negatives (FN) based on the pairwise distances. Sequence pairs that are
+within the user-specified threshold and are clustered together represent
+TPs and those in different OTUs are FNs. Those sequence pairs that have
+a distance larger than the threshold and are not clustered in the same
+OTU are TNs and those in the same OTU are FPs. These values can be
 synthesized into a single correlation coefficient, the Matthews'
@@
 Consistently, the average neighbor algorithm was identified as among the
-best or the best algorithm. Other hierarchical algorithms such as
-furthest and nearest neighbor, which seek to minimize the number of FP
-and FN, respectively, fared significantly worse. The distance-based
-greedy clustering as implemented in VSEARCH has also performed well. The
-computational resources required to complete the average neighbor
-algorithm can be significant for large datasets and so there is a need
-for an algorithm that efficiently produces consistently high quality OTU
-assignments.
+best or the best algorithm. The distance-based greedy clustering as
+implemented in VSEARCH has also performed well. The computational
+resources required to complete the average neighbor algorithm can be
+significant for large datasets and so there is a need for an algorithm
+that efficiently produces consistently high quality OTU assignments.
 
@@
 consists of interrogating every sequence in the dataset. Although
-numerous options are available for optimizing the clusters and for
-assessing the quality of the clusters within the mothur-based
-implementation of the algorithm (e.g.~sensitivity, specificity,
-accuracy, F1 score, etc.), the default metric for optimization and
-assessment is MCC because it includes all four parameters from the
-confusion matrix (Figure S1). The algorithm continues until the
-optimization metric stabilizes or until it reaches a defined stopping
-criteria (Figure S2).
+numerous options are available within the mothur-based implementation of
+the algorithm (e.g.~sensitivity, specificity, accuracy, F1 score, etc.),
+the default metric is MCC because it includes all four parameters from
+the confusion matrix. The algorithm continues until the optimization
+metric stabilizes or until it reaches a defined stopping criteria.
 
@@
 algorithms is that their output is sensitive to the initial order of the
-sequences because each algorithm must break ties where a sequence could
-be assigned to multiple OTUs. An additional concern with the OptiClust
-algorithm is that it may stabilize at a local optimum. To evaluate these
-concerns we compared the results obtained using ten randomizations of
-the order that sequences were given to the algorithm. The median the
-coefficient of variation across the six datasets for MCC values obtained
-from the replicate clusterings using OptiClust was 0.1\% (Figure 1). We
-also measured the coefficient of variation for the number of OTUs across
-the six datasets for each method. The median coefficient of variation
-for the number of OTUs generated using OptiClust was 0.1\%. Confirming
-our previous results, all of the methods we tested were stable to
-stochastic processes. Of the methods that involved randomization, the
-coefficient of variation for MCC values was considerably smaller with
-OptiClust than the other methods and the coefficient of variation for
-the number of OTUs was comparable to the other methods. The variation
-observed in clustering quality suggested that the algorithm does not
-appear to converge to a locally optimum MCC value. More importantly, the
-random variation does yield output of a similarly high quality.
+sequences. An additional concern with the OptiClust algorithm is that it
+may stabilize at a local optimum. To evaluate these concerns we compared
+the results obtained using ten randomizations of the order that
+sequences were given to the algorithm. The median the coefficient of
+variation across the six datasets for MCC values obtained from the
+replicate clusterings using OptiClust was 0.1\% (Figure 1). We also
+measured the coefficient of variation for the number of OTUs across the
+six datasets for each method. The median coefficient of variation for
+the number of OTUs generated using OptiClust was 0.1\%. Confirming our
+previous results, all of the methods we tested were stable to stochastic
+processes. Of the methods that involved randomization, the coefficient
+of variation for MCC values considerably smaller with OptiClust than the
+other methods and the coefficient of variation for the number of OTUs
+was comparable to the other methods. The variation observed in
+clustering quality suggested that the algorithm does not appear to
+converge to a locally optimum MCC value. More importantly, the random
+variation does yield output of a similarly high quality.
 
@@
 quality of clustering, the amount of time and memory required to assign
-sequences to OTUs is a legitimate concern. We observed that the time
-required to complete the OptiClust algorithm (Figure 1C) paralleled the
-number of pairwise distances that were smaller than 0.03. To further
-evaluate how the speed and memory usage scaled with the number of
-sequences in the dataset, we measured the time required and maximum RAM
-usage to cluster 20, 40, 60, 80, and 100\% of the unique sequences from
-each of the natural datasets using the OptiClust algorithm (Figure 2).
-Within each iteration of the algorithm, each sequence is compared to
-every other sequence and each comparison requires a recalculation of the
-confusion matrix. This would result in a worst case algorithmic
-complexity on the order of N\^{}3, where N is the number of unique
-sequences. Because the algorithm only needs to keep track of the
-sequence pairs that are within the threshold of each other, it is likely
-that the implementation of the algorithm is more efficient. To
-empirically determine the algorithmic complexity, we fit a power law
-function to the data in Figure 2A. We observed power coefficients
-between 1.7 and 2.5 for the marine and human datasets, respectively. The
-algorithm requires storing a matrix that contains the pairs of sequences
-that are close to each other as well as a matrix that indicates which
-sequences are clustered together. The memory required to store these
-matrices is on the order of N\^{}2, where N is the number of unique
-sequences. In fact, when we fit a power law function to the data in
-Figure 2B, the power coefficients were 1.9. Using the four natural
-community datasets, doubling the number of sequences in a dataset would
-increase the time required to cluster the data by 4 to 8-fold and
-increase the RAM required by 4-fold. It is possible that future
-improvements to the implementation of the algorithm could improve this
-performance.
+sequences to OTUs is a legitimate concern. To evaluate how the speed and
+memory usage scaled with the number of sequences in the dataset, we
+measured the time required and maximum RAM usage to cluster 20, 40, 60,
+80, and 100\% of the unique sequences from each of the natural datasets
+using the OptiClust algorithm (Figure 2). Within each iteration of the
+algorithm, each sequence is compared to every other sequence and each
+comparison requires a recalculation of the confusion matrix. This would
+result in a worst case algorithmic complexity on the order of N\^{}3,
+where N is the number of unique sequences. Because the algorithm only
+needs to keep track of the sequence pairs that are within the threshold
+of each other, it is likely that the implementation of the algorithm is
+more efficient. To empirically determine the algorithmic complexity, we
+fit a power law function to the data in Figure 2A. We observed power
+coefficients between 1.7 and 2.5 for the marine and human datasets,
+respectively. The algorithm requires storing a matrix that contains the
+pairs of sequences that are close to each other as well as a matrix that
+indicates which sequences are clustered together. The memory required to
+store these matrices is on the order of N\^{}2, where N is the number of
+unique sequences. In fact, when we fit a power law function to the data
+in Figure 2B, the power coefficients were 1.9. This analysis suggests
+that doubling the number of sequences in a dataset would increase the
+time required to cluster the data by 4 to 8-fold and increase the RAM
+required by 4-fold. It is possible that future improvements to the
+implementation of the algorithm could improve this performance.
 
@@
 taxonomic groups and within each taxon sequences were assigned to OTUs
-using the average neighbor clustering algorithm (13). This method is
-similar to open reference clustering except that in our approach all
-sequences are subjected to \emph{de novo} clustering following
-classification whereas in open reference clustering only those sequences
-that cannot be classified are subjected to \emph{de novo} clustering.
-Our cluster splitting approach accelerated the clustering and reduced
-the memory requirements because the number of unique sequences is
-effectively reduced by splitting sequences across taxonomic groups.
-Furthermore, because sequences in different taxonomic groups are assumed
-to belong to different OTUs they are independent, which permits
-parallelization and additional reduction in computation time. Reduction
-in clustering quality are encountered in this approach if there are
-errors in classification or if two sequences within the desired
-threshold belong to different taxonomic groups. It is expected that
-these errors would increase as the taxonomic level goes from kingdom to
-genus. To characterize the clustering quality, we classified each
-sequence at each taxonomic level and calculated the MCC values using
-OptiClust, average neighbor, and DGC with VSEARCH when splitting at each
-taxonomic level (Figure 3). For each method, the MCC values decreased as
-the taxonomic resolution increased; however, the decrease in MCC as not
-as large as the difference between clustering methods. As the resolution
-of the taxonomic levels increased, the clustering quality remained high,
-relative to clusters formed from the entire dataset
-(i.e.~kingdom-level). The MCC values when splitting the datasets at the
-class and genus levels were within 98.0 and 93.0\%, respectively, of the
-MCC values obtained from the entire dataset. These decreases in MCC
-value resulted in the formation of as many as 4.7 and 22.5\% more OTUs,
-respectively, than were observed from the entire dataset. These errors
-were due to the generation of additional false negatives due to
-splitting similar sequences into different taxonomic groups. For the
-datasets included in the current analysis, the use of the cluster
-splitting heuristic was probably not worth the loss in clustering
-quality. However, as datasets become larger, it may be necessary to use
-the heuristic to clustering the data into OTUs.
+using the average neighbor clustering algorithm (13). This accelerated
+the clustering and reduce the memory requirements because the number of
+unique sequences is effectively reduced by splitting sequences across
+taxonomic groups. Furthermore, because sequences in different taxonomic
+groups are assumed to belong to different OTUs they are independent,
+which permits parallelization and additional reduction in computation
+time. Reduction in clustering quality are encountered in this approach
+if there are errors in classification or if two sequences within the
+desired threshold belong to different taxonomic groups. It is expected
+that these errors would increase as the taxonomic level goes from
+kingdom to genus. To characterize the clustering quality, we calculated
+the MCC values using OptiClust, average neighbor, and DGC with VSEARCH
+when splitting at each taxonomic level (Figure 3). For each method, the
+MCC values decreased as the taxonomic resolution increased; however, the
+decrease in MCC as not as large as the difference between clustering
+methods. As the resolution of the taxonomic levels increased, the
+clustering quality remained high, relative to clusters formed from the
+entire dataset (i.e.~kingdom-level). The MCC values when splitting the
+datasets at the class and genus levels were within 98.0 and 93.0\%,
+respectively, of the MCC values obtained from the entire dataset. These
+decreases in MCC value resulted in the formation of as many as 4.7 and
+22.5\% more OTUs, respectively, than were observed from the entire
+dataset. For the datasets included in the current analysis, the use of
+the cluster splitting heuristic was probably not worth the loss in
+clustering quality. However, as datasets become larger, it may be
+necessary to use the heuristic to clustering the data into OTUs.
 
@@
 FP and FN, respectively; however, these suffer because the number of FN
-and FP are not controlled, respectively (13, 15). Alternatively, one
-could optimize based on the sensitivity, specificity, or accuracy, which
-are each based on two values from the confusion matrix or they could
-optimize based on the F1 score, which is based on three values from the
-confusion matrix. Because these metrics do not balance all four
-parameters equally, it is likely that one parameter will dominate in the
-optimization procedure. For example, optimizing for sensitivity could
-lead to a large number of FPs. More FPs increases the number of OTUs
-while more FNs collapses OTUs together. It is difficult to know which is
-worse since community richness and diversity are linked to the number of
-OTUs. In addition, increasing the number of FNs would overstate the
-differences between communities while increasing the number of FPs would
-overstate their similarity. Therefore, it is important to jointly
-minimize the number of FPs and FNs. With this in mind, we decided to
-optimize utilizing the MCC. It is possible that other metrics that
-balance the four parameters could be developed and employed for
-optimization of the clustering.
+and FP are not controlled (13, 15). Alternatively, one could optimize
+based on the sensitivity, specificity, or accuracy, which are each based
+on two values from the confusion matrix or they could optimize based on
+the F1 score, which is based on three values from the confusion matrix.
+Because these metrics do not balance all four parameters equally, it is
+likely that one parameter will dominate in the optimization procedure.
+For example, optimizing for sensitivity could lead to a large number of
+FPs. Since we would like to minimize both FPs and FNs and not just the
+total number of false assignments, we decided to optimize utilizing the
+MCC. It is possible that other metrics could be developed and employed
+for optimization of the clustering.
 
 The OptiClust algorithm is relatively simple. For each sequence it
-effectively asks whether the MCC value will increase if the sequence is
+effectively ask whether the MCC value will increase if the sequence is
 moved to a different OTU including creating a new OTU. If the value does
@@
 the algorithm could provide optimization to further improve its speed
-and susceptibility to find a local optimum. Users are encouraged to
+and susceptibility to find a local optimum. Users are encourage to
 repeat the OTU assignment several times to confirm that they have found
@@
 algorithm and compare its performance to other algorithms.} Each dataset
-contains sequences from the V4 region of the 16S rRNA gene. The number
-of distances for each dataset are those that were less than or equal to
-0.03. The number of OTUs were determined using the OptiClust algorithm.
-The even and staggered datasets were generated by extracting the V4
-region from full length reference sequences and the datasets from the
-natural communities were generated by sequencing the V4 region using a
-Illumina MiSeq with either paired 150 or 250 nt reads.
-
-\begin{longtable}[c]{@{}lcccccc@{}}
+contains sequences from the V4 region of the 16S rRNA gene. The even and
+staggered datasets were generated by extracting the V4 region from full
+length reference sequences and the datasets from the natural communities
+were generated by sequencing the V4 region using a Illumina MiSeq with
+either paired 150 or 250 nt reads.
+
+\begin{longtable}[c]{@{}lcccc@{}}
 \toprule
 \textbf{Dataset (Ref.)} & \textbf{Read Length} & \textbf{Samples} &
-\textbf{Total Seqs.} & \textbf{Unique Seqs.} & \textbf{Distances} &
-\textbf{OTUs}\tabularnewline
+\textbf{Total Seqs.} & \textbf{Unique Seqs.}\tabularnewline
 \midrule
 \endhead
-Soil (41) & 150 & 18 & 948,243 & 143,677 & 11,775,167 &
-40,216\tabularnewline
-Marine (42) & 250 & 7 & 1,384,988 & 75,923 & 12,908,857 &
-25,787\tabularnewline
-Mice (40) & 250 & 360 & 2,825,495 & 32,447 & 6,988,306 &
-2,658\tabularnewline
-Human (39) & 250 & 489 & 20,951,841 & 121,281 & 38,544,315 &
-11,648\tabularnewline
-Even (33, 35) & NA & NA & 1,155,800 & 11,558 & 29,694 &
-7,651\tabularnewline
-Staggered (33, 35) & NA & NA & 1,156,550 & 11,558 & 29,694 &
-7,653\tabularnewline
+Soil (41) & 150 & 18 & 948,243 & 143,677\tabularnewline
+Marine (42) & 250 & 7 & 1,384,988 & 75,923\tabularnewline
+Mice (40) & 250 & 360 & 2,825,495 & 32,447\tabularnewline
+Human (39) & 250 & 489 & 20,951,841 & 121,281\tabularnewline
+Even (33, 35) & NA & NA & 1,155,800 & 11,558\tabularnewline
+Staggered (33, 35) & NA & NA & 1,156,550 & 11,558\tabularnewline
 \bottomrule
@@
 
-\includegraphics[width=6.0in]{../results/figures/performance_other.png}
-
-\textbf{Figure S1. The OptiClust algorithm is able to effectively
-cluster sequences into OTUs by minimizing or maximizing numerous
-metrics.} Plot of MCC (A), number of OTUs (B), and execution times (C)
-for the comparison of output from the OptiClust algorithm when to
-minimizing or maximizing a variety of parameters when applied to four
-natural and two synthetic datasets. Within mothur, OTU assignments can
-also be made using other metrics including minimizing false positives
-and maximizing the specificity, positive predictive value, and true
-negatives; however, these all resulted in sequences being assigned to
-separate OTUs, which resulted in no false positives and the maximum
-number of true negatives. The error bars indicate the range of values
-observed for 10 replicates.
-
-\newpage
-
-\includegraphics[width=3.5in]{../results/figures/optimization.png}
-
-\textbf{Figure S2. The OptiClust algorithm rapidly converges to optimize
-the Matthew's Correlation Coefficient.} The six datasets were clustered
-into OTUs using the OptiClust algorithm seeking to maximize the
-Matthew's Correlation Coefficient. This was repeated 10 times for each
-dataset.
-
-\newpage
-
 \textbf{Supplemental text.} Worked example of how OptiClust algorithm

### 2017-02-03 submission/Westcott_OptiClust_mSphere_2016.Rmd
Editing

@@
 Sequences are compared to a reference collection and clustered based on the reference sequences that they are similar to.
-This approach is fast; however, the method struggles when a sequence is similar to multiple reference sequences that may have different taxonomies and when it is not similar to sequences in the reference.
+This approach is fast; however, the method struggles when a sequence is similar to multiple reference sequences that may have different taxonomies and when it is not similar to sequences in the reference [@Westcott2015].
 The second category of algorithms has been called *de novo* because they assign sequences to OTUs without the use of a reference [@NavasMolina2013].
@@
 This category blends the strengths and weaknesses of the other method and adds the complication that closed-reference and *de novo* clustering use different OTU definitions.
-These algorithms take different approaches to handling large datasets to minimize the time and memory requirements while attempting to assign sequences to meaningful OTUs.
+These three categories of algorithms take different approaches to handling large datasets to minimize the time and memory requirements while attempting to assign sequences to meaningful OTUs.
 
@@
 Those sequence pairs that have a distance larger than the threshold and are not clustered in the same OTU are TNs and those in the same OTU are FPs.
-These values can be synthesized into a single correlation coefficient, the Matthews' Correlation Coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix [@Matthews1975].
+These values can be synthesized into a single correlation coefficient, the Matthews correlation coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix [@Matthews1975].
 Consistently, the average neighbor algorithm was identified as among the best or the best algorithm.
-Other hierarchical algorithms such as furthest and nearest neighbor, which seek to minimize the number of FP and FN, respectively, fared significantly worse.
+Other hierarchical algorithms such as furthest and nearest neighbor, which do not permit the formation of FPs or FNs, respectively, fared significantly worse.
 The distance-based greedy clustering as implemented in VSEARCH has also performed well.
@@
 
-These previous efforts have assessed the quality of the clusters after the completion of the algorithm.
+These benchmarking efforts have assessed the quality of the clusters after the completion of the algorithm.
 In the current study we developed and benchmarked a new *de novo* clustering algorithm that uses real time calculation of the MCC to direct the progress of the clustering.
@@
 Although numerous options are available for optimizing the clusters and for assessing the quality of the clusters within the mothur-based implementation of the algorithm (e.g.
-sensitivity, specificity, accuracy, F1 score, etc.), the default metric for optimization and assessment is MCC because it includes all four parameters from the confusion matrix (Figure S1).
-The algorithm continues until the optimization metric stabilizes or until it reaches a defined stopping criteria (Figure S2).
+sensitivity, specificity, accuracy, F1-score, etc.), the default metric for optimization and assessment is MCC because it includes all four parameters from the confusion matrix (Figure S1).
+The algorithm continues until the optimization metric stabilizes or until it reaches a defined stopping criteria.
 
@@
 When we seeded the OptiClust algorithm with each sequence in a separate OTU and ran the algorithm until complete convergence, the MCC values averaged `r round(mcc_rel_an, 1)` and `r round(mcc_rel_vdgc, 1)`% higher than the OTUs using average neighbor and distance-based greedy clustering (DGC) with VSEARCH, respectively (Figure 1).
-The number of OTUs formed by the various methods was negatively correlated with their MCC value ($\rho$=`r round(mcc_sobs_cor$estimate, 2)`; p=`r round(mcc_sobs_cor$p.value, 2)`).
+The number of OTUs formed by the various methods was negatively correlated with their MCC value ($\rho$=`r round(mcc_sobs_cor$estimate, 2)`; p`r p<-mcc_sobs_cor$p.value; ifelse(p<0.001, "<0.001", paste0("=", p))`).
 The OptiClust algorithm was considerably faster than the hierarchical algorithms and somewhat slower than the heuristic-based algorithms.
@@
 By default mothur uses a maximum value of 100 iterations.
-The justification for allowing incomplete convergence was based on the observation that numerous iterations are performed that extend the time required to complete the clustering with minimal improvement in clustering.
+The justification for allowing incomplete convergence was based on the observation that numerous iterations are performed that extend the time required to complete the clustering with minimal improvement in clustering (Figure S2).
 We evaluated the results of clustering to partial convergence (i.e.
@@
 This represented a difference of less than `r format(round(max(steps$per_diff_numotus), 2), nsmall=1L)`%.
-Among the four natural datasets, between `r min(steps$med_close_steps)` and `r max(steps$med_close_steps)` were needed to achieve partial convergence and between `r min(steps$med_full_steps)` and `r max(steps$med_full_steps)` iterations were needed to reach full convergence.
+Among the four natural datasets, between `r min(steps$med_close_steps)` and `r max(steps$med_close_steps)` were needed to achieve partial convergence and between `r min(steps$med_full_steps)` and `r round(max(steps$med_full_steps))` iterations were needed to reach full convergence.
 The additional steps required between `r paste(format(round(range(steps$ratio_secs), 1), nsmall=1L), collapse=" and ")` times longer to complete the algorithm.
@@
 One concern that many have with *de novo* clustering algorithms is that their output is sensitive to the initial order of the sequences because each algorithm must break ties where a sequence could be assigned to multiple OTUs.
-An additional concern with the OptiClust algorithm is that it may stabilize at a local optimum.
+An additional concern specific to the OptiClust algorithm is that it may stabilize at a local optimum.
 To evaluate these concerns we compared the results obtained using ten randomizations of the order that sequences were given to the algorithm.
-The median the coefficient of variation across the six datasets for MCC values obtained from the replicate clusterings using OptiClust was `r format(round(stability %>% filter(method=="mcc") %>% select(med_cov_mcc), 1), nsmall=1L)`% (Figure 1).
+The median coefficient of variation across the six datasets for MCC values obtained from the replicate clusterings using OptiClust was `r format(round(stability %>% filter(method=="mcc") %>% select(med_cov_mcc), 1), nsmall=1L)`% (Figure 1).
 We also measured the coefficient of variation for the number of OTUs across the six datasets for each method.
 The median coefficient of variation for the number of OTUs generated using OptiClust was `r format(round(stability %>% filter(method=="mcc") %>% select(med_cov_num_otus), 1), nsmall=1L)`%.
-Confirming our previous results, all of the methods we tested were stable to stochastic processes.
+Confirming our previous results [@Westcott2015], all of the methods we tested were stable to stochastic processes.
 Of the methods that involved randomization, the coefficient of variation for MCC values was considerably smaller with OptiClust than the other methods and the coefficient of variation for the number of OTUs was comparable to the other methods.
@@
 Although not as important as the quality of clustering, the amount of time and memory required to assign sequences to OTUs is a legitimate concern.
-We observed that the time required to complete the OptiClust algorithm (Figure 1C) paralleled the number of pairwise distances that were smaller than 0.03.
+We observed that the time required to complete the OptiClust algorithm (Figure 1C) paralleled the number of pairwise distances that were smaller than 0.03 (Table 1).
 To further evaluate how the speed and memory usage scaled with the number of sequences in the dataset, we measured the time required and maximum RAM usage to cluster 20, 40, 60, 80, and 100% of the unique sequences from each of the natural datasets using the OptiClust algorithm (Figure 2).
 Within each iteration of the algorithm, each sequence is compared to every other sequence and each comparison requires a recalculation of the confusion matrix.
-This would result in a worst case algorithmic complexity on the order of N^3, where N is the number of unique sequences.
+This would result in a worst case algorithmic complexity on the order of N^3^, where N is the number of unique sequences.
 Because the algorithm only needs to keep track of the sequence pairs that are within the threshold of each other, it is likely that the implementation of the algorithm is more efficient.
@@
 The algorithm requires storing a matrix that contains the pairs of sequences that are close to each other as well as a matrix that indicates which sequences are clustered together.
-The memory required to store these matrices is on the order of N^2, where N is the number of unique sequences.
+The memory required to store these matrices is on the order of N^2^, where N is the number of unique sequences.
 In fact, when we fit a power law function to the data in Figure 2B, the power coefficients were  `r format(round(median(memory_coef), 1), nsmall=1L)`.
@@
 This method is similar to open reference clustering except that in our approach all sequences are subjected to *de novo* clustering following classification whereas in open reference clustering only those sequences that cannot be classified are subjected to *de novo* clustering.
-Our cluster splitting approach accelerated the clustering and reduced the memory requirements because the number of unique sequences is effectively reduced by splitting sequences across taxonomic groups.
+Our cluster splitting approach accelerated the clustering and reduced the memory requirements because the number of unique sequences was effectively reduced by splitting sequences across taxonomic groups.
 Furthermore, because sequences in different taxonomic groups are assumed to belong to different OTUs they are independent, which permits parallelization and additional reduction in computation time.
-Reduction in clustering quality are encountered in this approach if there are errors in classification or if two sequences within the desired threshold belong to different taxonomic groups.
+Reduction in clustering quality is encountered in this approach if there are errors in classification or if two sequences within the desired threshold belong to different taxonomic groups.
 It is expected that these errors would increase as the taxonomic level goes from kingdom to genus.
 To characterize the clustering quality, we classified each sequence at each taxonomic level and calculated the MCC values using OptiClust, average neighbor, and DGC with VSEARCH when splitting at each taxonomic level (Figure 3).
-For each method, the MCC values decreased as the taxonomic resolution increased; however, the decrease in MCC as not as large as the difference between clustering methods.
+For each method, the MCC values decreased as the taxonomic resolution increased; however, the decrease in MCC was not as large as the difference between clustering methods.
 As the resolution of the taxonomic levels increased, the clustering quality remained high, relative to clusters formed from the entire dataset (i.e.
@@
 ## Discussion
-Myriad methods have been proposed for assigning 16S rRNA gene sequences to OTUs that each claim improved performance based on speed, memory usage, representation of taxonomic information, and number of OTUs.
+Myriad methods have been proposed for assigning 16S rRNA gene sequences to OTUs.
+Each claim improved performance based on speed, memory usage, representation of taxonomic information, and number of OTUs.
 Each of these metrics is subjective and do not actually indicate the quality of the clustering.
@@
 Here, we described a new clustering method that seeks to optimize clustering based on an objective criterion that measures clustering quality in real time.
-In the OptiClust algorithm clustering is driven by optimizing a metric that assesses whether any two sequences should be grouped into the same OTU.
+In the OptiClust algorithm, clustering is driven by optimizing a metric that assesses whether any two sequences should be grouped into the same OTU.
 The result is clusters that are significantly more robust and is efficient in the time and memory required to cluster the sequences into OTUs.
@@
 Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled, respectively [@Schloss2005; @Schloss2011].
-Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1 score, which is based on three values from the confusion matrix.
+Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1-score, which is based on three values from the confusion matrix.
 Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure.
@@
 With the exception of Swarm each of these methods uses distance-based thresholds to report OTU assignments.
+We also evalauted the ability of OptiClust to optimize to metrics other than MCC.
+These included accuracy, F1-score, negative predictive value, positive predictive value, false discovery rate, senitivity, specificity, the sum of TPs and TNs, the sum of FPs and FNs, and the number of FNs, FPs, TNs, and TPs (Figure S1).
 
@@
 Pairs of sequences that were in the same OTU and had a distance less than 3% were true positives (TPs), those that were in different OTUs and had a distance greater than 3% were true negatives (TNs), those that were in the same OTU and had a distance greater than 3% were false positives (FPs), and those that were in different OTUs and had a distance less than 3% were false negatives (FNs).
-To synthesize the matrix into a single metric we used the Matthews Correlation Coefficient using the `sens.spec` command in mothur using the following equations.
+To synthesize the matrix into a single metric we used the Matthews correlation coefficient using the `sens.spec` command in mothur using the following equations.
 
@@
 We obtained 10 subsamples at each depth for each dataset and ran each collection (N= 50 = 5 sequencing depths x 10 replicates) through each of the algorithms.
-We used the timeout script to quantify the maximum RAM used and the amount of time required to process each sequence collection (https://github.com/pshved/timeout).
+We used the `timeout` script to quantify the maximum RAM used and the amount of time required to process each sequence collection (https://github.com/pshved/timeout).
 We limited each algorithm to 45 GB of RAM and 50 hours using a single processor.
@@
 **Figure S2.
-The OptiClust algorithm rapidly converges to optimize the Matthew's Correlation Coefficient.** The six datasets were clustered into OTUs using the OptiClust algorithm seeking to maximize the Matthew's Correlation Coefficient.
+The OptiClust algorithm rapidly converges to optimize the Matthews correlation coefficient.** The six datasets were clustered into OTUs using the OptiClust algorithm seeking to maximize the Matthews correlation coefficient.
 This was repeated 10 times for each dataset.

### 2017-02-03 code/build_optimization_figure.R

@@ -55,7 +55,7 @@ all_step_data %>%
 	geom_line() +
 	scale_color_manual(breaks=levels(all_step_data$dataset), labels=pretty_datasets[levels(all_step_data$dataset)],
 											values=c('black', wes_palette("Darjeeling")), name=NULL) +
-	xlab("Number of Iterations") + ylab("Matthew's\nCorrelation Coefficient") +
+	xlab("Number of Iterations") + ylab("Matthews\nCorrelation Coefficient") +
 	expand_limits(y=c(0.7,0.9), x=c(0,17)) +
 	scale_x_continuous(breaks=c(0,5,10,15,20), labels=c(0,5,10,15,20)) +
 	my_theme +

### 2017-02-03 code/build_performance_figure.R

@@ -18,7 +18,7 @@ pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
 names(pretty_datasets) <- datasets
 
 metrics <- c("mcc", "num_otus", "cluster_secs")
-pretty_metrics <- c("Matthew's\nCorrelation Coefficient", "Number of OTUs", "Time to Cluster\n(seconds)")
+pretty_metrics <- c("Matthews\nCorrelation Coefficient", "Number of OTUs", "Time to Cluster\n(seconds)")
 names(pretty_metrics) <- metrics
 
 

### 2017-02-03 code/build_performance_other_figure.R

@@ -11,7 +11,7 @@ basic_methods <- c('accuracy', 'f1score', 'fpfn', 'mcc', 'npv', 'opti_fn', 'sens
 
 pretty_methods <- c('Accuracy (max.)', 'F1-score (max.)',
 										'False Positives +\nFalse Negatives (min.)',
-										"Matthew's Correlation\nCoefficient (max.)",
+										"Matthews Correlation\nCoefficient (max.)",
 										"Negative Predictive\nValue (max.)", "False Negatives (min.)", "Sensitivity (max.)",
  										"True Positives (max.)", "True Positives +\nTrue Negatives (max.)")
 
@@ -22,7 +22,7 @@ pretty_datasets <- c("Soil", "Marine", "Mice", "Human", "Even", "Staggered")
 names(pretty_datasets) <- datasets
 
 metrics <- c("mcc", "num_otus", "cluster_secs")
-pretty_metrics <- c("Matthew's\nCorrelation Coefficient", "Number of OTUs", "Time to Cluster\n(seconds)")
+pretty_metrics <- c("Matthews\nCorrelation Coefficient", "Number of OTUs", "Time to Cluster\n(seconds)")
 names(pretty_metrics) <- metrics
 
 

### 2017-02-03 submission/Westcott_OptiClust_mSphere_2016.tex
Re render with edits

@@
     \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
   \newcommand{\euro}{€}
@@
 \usepackage[margin=1.0in]{geometry}
-\usepackage{hyperref}
-\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
-\hypersetup{unicode=true,
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
             pdftitle={OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units},
-            pdfborder={0 0 0},
-            breaklinks=true}
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
 \urlstyle{same}  % don't use monospace font for urls
@@
 that may have different taxonomies and when it is not similar to
-sequences in the reference. The second category of algorithms has been
-called \emph{de novo} because they assign sequences to OTUs without the
-use of a reference (14). These include hierarchical algorithms such as
-nearest, furthest, and average neighbor (15) and algorithms that employ
-heuristics such as abundance or distance-based greedy clustering as
-implemented in USEARCH (16) or VSEARCH (17), Sumaclust, OTUCLUST (18),
-and Swarm (19). \emph{De novo} methods are agglomerative and tend to be
-more computationally intense. It has proven difficult to know which
-method generates the best assignments. A third category of algorithm is
-open reference clustering, which is a hybrid approach (3, 14). Here
-sequences are assigned to OTUs using closed-reference clustering and
-sequences that are not within a threshold of a reference sequence are
-then clustered using a \emph{de novo} approach. This category blends the
-strengths and weaknesses of the other method and adds the complication
-that closed-reference and \emph{de novo} clustering use different OTU
-definitions. These algorithms take different approaches to handling
-large datasets to minimize the time and memory requirements while
-attempting to assign sequences to meaningful OTUs.
+sequences in the reference (15). The second category of algorithms has
+been called \emph{de novo} because they assign sequences to OTUs without
+the use of a reference (14). These include hierarchical algorithms such
+as nearest, furthest, and average neighbor (16) and algorithms that
+employ heuristics such as abundance or distance-based greedy clustering
+as implemented in USEARCH (17) or VSEARCH (18), Sumaclust, OTUCLUST
+(19), and Swarm (20). \emph{De novo} methods are agglomerative and tend
+to be more computationally intense. It has proven difficult to know
+which method generates the best assignments. A third category of
+algorithm is open reference clustering, which is a hybrid approach (3,
+14). Here sequences are assigned to OTUs using closed-reference
+clustering and sequences that are not within a threshold of a reference
+sequence are then clustered using a \emph{de novo} approach. This
+category blends the strengths and weaknesses of the other method and
+adds the complication that closed-reference and \emph{de novo}
+clustering use different OTU definitions. These three categories of
+algorithms take different approaches to handling large datasets to
+minimize the time and memory requirements while attempting to assign
+sequences to meaningful OTUs.
 
@@
 algorithms. These have included the time and memory required to run the
-algorithm (3, 19--21), agreement between OTU assignments and the
-sequences' taxonomy (19, 21--31), sensitivity of an algorithm to
-stochastic processes (32), the number of OTUs generated by the algorithm
-(22, 33), and the ability to regenerate the assignments made by other
-algorithms (3, 34). Unfortunately, these methods fail to directly
+algorithm (3, 20--22), agreement between OTU assignments and the
+sequences' taxonomy (20, 22--32), sensitivity of an algorithm to
+stochastic processes (33), the number of OTUs generated by the algorithm
+(23, 34), and the ability to regenerate the assignments made by other
+algorithms (3, 35). Unfortunately, these methods fail to directly
 quantify the quality of the OTU assignments. An algorithm may complete
@@
 to misclassification. As an alternative, we developed an approach to
-objectively benchmark the clustering quality of OTU assignments (13, 35,
+objectively benchmark the clustering quality of OTU assignments (13, 15,
 36). This approach counts the number of true positives (TP), true
@@
 are TNs and those in the same OTU are FPs. These values can be
-synthesized into a single correlation coefficient, the Matthews'
-Correlation Coefficient (MCC), which measures the correlation between
+synthesized into a single correlation coefficient, the Matthews
+correlation coefficient (MCC), which measures the correlation between
 observed and predicted classifications and is robust to cases where
@@
 best or the best algorithm. Other hierarchical algorithms such as
-furthest and nearest neighbor, which seek to minimize the number of FP
-and FN, respectively, fared significantly worse. The distance-based
+furthest and nearest neighbor, which do not permit the formation of FPs
+or FNs, respectively, fared significantly worse. The distance-based
 greedy clustering as implemented in VSEARCH has also performed well. The
@@
 
-These previous efforts have assessed the quality of the clusters after
-the completion of the algorithm. In the current study we developed and
-benchmarked a new \emph{de novo} clustering algorithm that uses real
+These benchmarking efforts have assessed the quality of the clusters
+after the completion of the algorithm. In the current study we developed
+and benchmarked a new \emph{de novo} clustering algorithm that uses real
 time calculation of the MCC to direct the progress of the clustering.
@@
 implementation of the algorithm (e.g.~sensitivity, specificity,
-accuracy, F1 score, etc.), the default metric for optimization and
+accuracy, F1-score, etc.), the default metric for optimization and
 assessment is MCC because it includes all four parameters from the
@@
 optimization metric stabilizes or until it reaches a defined stopping
-criteria (Figure S2).
+criteria.
 
@@
 (Figure 1). The number of OTUs formed by the various methods was
-negatively correlated with their MCC value (\(\rho\)=-0.47; p=0). The
-OptiClust algorithm was considerably faster than the hierarchical
-algorithms and somewhat slower than the heuristic-based algorithms.
-Across the six datasets, the OptiClust algorithm was 94.6-times faster
-than average neighbor and just as fast as DGC with VSEARCH. The human
-dataset was a challenge for a number of the algorithms. OTUCLUST and
-SumaClust were unable to cluster the human dataset in less than 50 hours
-and the average neighbor algorithm required more than 45 GB of RAM. The
-USEARCH-based methods were unable to cluster the human data using the
-32-bit free version of the software that limits the amount of RAM to
-approximately 3.5 GB. These data demonstrate that OptiClust generated
-significantly more robust OTU assignments than existing methods across a
-diverse collection of datasets with performance that was comparable to
-popular methods.
+negatively correlated with their MCC value (\(\rho\)=-0.47;
+p\textless{}0.001). The OptiClust algorithm was considerably faster than
+the hierarchical algorithms and somewhat slower than the heuristic-based
+algorithms. Across the six datasets, the OptiClust algorithm was
+94.6-times faster than average neighbor and just as fast as DGC with
+VSEARCH. The human dataset was a challenge for a number of the
+algorithms. OTUCLUST and SumaClust were unable to cluster the human
+dataset in less than 50 hours and the average neighbor algorithm
+required more than 45 GB of RAM. The USEARCH-based methods were unable
+to cluster the human data using the 32-bit free version of the software
+that limits the amount of RAM to approximately 3.5 GB. These data
+demonstrate that OptiClust generated significantly more robust OTU
+assignments than existing methods across a diverse collection of
+datasets with performance that was comparable to popular methods.
 
@@
 iterations are performed that extend the time required to complete the
-clustering with minimal improvement in clustering. We evaluated the
-results of clustering to partial convergence (i.e.~a change in the MCC
-value that was less than 0.0001) or until complete convergence of the
-MCC value (i.e.~until it did not change between iterations) when seeding
-the algorithm with each sequence in a separate OTU (Figure 1). The small
-difference in MCC values between the output from partial and complete
-convergence resulted in a difference in the median number of OTUs that
-ranged between 1.5 and 17.0 OTUs. This represented a difference of less
-than 0.15\%. Among the four natural datasets, between 3 and 6 were
-needed to achieve partial convergence and between 8 and 12.50 iterations
-were needed to reach full convergence. The additional steps required
-between 1.4 and 1.7 times longer to complete the algorithm. These
-results suggest that achieving full convergence of the optimization
-metric adds computational effort; however, considering full convergence
-took between 2 and 17 minutes the extra effort was relatively small.
-Although the mothur's default setting is partial convergence, the
-remainder of our analysis used complete convergence to be more
-conservative.
+clustering with minimal improvement in clustering (Figure S2). We
+evaluated the results of clustering to partial convergence (i.e.~a
+change in the MCC value that was less than 0.0001) or until complete
+convergence of the MCC value (i.e.~until it did not change between
+iterations) when seeding the algorithm with each sequence in a separate
+OTU (Figure 1). The small difference in MCC values between the output
+from partial and complete convergence resulted in a difference in the
+median number of OTUs that ranged between 1.5 and 17.0 OTUs. This
+represented a difference of less than 0.15\%. Among the four natural
+datasets, between 3 and 6 were needed to achieve partial convergence and
+between 8 and 12 iterations were needed to reach full convergence. The
+additional steps required between 1.4 and 1.7 times longer to complete
+the algorithm. These results suggest that achieving full convergence of
+the optimization metric adds computational effort; however, considering
+full convergence took between 2 and 17 minutes the extra effort was
+relatively small. Although the mothur's default setting is partial
+convergence, the remainder of our analysis used complete convergence to
+be more conservative.
 
@@
 sequences because each algorithm must break ties where a sequence could
-be assigned to multiple OTUs. An additional concern with the OptiClust
-algorithm is that it may stabilize at a local optimum. To evaluate these
-concerns we compared the results obtained using ten randomizations of
-the order that sequences were given to the algorithm. The median the
-coefficient of variation across the six datasets for MCC values obtained
-from the replicate clusterings using OptiClust was 0.1\% (Figure 1). We
-also measured the coefficient of variation for the number of OTUs across
-the six datasets for each method. The median coefficient of variation
-for the number of OTUs generated using OptiClust was 0.1\%. Confirming
-our previous results, all of the methods we tested were stable to
-stochastic processes. Of the methods that involved randomization, the
-coefficient of variation for MCC values was considerably smaller with
-OptiClust than the other methods and the coefficient of variation for
-the number of OTUs was comparable to the other methods. The variation
-observed in clustering quality suggested that the algorithm does not
-appear to converge to a locally optimum MCC value. More importantly, the
-random variation does yield output of a similarly high quality.
+be assigned to multiple OTUs. An additional concern specific to the
+OptiClust algorithm is that it may stabilize at a local optimum. To
+evaluate these concerns we compared the results obtained using ten
+randomizations of the order that sequences were given to the algorithm.
+The median coefficient of variation across the six datasets for MCC
+values obtained from the replicate clusterings using OptiClust was 0.1\%
+(Figure 1). We also measured the coefficient of variation for the number
+of OTUs across the six datasets for each method. The median coefficient
+of variation for the number of OTUs generated using OptiClust was 0.1\%.
+Confirming our previous results (15), all of the methods we tested were
+stable to stochastic processes. Of the methods that involved
+randomization, the coefficient of variation for MCC values was
+considerably smaller with OptiClust than the other methods and the
+coefficient of variation for the number of OTUs was comparable to the
+other methods. The variation observed in clustering quality suggested
+that the algorithm does not appear to converge to a locally optimum MCC
+value. More importantly, the random variation does yield output of a
+similarly high quality.
 
@@
 required to complete the OptiClust algorithm (Figure 1C) paralleled the
-number of pairwise distances that were smaller than 0.03. To further
-evaluate how the speed and memory usage scaled with the number of
-sequences in the dataset, we measured the time required and maximum RAM
-usage to cluster 20, 40, 60, 80, and 100\% of the unique sequences from
-each of the natural datasets using the OptiClust algorithm (Figure 2).
-Within each iteration of the algorithm, each sequence is compared to
+number of pairwise distances that were smaller than 0.03 (Table 1). To
+further evaluate how the speed and memory usage scaled with the number
+of sequences in the dataset, we measured the time required and maximum
+RAM usage to cluster 20, 40, 60, 80, and 100\% of the unique sequences
+from each of the natural datasets using the OptiClust algorithm (Figure
+2). Within each iteration of the algorithm, each sequence is compared to
 every other sequence and each comparison requires a recalculation of the
 confusion matrix. This would result in a worst case algorithmic
-complexity on the order of N\^{}3, where N is the number of unique
-sequences. Because the algorithm only needs to keep track of the
-sequence pairs that are within the threshold of each other, it is likely
-that the implementation of the algorithm is more efficient. To
+complexity on the order of N\textsuperscript{3}, where N is the number
+of unique sequences. Because the algorithm only needs to keep track of
+the sequence pairs that are within the threshold of each other, it is
+likely that the implementation of the algorithm is more efficient. To
 empirically determine the algorithmic complexity, we fit a power law
@@
 sequences are clustered together. The memory required to store these
-matrices is on the order of N\^{}2, where N is the number of unique
-sequences. In fact, when we fit a power law function to the data in
-Figure 2B, the power coefficients were 1.9. Using the four natural
-community datasets, doubling the number of sequences in a dataset would
-increase the time required to cluster the data by 4 to 8-fold and
-increase the RAM required by 4-fold. It is possible that future
-improvements to the implementation of the algorithm could improve this
-performance.
+matrices is on the order of N\textsuperscript{2}, where N is the number
+of unique sequences. In fact, when we fit a power law function to the
+data in Figure 2B, the power coefficients were 1.9. Using the four
+natural community datasets, doubling the number of sequences in a
+dataset would increase the time required to cluster the data by 4 to
+8-fold and increase the RAM required by 4-fold. It is possible that
+future improvements to the implementation of the algorithm could improve
+this performance.
 
@@
 Our cluster splitting approach accelerated the clustering and reduced
-the memory requirements because the number of unique sequences is
+the memory requirements because the number of unique sequences was
 effectively reduced by splitting sequences across taxonomic groups.
@@
 parallelization and additional reduction in computation time. Reduction
-in clustering quality are encountered in this approach if there are
+in clustering quality is encountered in this approach if there are
 errors in classification or if two sequences within the desired
@@
 taxonomic level (Figure 3). For each method, the MCC values decreased as
-the taxonomic resolution increased; however, the decrease in MCC as not
+the taxonomic resolution increased; however, the decrease in MCC was not
 as large as the difference between clustering methods. As the resolution
@@
 Myriad methods have been proposed for assigning 16S rRNA gene sequences
-to OTUs that each claim improved performance based on speed, memory
-usage, representation of taxonomic information, and number of OTUs. Each
-of these metrics is subjective and do not actually indicate the quality
-of the clustering. This led us to propose using the MCC as a metric for
+to OTUs. Each claim improved performance based on speed, memory usage,
+representation of taxonomic information, and number of OTUs. Each of
+these metrics is subjective and do not actually indicate the quality of
+the clustering. This led us to propose using the MCC as a metric for
 assessing the quality of clustering, post hoc. Here, we described a new
@@
 objective criterion that measures clustering quality in real time. In
-the OptiClust algorithm clustering is driven by optimizing a metric that
-assesses whether any two sequences should be grouped into the same OTU.
-The result is clusters that are significantly more robust and is
+the OptiClust algorithm, clustering is driven by optimizing a metric
+that assesses whether any two sequences should be grouped into the same
+OTU. The result is clusters that are significantly more robust and is
 efficient in the time and memory required to cluster the sequences into
@@
 FP and FN, respectively; however, these suffer because the number of FN
-and FP are not controlled, respectively (13, 15). Alternatively, one
+and FP are not controlled, respectively (13, 16). Alternatively, one
 could optimize based on the sensitivity, specificity, or accuracy, which
 are each based on two values from the confusion matrix or they could
-optimize based on the F1 score, which is based on three values from the
+optimize based on the F1-score, which is based on three values from the
 confusion matrix. Because these metrics do not balance all four
@@
 (7). The method for generating the simulated datasets followed the
-approach used by Kopylova et al. (33) and Schloss (35). Briefly, we
+approach used by Kopylova et al. (34) and Schloss (36). Briefly, we
 randomly selected 10,000 uniques V4 fragments from 16S rRNA gene
@@
 (distance-based greedy clustering) DGC as implemented in USEARCH (v.6.1)
-(16) and VSEARCH (v.2.3.3) ((17){]}, OTUCLUST (v.0.1) (18), SumaClust
-(v.1.0.20), and Swarm (v.2.1.9) (19). With the exception of Swarm each
+(17) and VSEARCH (v.2.3.3) ((18){]}, OTUCLUST (v.0.1) (19), SumaClust
+(v.1.0.20), and Swarm (v.2.1.9) (20). With the exception of Swarm each
 of these methods uses distance-based thresholds to report OTU
-assignments.
+assignments. We also evalauted the ability of OptiClust to optimize to
+metrics other than MCC. These included accuracy, F1-score, negative
+predictive value, positive predictive value, false discovery rate,
+senitivity, specificity, the sum of TPs and TNs, the sum of FPs and FNs,
+and the number of FNs, FPs, TNs, and TPs (Figure S1).
 
@@
 than 3\% were false negatives (FNs). To synthesize the matrix into a
-single metric we used the Matthews Correlation Coefficient using the
+single metric we used the Matthews correlation coefficient using the
 \texttt{sens.spec} command in mothur using the following equations.
@@
 ran each collection (N= 50 = 5 sequencing depths x 10 replicates)
-through each of the algorithms. We used the timeout script to quantify
-the maximum RAM used and the amount of time required to process each
-sequence collection (\url{https://github.com/pshved/timeout}). We
+through each of the algorithms. We used the \texttt{timeout} script to
+quantify the maximum RAM used and the amount of time required to process
+each sequence collection (\url{https://github.com/pshved/timeout}). We
 limited each algorithm to 45 GB of RAM and 50 hours using a single
@@
 (v.0.3.2) (45), dplyr (v.0.5.0) (46), tidyr (v.0.6.0) (47), cowplot
-(v.0.6.3) (48), and ggplot2 (v.2.2.0.9000) (49) packages. A reproducible
-version of this manuscript and analysis is available at
+(v.0.6.9990) (48), and ggplot2 (v.2.1.0.9001) (49) packages. A
+reproducible version of this manuscript and analysis is available at
 \url{https://github.com/SchlossLab/Westcott_OptiClust_mSphere_2017}.
@@
 11,648\tabularnewline
-Even (33, 35) & NA & NA & 1,155,800 & 11,558 & 29,694 &
+Even (34, 36) & NA & NA & 1,155,800 & 11,558 & 29,694 &
 7,651\tabularnewline
-Staggered (33, 35) & NA & NA & 1,156,550 & 11,558 & 29,694 &
+Staggered (34, 36) & NA & NA & 1,156,550 & 11,558 & 29,694 &
 7,653\tabularnewline
@@
 
-\includegraphics[width=6.0in]{../results/figures/performance.png}
+\includegraphics[width=6.0in]{results/figures/performance.png}
 
@@
 
-\includegraphics[width=3.5in]{../results/figures/speed_memory.png}
+\includegraphics[width=3.5in]{results/figures/speed_memory.png}
 
@@
 
-\includegraphics[width=3.5in]{../results/figures/split_mcc.png}
+\includegraphics[width=3.5in]{results/figures/split_mcc.png}
 
@@
 
-\includegraphics[width=6.0in]{../results/figures/performance_other.png}
+\includegraphics[width=6.0in]{results/figures/performance_other.png}
 
@@
 
-\includegraphics[width=3.5in]{../results/figures/optimization.png}
+\includegraphics[width=3.5in]{results/figures/optimization.png}
 
 \textbf{Figure S2. The OptiClust algorithm rapidly converges to optimize
-the Matthew's Correlation Coefficient.} The six datasets were clustered
-into OTUs using the OptiClust algorithm seeking to maximize the
-Matthew's Correlation Coefficient. This was repeated 10 times for each
-dataset.
+the Matthews correlation coefficient.} The six datasets were clustered
+into OTUs using the OptiClust algorithm seeking to maximize the Matthews
+correlation coefficient. This was repeated 10 times for each dataset.
 
@@
 
+\hyperdef{}{references}{\label{references}}
 \subsection*{References}\label{references}
@@
 
-\hypertarget{refs}{}
-\hypertarget{ref-Schloss2016b}{}
+\hyperdef{}{ref-Schloss2016b}{\label{ref-Schloss2016b}}
 1. \textbf{Schloss PD}, \textbf{Girard RA}, \textbf{Martin T},
@@
 bacterial census: An update. mBio \textbf{7}:e00201--16.
-doi:\href{https://doi.org/10.1128/mbio.00201-16}{10.1128/mbio.00201-16}.
-
-\hypertarget{ref-Locey2016}{}
+doi:\url{http://doi.org/10.1128/mbio.00201-16}.
+
+\hyperdef{}{ref-Locey2016}{\label{ref-Locey2016}}
 2. \textbf{Locey KJ}, \textbf{Lennon JT}. 2016. Scaling laws predict
@@
 Sciences \textbf{113}:5970--5975.
-doi:\href{https://doi.org/10.1073/pnas.1521291113}{10.1073/pnas.1521291113}.
-
-\hypertarget{ref-Rideout2014}{}
+doi:\url{http://doi.org/10.1073/pnas.1521291113}.
+
+\hyperdef{}{ref-Rideout2014}{\label{ref-Rideout2014}}
 3. \textbf{Rideout JR}, \textbf{He Y}, \textbf{Navas-Molina JA},
@@
 comprehensive OTU definitions and scales to billions of sequences. PeerJ
-\textbf{2}:e545.
-doi:\href{https://doi.org/10.7717/peerj.545}{10.7717/peerj.545}.
-
-\hypertarget{ref-Huttenhower2012}{}
+\textbf{2}:e545. doi:\url{http://doi.org/10.7717/peerj.545}.
+
+\hyperdef{}{ref-Huttenhower2012}{\label{ref-Huttenhower2012}}
 4. \textbf{Consortium THMP}. 2012. Structure, function and diversity of
 the healthy human microbiome. Nature \textbf{486}:207--214.
-doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.
-
-\hypertarget{ref-Eckburg2005}{}
+doi:\url{http://doi.org/10.1038/nature11234}.
+
+\hyperdef{}{ref-Eckburg2005}{\label{ref-Eckburg2005}}
 5. \textbf{Eckburg PB}, \textbf{Bik EM}, \textbf{Bernstein CN},
@@
 
-\hypertarget{ref-Elshahed2008}{}
+\hyperdef{}{ref-Elshahed2008}{\label{ref-Elshahed2008}}
 6. \textbf{Elshahed MS}, \textbf{Youssef NH}, \textbf{Spain AM},
@@
 members of the soil biosphere. Applied and Environmental Microbiology
-\textbf{74}:5422--5428.
-doi:\href{https://doi.org/10.1128/aem.00410-08}{10.1128/aem.00410-08}.
-
-\hypertarget{ref-Kozich2013}{}
+\textbf{74}:5422--5428. doi:\url{http://doi.org/10.1128/aem.00410-08}.
+
+\hyperdef{}{ref-Kozich2013}{\label{ref-Kozich2013}}
 7. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
@@
 Applied and Environmental Microbiology \textbf{79}:5112--5120.
-doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
-
-\hypertarget{ref-Schloss2009a}{}
+doi:\url{http://doi.org/10.1128/aem.01043-13}.
+
+\hyperdef{}{ref-Schloss2009a}{\label{ref-Schloss2009a}}
 8. \textbf{Schloss PD}. 2009. A high-throughput DNA sequence aligner for
 microbial ecology studies. PLOS ONE \textbf{4}:e8230.
-doi:\href{https://doi.org/10.1371/journal.pone.0008230}{10.1371/journal.pone.0008230}.
-
-\hypertarget{ref-Edgar2011}{}
+doi:\url{http://doi.org/10.1371/journal.pone.0008230}.
+
+\hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
 9. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
@@
 and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
-doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
-
-\hypertarget{ref-Wang2007}{}
+doi:\url{http://doi.org/10.1093/bioinformatics/btr381}.
+
+\hyperdef{}{ref-Wang2007}{\label{ref-Wang2007}}
 10. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
@@
 Environmental Microbiology \textbf{73}:5261--5267.
-doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
-
-\hypertarget{ref-Schloss2009b}{}
+doi:\url{http://doi.org/10.1128/aem.00062-07}.
+
+\hyperdef{}{ref-Schloss2009b}{\label{ref-Schloss2009b}}
 11. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
@@
 communities. Applied and Environmental Microbiology
-\textbf{75}:7537--7541.
-doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
-
-\hypertarget{ref-Caporaso2010}{}
+\textbf{75}:7537--7541. doi:\url{http://doi.org/10.1128/aem.01541-09}.
+
+\hyperdef{}{ref-Caporaso2010}{\label{ref-Caporaso2010}}
 12. \textbf{Caporaso JG}, \textbf{Kuczynski J}, \textbf{Stombaugh J},
@@
 Nature Methods \textbf{7}:335--336.
-doi:\href{https://doi.org/10.1038/nmeth.f.303}{10.1038/nmeth.f.303}.
-
-\hypertarget{ref-Schloss2011}{}
+doi:\url{http://doi.org/10.1038/nmeth.f.303}.
+
+\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
 13. \textbf{Schloss PD}, \textbf{Westcott SL}. 2011. Assessing and
@@
 Microbiology \textbf{77}:3219--3226.
-doi:\href{https://doi.org/10.1128/aem.02810-10}{10.1128/aem.02810-10}.
-
-\hypertarget{ref-NavasMolina2013}{}
+doi:\url{http://doi.org/10.1128/aem.02810-10}.
+
+\hyperdef{}{ref-NavasMolina2013}{\label{ref-NavasMolina2013}}
 14. \textbf{Navas-Molina JA}, \textbf{Peralta-Sánchez JM},
@@
 
-\hypertarget{ref-Schloss2005}{}
-15. \textbf{Schloss PD}, \textbf{Handelsman J}. 2005. Introducing DOTUR,
+\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
+15. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
+methods outperform reference-based methods for assigning 16S rRNA gene
+sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
+doi:\url{http://doi.org/10.7717/peerj.1487}.
+
+\hyperdef{}{ref-Schloss2005}{\label{ref-Schloss2005}}
+16. \textbf{Schloss PD}, \textbf{Handelsman J}. 2005. Introducing DOTUR,
 a computer program for defining operational taxonomic units and
@@
 
-\hypertarget{ref-Edgar2010}{}
-16. \textbf{Edgar RC}. 2010. Search and clustering orders of magnitude
+\hyperdef{}{ref-Edgar2010}{\label{ref-Edgar2010}}
+17. \textbf{Edgar RC}. 2010. Search and clustering orders of magnitude
 faster than BLAST. Bioinformatics \textbf{26}:2460--2461.
-doi:\href{https://doi.org/10.1093/bioinformatics/btq461}{10.1093/bioinformatics/btq461}.
-
-\hypertarget{ref-Rognes2016}{}
-17. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+doi:\url{http://doi.org/10.1093/bioinformatics/btq461}.
+
+\hyperdef{}{ref-Rognes2016}{\label{ref-Rognes2016}}
+18. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
 \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
 source tool for metagenomics. PeerJ \textbf{4}:e2584.
-doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
-
-\hypertarget{ref-Albanese2015}{}
-18. \textbf{Albanese D}, \textbf{Fontana P}, \textbf{Filippo CD},
+doi:\url{http://doi.org/10.7717/peerj.2584}.
+
+\hyperdef{}{ref-Albanese2015}{\label{ref-Albanese2015}}
+19. \textbf{Albanese D}, \textbf{Fontana P}, \textbf{Filippo CD},
 \textbf{Cavalieri D}, \textbf{Donati C}. 2015. MICCA: A complete and
@@
 Scientific Reports \textbf{5}:9743.
-doi:\href{https://doi.org/10.1038/srep09743}{10.1038/srep09743}.
-
-\hypertarget{ref-Mah2014}{}
-19. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
+doi:\url{http://doi.org/10.1038/srep09743}.
+
+\hyperdef{}{ref-Mah2014}{\label{ref-Mah2014}}
+20. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
 \textbf{Vargas C de}, \textbf{Dunthorn M}. 2014. Swarm: Robust and fast
 clustering method for amplicon-based studies. PeerJ \textbf{2}:e593.
-doi:\href{https://doi.org/10.7717/peerj.593}{10.7717/peerj.593}.
-
-\hypertarget{ref-Sun2009}{}
-20. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Liu L}, \textbf{Yu F},
+doi:\url{http://doi.org/10.7717/peerj.593}.
+
+\hyperdef{}{ref-Sun2009}{\label{ref-Sun2009}}
+21. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Liu L}, \textbf{Yu F},
 \textbf{Farrell ML}, \textbf{McKendree W}, \textbf{Farmerie W}. 2009.
@@
 pyrosequences. Nucleic Acids Research \textbf{37}:e76--e76.
-doi:\href{https://doi.org/10.1093/nar/gkp285}{10.1093/nar/gkp285}.
-
-\hypertarget{ref-Cai2011}{}
-21. \textbf{Cai Y}, \textbf{Sun Y}. 2011. ESPRIT-tree: Hierarchical
+doi:\url{http://doi.org/10.1093/nar/gkp285}.
+
+\hyperdef{}{ref-Cai2011}{\label{ref-Cai2011}}
+22. \textbf{Cai Y}, \textbf{Sun Y}. 2011. ESPRIT-tree: Hierarchical
 clustering analysis of millions of 16S rRNA pyrosequences in quasilinear
 computational time. Nucleic Acids Research \textbf{39}:e95--e95.
-doi:\href{https://doi.org/10.1093/nar/gkr349}{10.1093/nar/gkr349}.
-
-\hypertarget{ref-Edgar2013}{}
-22. \textbf{Edgar RC}. 2013. UPARSE: Highly accurate OTU sequences from
+doi:\url{http://doi.org/10.1093/nar/gkr349}.
+
+\hyperdef{}{ref-Edgar2013}{\label{ref-Edgar2013}}
+23. \textbf{Edgar RC}. 2013. UPARSE: Highly accurate OTU sequences from
 microbial amplicon reads. Nature Methods \textbf{10}:996--998.
-doi:\href{https://doi.org/10.1038/nmeth.2604}{10.1038/nmeth.2604}.
-
-\hypertarget{ref-Mah2015}{}
-23. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
+doi:\url{http://doi.org/10.1038/nmeth.2604}.
+
+\hyperdef{}{ref-Mah2015}{\label{ref-Mah2015}}
+24. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
 \textbf{Vargas C de}, \textbf{Dunthorn M}. 2015. Swarm v2:
 Highly-scalable and high-resolution amplicon clustering. PeerJ
-\textbf{3}:e1420.
-doi:\href{https://doi.org/10.7717/peerj.1420}{10.7717/peerj.1420}.
-
-\hypertarget{ref-Barriuso2011}{}
-24. \textbf{Barriuso J}, \textbf{Valverde JR}, \textbf{Mellado RP}.
+\textbf{3}:e1420. doi:\url{http://doi.org/10.7717/peerj.1420}.
+
+\hyperdef{}{ref-Barriuso2011}{\label{ref-Barriuso2011}}
+25. \textbf{Barriuso J}, \textbf{Valverde JR}, \textbf{Mellado RP}.
 2011. Estimation of bacterial diversity using next generation sequencing
 of 16S rDNA: A comparison of different workflows. BMC Bioinformatics
-\textbf{12}:473.
-doi:\href{https://doi.org/10.1186/1471-2105-12-473}{10.1186/1471-2105-12-473}.
-
-\hypertarget{ref-Bonder2012}{}
-25. \textbf{Bonder MJ}, \textbf{Abeln S}, \textbf{Zaura E},
+\textbf{12}:473. doi:\url{http://doi.org/10.1186/1471-2105-12-473}.
+
+\hyperdef{}{ref-Bonder2012}{\label{ref-Bonder2012}}
+26. \textbf{Bonder MJ}, \textbf{Abeln S}, \textbf{Zaura E},
 \textbf{Brandt BW}. 2012. Comparing clustering and pre-processing in
 taxonomy analysis. Bioinformatics \textbf{28}:2891--2897.
-doi:\href{https://doi.org/10.1093/bioinformatics/bts552}{10.1093/bioinformatics/bts552}.
-
-\hypertarget{ref-Chen2013}{}
-26. \textbf{Chen W}, \textbf{Zhang CK}, \textbf{Cheng Y}, \textbf{Zhang
+doi:\url{http://doi.org/10.1093/bioinformatics/bts552}.
+
+\hyperdef{}{ref-Chen2013}{\label{ref-Chen2013}}
+27. \textbf{Chen W}, \textbf{Zhang CK}, \textbf{Cheng Y}, \textbf{Zhang
 S}, \textbf{Zhao H}. 2013. A comparison of methods for clustering 16S
 rRNA sequences into OTUs. PLOS ONE \textbf{8}:e70837.
-doi:\href{https://doi.org/10.1371/journal.pone.0070837}{10.1371/journal.pone.0070837}.
-
-\hypertarget{ref-Huse2010}{}
-27. \textbf{Huse SM}, \textbf{Welch DM}, \textbf{Morrison HG},
+doi:\url{http://doi.org/10.1371/journal.pone.0070837}.
+
+\hyperdef{}{ref-Huse2010}{\label{ref-Huse2010}}
+28. \textbf{Huse SM}, \textbf{Welch DM}, \textbf{Morrison HG},
 \textbf{Sogin ML}. 2010. Ironing out the wrinkles in the rare biosphere
@@
 \textbf{12}:1889--1898.
-doi:\href{https://doi.org/10.1111/j.1462-2920.2010.02193.x}{10.1111/j.1462-2920.2010.02193.x}.
-
-\hypertarget{ref-May2014}{}
-28. \textbf{May A}, \textbf{Abeln S}, \textbf{Crielaard W},
+doi:\url{http://doi.org/10.1111/j.1462-2920.2010.02193.x}.
+
+\hyperdef{}{ref-May2014}{\label{ref-May2014}}
+29. \textbf{May A}, \textbf{Abeln S}, \textbf{Crielaard W},
 \textbf{Heringa J}, \textbf{Brandt BW}. 2014. Unraveling the outcome of
@@
 Bioinformatics \textbf{30}:1530--1538.
-doi:\href{https://doi.org/10.1093/bioinformatics/btu085}{10.1093/bioinformatics/btu085}.
-
-\hypertarget{ref-Sun2011}{}
-29. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Huse SM}, \textbf{Knight R},
+doi:\url{http://doi.org/10.1093/bioinformatics/btu085}.
+
+\hyperdef{}{ref-Sun2011}{\label{ref-Sun2011}}
+30. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Huse SM}, \textbf{Knight R},
 \textbf{Farmerie WG}, \textbf{Wang X}, \textbf{Mai V}. 2011. A
@@
 Bioinformatics \textbf{13}:107--121.
-doi:\href{https://doi.org/10.1093/bib/bbr009}{10.1093/bib/bbr009}.
-
-\hypertarget{ref-White2010}{}
-30. \textbf{White JR}, \textbf{Navlakha S}, \textbf{Nagarajan N},
+doi:\url{http://doi.org/10.1093/bib/bbr009}.
+
+\hyperdef{}{ref-White2010}{\label{ref-White2010}}
+31. \textbf{White JR}, \textbf{Navlakha S}, \textbf{Nagarajan N},
 \textbf{Ghodsi M-R}, \textbf{Kingsford C}, \textbf{Pop M}. 2010.
@@
 microbial diversity studies. BMC Bioinformatics \textbf{11}:152.
-doi:\href{https://doi.org/10.1186/1471-2105-11-152}{10.1186/1471-2105-11-152}.
-
-\hypertarget{ref-AlGhalith2016}{}
-31. \textbf{Al-Ghalith GA}, \textbf{Montassier E}, \textbf{Ward HN},
+doi:\url{http://doi.org/10.1186/1471-2105-11-152}.
+
+\hyperdef{}{ref-AlGhalith2016}{\label{ref-AlGhalith2016}}
+32. \textbf{Al-Ghalith GA}, \textbf{Montassier E}, \textbf{Ward HN},
 \textbf{Knights D}. 2016. NINJA-OPS: Fast accurate marker gene alignment
@@
 \textbf{12}:e1004658.
-doi:\href{https://doi.org/10.1371/journal.pcbi.1004658}{10.1371/journal.pcbi.1004658}.
-
-\hypertarget{ref-He2015}{}
-32. \textbf{He Y}, \textbf{Caporaso JG}, \textbf{Jiang X-T},
+doi:\url{http://doi.org/10.1371/journal.pcbi.1004658}.
+
+\hyperdef{}{ref-He2015}{\label{ref-He2015}}
+33. \textbf{He Y}, \textbf{Caporaso JG}, \textbf{Jiang X-T},
 \textbf{Sheng H-F}, \textbf{Huse SM}, \textbf{Rideout JR}, \textbf{Edgar
@@
 Microbiome \textbf{3}.
-doi:\href{https://doi.org/10.1186/s40168-015-0081-x}{10.1186/s40168-015-0081-x}.
-
-\hypertarget{ref-Kopylova2016}{}
-33. \textbf{Kopylova E}, \textbf{Navas-Molina JA}, \textbf{Mercier C},
+doi:\url{http://doi.org/10.1186/s40168-015-0081-x}.
+
+\hyperdef{}{ref-Kopylova2016}{\label{ref-Kopylova2016}}
+34. \textbf{Kopylova E}, \textbf{Navas-Molina JA}, \textbf{Mercier C},
 \textbf{Xu ZZ}, \textbf{Mahé F}, \textbf{He Y}, \textbf{Zhou H-W},
@@
 mSystems \textbf{1}:e00003--15.
-doi:\href{https://doi.org/10.1128/msystems.00003-15}{10.1128/msystems.00003-15}.
-
-\hypertarget{ref-Schmidt2014}{}
-34. \textbf{Schmidt TSB}, \textbf{Rodrigues JFM}, \textbf{Mering C von}.
+doi:\url{http://doi.org/10.1128/msystems.00003-15}.
+
+\hyperdef{}{ref-Schmidt2014}{\label{ref-Schmidt2014}}
+35. \textbf{Schmidt TSB}, \textbf{Rodrigues JFM}, \textbf{Mering C von}.
 2014. Limits to robustness and reproducibility in the demarcation of
 operational taxonomic units. Environ Microbiol \textbf{17}:1689--1706.
-doi:\href{https://doi.org/10.1111/1462-2920.12610}{10.1111/1462-2920.12610}.
-
-\hypertarget{ref-Schloss2016a}{}
-35. \textbf{Schloss PD}. 2016. Application of a database-independent
+doi:\url{http://doi.org/10.1111/1462-2920.12610}.
+
+\hyperdef{}{ref-Schloss2016a}{\label{ref-Schloss2016a}}
+36. \textbf{Schloss PD}. 2016. Application of a database-independent
 approach to assess the quality of operational taxonomic unit picking
 methods. mSystems \textbf{1}:e00027--16.
-doi:\href{https://doi.org/10.1128/msystems.00027-16}{10.1128/msystems.00027-16}.
-
-\hypertarget{ref-Westcott2015}{}
-36. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
-methods outperform reference-based methods for assigning 16S rRNA gene
-sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
-
-\hypertarget{ref-Matthews1975}{}
+doi:\url{http://doi.org/10.1128/msystems.00027-16}.
+
+\hyperdef{}{ref-Matthews1975}{\label{ref-Matthews1975}}
 37. \textbf{Matthews B}. 1975. Comparison of the predicted and observed
@@
 (BBA) - Protein Structure \textbf{405}:442--451.
-doi:\href{https://doi.org/10.1016/0005-2795(75)90109-9}{10.1016/0005-2795(75)90109-9}.
-
-\hypertarget{ref-Pruesse2007}{}
+doi:\url{http://doi.org/10.1016/0005-2795(75)90109-9}.
+
+\hyperdef{}{ref-Pruesse2007}{\label{ref-Pruesse2007}}
 38. \textbf{Pruesse E}, \textbf{Quast C}, \textbf{Knittel K},
@@
 ARB. Nucleic Acids Research \textbf{35}:7188--7196.
-doi:\href{https://doi.org/10.1093/nar/gkm864}{10.1093/nar/gkm864}.
-
-\hypertarget{ref-Baxter2016}{}
+doi:\url{http://doi.org/10.1093/nar/gkm864}.
+
+\hyperdef{}{ref-Baxter2016}{\label{ref-Baxter2016}}
 39. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
@@
 Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
-
-\hypertarget{ref-Schloss2012}{}
+doi:\url{http://doi.org/10.1186/s13073-016-0290-3}.
+
+\hyperdef{}{ref-Schloss2012}{\label{ref-Schloss2012}}
 40. \textbf{Schloss PD}, \textbf{Schubert AM}, \textbf{Zackular JP},
@@
 Microbes \textbf{3}:383--393.
-doi:\href{https://doi.org/10.4161/gmic.21008}{10.4161/gmic.21008}.
-
-\hypertarget{ref-Johnston2016}{}
+doi:\url{http://doi.org/10.4161/gmic.21008}.
+
+\hyperdef{}{ref-Johnston2016}{\label{ref-Johnston2016}}
 41. \textbf{Johnston ER}, \textbf{Rodriguez-R LM}, \textbf{Luo C},
@@
 tundra ecosystem. Front Microbiol \textbf{7}.
-doi:\href{https://doi.org/10.3389/fmicb.2016.00579}{10.3389/fmicb.2016.00579}.
-
-\hypertarget{ref-Henson2016}{}
+doi:\url{http://doi.org/10.3389/fmicb.2016.00579}.
+
+\hyperdef{}{ref-Henson2016}{\label{ref-Henson2016}}
 42. \textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst JL},
@@
 \textbf{1}:e00028--16.
-doi:\href{https://doi.org/10.1128/msphere.00028-16}{10.1128/msphere.00028-16}.
-
-\hypertarget{ref-Schloss2010}{}
+doi:\url{http://doi.org/10.1128/msphere.00028-16}.
+
+\hyperdef{}{ref-Schloss2010}{\label{ref-Schloss2010}}
 43. \textbf{Schloss PD}. 2010. The effects of alignment quality,
@@
 \textbf{6}:e1000844.
-doi:\href{https://doi.org/10.1371/journal.pcbi.1000844}{10.1371/journal.pcbi.1000844}.
-
-\hypertarget{ref-language2015}{}
+doi:\url{http://doi.org/10.1371/journal.pcbi.1000844}.
+
+\hyperdef{}{ref-language2015}{\label{ref-language2015}}
 44. \textbf{R Core Team}. 2015. R: A language and environment for
@@
 
-\hypertarget{ref-wesanderson}{}
+\hyperdef{}{ref-wesanderson}{\label{ref-wesanderson}}
 45. \textbf{Ram K}, \textbf{Wickham H}. 2015. wesanderson: A wes
@@
 
-\hypertarget{ref-dplyr}{}
+\hyperdef{}{ref-dplyr}{\label{ref-dplyr}}
 46. \textbf{Wickham H}, \textbf{Francois R}. 2016. dplyr: A grammar of
@@
 
-\hypertarget{ref-tidyr}{}
+\hyperdef{}{ref-tidyr}{\label{ref-tidyr}}
 47. \textbf{Wickham H}. 2016. tidyr: Easily tidy data with `spread()`
@@
 
-\hypertarget{ref-cowplot}{}
+\hyperdef{}{ref-cowplot}{\label{ref-cowplot}}
 48. \textbf{Wilke CO}. cowplot: Streamlined plot theme and plot
@@
 
-\hypertarget{ref-ggplot2}{}
+\hyperdef{}{ref-ggplot2}{\label{ref-ggplot2}}
 49. \textbf{Wickham H}. 2009. ggplot2: Elegant graphics for data

### 2017-02-03 submission/response_to_reviewers.md
Finalize response to reviewers

@@
 ---
+
+The line numbers referred to in our response to reviewers correspond to those in the version of the manuscript with the "track changes" noted (i.e. `track_changes.pdf`).
 
@@
 
-We have added text to the Introduction to describe that the furthest and nearest neighbor algorithms seek to minimize the number of FPs and FNs, respectively (***LXXX***).
-We have also added text to the Discussion (***LXXX***) pointing out that More FPs would increase the number of OTUs while more FNs would collapse OTUs together.
+We have added text to the Introduction to describe that the furthest and nearest neighbor algorithms no not allow FPs and FNs, respectively (L83-84).
+We have also added text to the Discussion (L244-251) pointing out that More FPs would increase the number of OTUs while more FNs would collapse OTUs together.
 It is difficult to know which is worse since both alpha and beta diversity metrics are tied to the number and composition of OTUs.
@@
 Except under highly contrived conditions, we do not know what the true values are for output measures such as the number of clusters.
-This problem is outlined in ***LXXX-LXXX***.
+This problem is outlined in L62-88.
 This is why we have developed the concept of the MCC over our last several papers on the subject of OTU clustering.
@@
 Unfortunately, some of the datasets are too large and the sequencing data too noisy to permit the construction of the phylogenetic trees that would be needed to calculate the phylogenetic diversity for each study.
-As for comparing to percentage of sequences in a reference database, we feel that this is would be a misleading point as we have previously shown that reference-based methods are inferior to *de novo* clustering methods (***XXXX***) and the identity of the sequences does not affect the performance of the algorithm (i.e.
+As for comparing to percentage of sequences in a reference database, we feel that this is would be a misleading point as we have previously shown that reference-based methods are inferior to *de novo* clustering methods (L47; reference added) and the identity of the sequences does not affect the performance of the algorithm (i.e.
 we could have used synthetic DNA sequences and communities and reached the same conclusion).
@@
 
-We have added two supplementary figures.
+Because both reviewers inquired about this point, we added two supplementary figures.
 The first demonstrates the performance of the algorithm when we optimized clustering based on a variety of other metrics (Figure S1).
@@
 
-We have added a sentence to the Introduction (***LXXX***) indicating that all of the *de novo* methods used by microbial ecologists are agglomerative.
+We have added a sentence to the Introduction (L52) indicating that all of the *de novo* methods used by microbial ecologists are agglomerative.
 
@@
 
-We have added a line at ***LXXX*** and ***LXXX*** to indicate that speed was proportionate to the number of distances smaller than 0.03.
+We have added text (L175-177) to indicate that speed was proportionate to the number of distances smaller than 0.03.
 It does not necessarily appear to be related to the number of sequences or the number of OTUs.
@@
 
-A sentence has been added to this section to clarify the differnces between our method and open reference clustering.
+A sentence has been added to clarify the differences between our method and open reference clustering (L198-201).
 Briefly, in open reference clustering sequences are classified and any that do not map to a reference are clustered using a de novo approach.
@@
 
-At ***LXXX***, our goal wasn't to compare RAM usage between methods.
+At L120-123, our goal wasn't to compare RAM usage between methods.
 We merely wanted to point out that the USEARCH-based approaches could not handle large datasets without purchasing the 64-bit version of the software and that OTUCLUST and SumClust could not store the data within 45 GB of RAM.
@@
 
-These were run using the implementations found in mothur (***LXXXX***).
+These were run using the implementations found in mothur (L313).
 
@@
 
-We have added text to the manuscript in response to this comment and that of the first reviewer.
+We have added text and two supplemental figures to the manuscript in response to this comment and that of the first reviewer. We addressed this point above.

### 2017-02-09 results/tables/table_s1.Rmd
Finalize documents

---
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: table_header.tex
tables: true
geometry: margin=0.2in
---

**Table S1. Summary of the average number of true positives, true negatives, false positives, false negatives and the resulting Matthews correlation coefficient for each of the clustering methods that were analyzed in this study for each of the six datasets.** Blank values indicate that those conditions could not be completed in 50 hours with 45 GB of RAM.

```{r results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
library(xtable)
library(dplyr, warn.conflicts=FALSE)

my_methods <- c("accuracy", "an", "f1score", "fn", "fpfn", "mcc", "mcc_agg", "nn", "npv", "opti_fn",
							"otuclust", "sens", "sumaclust", "swarm", "tptn", "uagc", "udgc", "vagc_1",
							"vdgc_1")

pretty_methods <- c(accuracy = "Accuracy", an = "Average Neighbor", f1score = "F1-score",
 										fn = "Furthest Neighbor", fpfn = "FPs + FNs",
										mcc = "MCC (separate OTUs)", mcc_agg = "MCC (single OTU)",
										nn = "Nearest Neighbor", npv = "Negative Predictive Value",
										opti_fn = "FN", otuclust = "OTUCLUST", sens = "Sensitivity",
										sumaclust = "Sumaclust", swarm = "Swarm", tptn = "TP + TN",
										uagc = "USEARCH (w/AGC)", udgc = "USEARCH (w/DGC)",
										vagc_1 = "VSEARCH (w/AGC)", vdgc_1 = "VSEARCH (w/AGC)")

datasets <- c('soil', 'marine', 'mice', 'human', 'even', 'staggered')

pretty_datasets <- c(soil = "Soil", marine = "Marine", mice = "Mice",
										human = "Human", even = "Even", staggered = "Staggered")

data <- read.table(file="data/processed/cluster_data.summary", header=T, stringsAsFactors = T)
data$dataset = factor(data$dataset, levels=c("soil", "marine", "mice", "human", "even", "staggered"))

summary_data <- data %>% filter(frac==1.0, method %in% my_methods) %>%
													select(-c(frac, cluster_secs, cluster_kb, dist_secs)) %>%
													group_by(dataset, method) %>%
													summarize(tp.mean = round(mean(tp)),
																		tn.mean = round(mean(tn)),
																		fp.mean = round(mean(fp)),
																		fn.mean = round(mean(fn)),
																		mcc.mean = mean(mcc)) %>%
													arrange(dataset, -mcc.mean)

summary_data$dataset <- pretty_datasets[as.character(summary_data$dataset)]
summary_data$method <- pretty_methods[as.character(summary_data$method)]

colnames(summary_data) <- c("Dataset",
													"Method",
													"\\bigcell{c}{True \\\\ Positives}",
													"\\bigcell{c}{True \\\\ Negatives}",
													"\\bigcell{c}{False \\\\ Positives}",
													"\\bigcell{c}{False \\\\ Negatives}",
													"MCC")
colnames(summary_data) <- paste0("\\textbf{", colnames(summary_data), "}")

options(xtable.sanitize.text.function=identity)

composite <- xtable(summary_data, label="")

align(composite) <- "llcrrrrc"
digits(composite) <- c(0,0,0,0,0,0,0,4)
print(composite,
		include.rownames=FALSE,
		include.colnames=TRUE,
		hline.after = c(-1,0, nrow(composite)),
		comment=FALSE,
		size="tiny",
		table.placement="!ht"
		)
```

### 2017-02-09 submission/Westcott_OptiClust_mSphere_2016.Rmd
Finalize documents

@@
 Although numerous options are available for optimizing the clusters and for assessing the quality of the clusters within the mothur-based implementation of the algorithm (e.g.
-sensitivity, specificity, accuracy, F1-score, etc.), the default metric for optimization and assessment is MCC because it includes all four parameters from the confusion matrix (Figure S1).
+sensitivity, specificity, accuracy, F1-score, etc.), the default metric for optimization and assessment is MCC because it includes all four parameters from the confusion matrix (Figure S1; Table S1).
 The algorithm continues until the optimization metric stabilizes or until it reaches a defined stopping criteria.
@@
 To evaluate the OptiClust algorithm and compare its performance to other algorithms, we utilized six datasets including two synthetic communities and four previously published large datasets generated from soil, marine, human, and murine samples (Table 1).
-When we seeded the OptiClust algorithm with each sequence in a separate OTU and ran the algorithm until complete convergence, the MCC values averaged `r round(mcc_rel_an, 1)` and `r round(mcc_rel_vdgc, 1)`% higher than the OTUs using average neighbor and distance-based greedy clustering (DGC) with VSEARCH, respectively (Figure 1).
+When we seeded the OptiClust algorithm with each sequence in a separate OTU and ran the algorithm until complete convergence, the MCC values averaged `r round(mcc_rel_an, 1)` and `r round(mcc_rel_vdgc, 1)`% higher than the OTUs using average neighbor and distance-based greedy clustering (DGC) with VSEARCH, respectively (Figure 1; Table S1).
 The number of OTUs formed by the various methods was negatively correlated with their MCC value ($\rho$=`r round(mcc_sobs_cor$estimate, 2)`; p`r p<-mcc_sobs_cor$p.value; ifelse(p<0.001, "<0.001", paste0("=", p))`).
@@
 We also evalauted the ability of OptiClust to optimize to metrics other than MCC.
-These included accuracy, F1-score, negative predictive value, positive predictive value, false discovery rate, senitivity, specificity, the sum of TPs and TNs, the sum of FPs and FNs, and the number of FNs, FPs, TNs, and TPs (Figure S1).
+These included accuracy, F1-score, negative predictive value, positive predictive value, false discovery rate, senitivity, specificity, the sum of TPs and TNs, the sum of FPs and FNs, and the number of FNs, FPs, TNs, and TPs (Figure S1; Table S1).
 
@@
 
+\newpage
+
 **Supplemental text.** Worked example of how OptiClust algorithm clusters sequences into OTUs.
+
+**Table S1.
+Summary of the average number of true positives, true negatives, false positives, false negatives and the resulting Matthews correlation coefficient for each of the clustering methods that were analyzed in this study for each of the six datasets.** Blank values indicate that those conditions could not be completed in 50 hours with 45 GB of RAM.
 

### 2017-02-09 submission/Westcott_OptiClust_mSphere_2016.tex
Finalize documents

@@
     \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
   \newcommand{\euro}{€}
@@
 \usepackage[margin=1.0in]{geometry}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
+\usepackage{hyperref}
+\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
+\hypersetup{unicode=true,
             pdftitle={OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
+            pdfborder={0 0 0},
+            breaklinks=true}
 \urlstyle{same}  % don't use monospace font for urls
@@
 assessment is MCC because it includes all four parameters from the
-confusion matrix (Figure S1). The algorithm continues until the
-optimization metric stabilizes or until it reaches a defined stopping
-criteria.
+confusion matrix (Figure S1; Table S1). The algorithm continues until
+the optimization metric stabilizes or until it reaches a defined
+stopping criteria.
 
@@
 distance-based greedy clustering (DGC) with VSEARCH, respectively
-(Figure 1). The number of OTUs formed by the various methods was
-negatively correlated with their MCC value (\(\rho\)=-0.47;
+(Figure 1; Table S1). The number of OTUs formed by the various methods
+was negatively correlated with their MCC value (\(\rho\)=-0.47;
 p\textless{}0.001). The OptiClust algorithm was considerably faster than
@@
 senitivity, specificity, the sum of TPs and TNs, the sum of FPs and FNs,
-and the number of FNs, FPs, TNs, and TPs (Figure S1).
+and the number of FNs, FPs, TNs, and TPs (Figure S1; Table S1).
 
@@
 (v.0.3.2) (45), dplyr (v.0.5.0) (46), tidyr (v.0.6.0) (47), cowplot
-(v.0.6.9990) (48), and ggplot2 (v.2.1.0.9001) (49) packages. A
-reproducible version of this manuscript and analysis is available at
+(v.0.6.3) (48), and ggplot2 (v.2.2.0.9000) (49) packages. A reproducible
+version of this manuscript and analysis is available at
 \url{https://github.com/SchlossLab/Westcott_OptiClust_mSphere_2017}.
@@
 
+\newpage
+
 \textbf{Supplemental text.} Worked example of how OptiClust algorithm
@@
 
+\textbf{Table S1. Summary of the average number of true positives, true
+negatives, false positives, false negatives and the resulting Matthews
+correlation coefficient for each of the clustering methods that were
+analyzed in this study for each of the six datasets.} Blank values
+indicate that those conditions could not be completed in 50 hours with
+45 GB of RAM.
+
 \newpage
 
-\hyperdef{}{references}{\label{references}}
 \subsection*{References}\label{references}
@@
 
-\hyperdef{}{ref-Schloss2016b}{\label{ref-Schloss2016b}}
+\hypertarget{refs}{}
+\hypertarget{ref-Schloss2016b}{}
 1. \textbf{Schloss PD}, \textbf{Girard RA}, \textbf{Martin T},
@@
 bacterial census: An update. mBio \textbf{7}:e00201--16.
-doi:\url{http://doi.org/10.1128/mbio.00201-16}.
-
-\hyperdef{}{ref-Locey2016}{\label{ref-Locey2016}}
+doi:\href{https://doi.org/10.1128/mbio.00201-16}{10.1128/mbio.00201-16}.
+
+\hypertarget{ref-Locey2016}{}
 2. \textbf{Locey KJ}, \textbf{Lennon JT}. 2016. Scaling laws predict
@@
 Sciences \textbf{113}:5970--5975.
-doi:\url{http://doi.org/10.1073/pnas.1521291113}.
-
-\hyperdef{}{ref-Rideout2014}{\label{ref-Rideout2014}}
+doi:\href{https://doi.org/10.1073/pnas.1521291113}{10.1073/pnas.1521291113}.
+
+\hypertarget{ref-Rideout2014}{}
 3. \textbf{Rideout JR}, \textbf{He Y}, \textbf{Navas-Molina JA},
@@
 comprehensive OTU definitions and scales to billions of sequences. PeerJ
-\textbf{2}:e545. doi:\url{http://doi.org/10.7717/peerj.545}.
-
-\hyperdef{}{ref-Huttenhower2012}{\label{ref-Huttenhower2012}}
+\textbf{2}:e545.
+doi:\href{https://doi.org/10.7717/peerj.545}{10.7717/peerj.545}.
+
+\hypertarget{ref-Huttenhower2012}{}
 4. \textbf{Consortium THMP}. 2012. Structure, function and diversity of
 the healthy human microbiome. Nature \textbf{486}:207--214.
-doi:\url{http://doi.org/10.1038/nature11234}.
-
-\hyperdef{}{ref-Eckburg2005}{\label{ref-Eckburg2005}}
+doi:\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.
+
+\hypertarget{ref-Eckburg2005}{}
 5. \textbf{Eckburg PB}, \textbf{Bik EM}, \textbf{Bernstein CN},
@@
 
-\hyperdef{}{ref-Elshahed2008}{\label{ref-Elshahed2008}}
+\hypertarget{ref-Elshahed2008}{}
 6. \textbf{Elshahed MS}, \textbf{Youssef NH}, \textbf{Spain AM},
@@
 members of the soil biosphere. Applied and Environmental Microbiology
-\textbf{74}:5422--5428. doi:\url{http://doi.org/10.1128/aem.00410-08}.
-
-\hyperdef{}{ref-Kozich2013}{\label{ref-Kozich2013}}
+\textbf{74}:5422--5428.
+doi:\href{https://doi.org/10.1128/aem.00410-08}{10.1128/aem.00410-08}.
+
+\hypertarget{ref-Kozich2013}{}
 7. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
@@
 Applied and Environmental Microbiology \textbf{79}:5112--5120.
-doi:\url{http://doi.org/10.1128/aem.01043-13}.
-
-\hyperdef{}{ref-Schloss2009a}{\label{ref-Schloss2009a}}
+doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\hypertarget{ref-Schloss2009a}{}
 8. \textbf{Schloss PD}. 2009. A high-throughput DNA sequence aligner for
 microbial ecology studies. PLOS ONE \textbf{4}:e8230.
-doi:\url{http://doi.org/10.1371/journal.pone.0008230}.
-
-\hyperdef{}{ref-Edgar2011}{\label{ref-Edgar2011}}
+doi:\href{https://doi.org/10.1371/journal.pone.0008230}{10.1371/journal.pone.0008230}.
+
+\hypertarget{ref-Edgar2011}{}
 9. \textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente JC},
@@
 and speed of chimera detection. Bioinformatics \textbf{27}:2194--2200.
-doi:\url{http://doi.org/10.1093/bioinformatics/btr381}.
-
-\hyperdef{}{ref-Wang2007}{\label{ref-Wang2007}}
+doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
+
+\hypertarget{ref-Wang2007}{}
 10. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
@@
 Environmental Microbiology \textbf{73}:5261--5267.
-doi:\url{http://doi.org/10.1128/aem.00062-07}.
-
-\hyperdef{}{ref-Schloss2009b}{\label{ref-Schloss2009b}}
+doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
+
+\hypertarget{ref-Schloss2009b}{}
 11. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
@@
 communities. Applied and Environmental Microbiology
-\textbf{75}:7537--7541. doi:\url{http://doi.org/10.1128/aem.01541-09}.
-
-\hyperdef{}{ref-Caporaso2010}{\label{ref-Caporaso2010}}
+\textbf{75}:7537--7541.
+doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+
+\hypertarget{ref-Caporaso2010}{}
 12. \textbf{Caporaso JG}, \textbf{Kuczynski J}, \textbf{Stombaugh J},
@@
 Nature Methods \textbf{7}:335--336.
-doi:\url{http://doi.org/10.1038/nmeth.f.303}.
-
-\hyperdef{}{ref-Schloss2011}{\label{ref-Schloss2011}}
+doi:\href{https://doi.org/10.1038/nmeth.f.303}{10.1038/nmeth.f.303}.
+
+\hypertarget{ref-Schloss2011}{}
 13. \textbf{Schloss PD}, \textbf{Westcott SL}. 2011. Assessing and
@@
 Microbiology \textbf{77}:3219--3226.
-doi:\url{http://doi.org/10.1128/aem.02810-10}.
-
-\hyperdef{}{ref-NavasMolina2013}{\label{ref-NavasMolina2013}}
+doi:\href{https://doi.org/10.1128/aem.02810-10}{10.1128/aem.02810-10}.
+
+\hypertarget{ref-NavasMolina2013}{}
 14. \textbf{Navas-Molina JA}, \textbf{Peralta-Sánchez JM},
@@
 
-\hyperdef{}{ref-Westcott2015}{\label{ref-Westcott2015}}
+\hypertarget{ref-Westcott2015}{}
 15. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
@@
 sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\url{http://doi.org/10.7717/peerj.1487}.
-
-\hyperdef{}{ref-Schloss2005}{\label{ref-Schloss2005}}
+doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
+
+\hypertarget{ref-Schloss2005}{}
 16. \textbf{Schloss PD}, \textbf{Handelsman J}. 2005. Introducing DOTUR,
@@
 
-\hyperdef{}{ref-Edgar2010}{\label{ref-Edgar2010}}
+\hypertarget{ref-Edgar2010}{}
 17. \textbf{Edgar RC}. 2010. Search and clustering orders of magnitude
 faster than BLAST. Bioinformatics \textbf{26}:2460--2461.
-doi:\url{http://doi.org/10.1093/bioinformatics/btq461}.
-
-\hyperdef{}{ref-Rognes2016}{\label{ref-Rognes2016}}
+doi:\href{https://doi.org/10.1093/bioinformatics/btq461}{10.1093/bioinformatics/btq461}.
+
+\hypertarget{ref-Rognes2016}{}
 18. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
@@
 source tool for metagenomics. PeerJ \textbf{4}:e2584.
-doi:\url{http://doi.org/10.7717/peerj.2584}.
-
-\hyperdef{}{ref-Albanese2015}{\label{ref-Albanese2015}}
+doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
+
+\hypertarget{ref-Albanese2015}{}
 19. \textbf{Albanese D}, \textbf{Fontana P}, \textbf{Filippo CD},
@@
 Scientific Reports \textbf{5}:9743.
-doi:\url{http://doi.org/10.1038/srep09743}.
-
-\hyperdef{}{ref-Mah2014}{\label{ref-Mah2014}}
+doi:\href{https://doi.org/10.1038/srep09743}{10.1038/srep09743}.
+
+\hypertarget{ref-Mah2014}{}
 20. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
@@
 clustering method for amplicon-based studies. PeerJ \textbf{2}:e593.
-doi:\url{http://doi.org/10.7717/peerj.593}.
-
-\hyperdef{}{ref-Sun2009}{\label{ref-Sun2009}}
+doi:\href{https://doi.org/10.7717/peerj.593}{10.7717/peerj.593}.
+
+\hypertarget{ref-Sun2009}{}
 21. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Liu L}, \textbf{Yu F},
@@
 pyrosequences. Nucleic Acids Research \textbf{37}:e76--e76.
-doi:\url{http://doi.org/10.1093/nar/gkp285}.
-
-\hyperdef{}{ref-Cai2011}{\label{ref-Cai2011}}
+doi:\href{https://doi.org/10.1093/nar/gkp285}{10.1093/nar/gkp285}.
+
+\hypertarget{ref-Cai2011}{}
 22. \textbf{Cai Y}, \textbf{Sun Y}. 2011. ESPRIT-tree: Hierarchical
@@
 computational time. Nucleic Acids Research \textbf{39}:e95--e95.
-doi:\url{http://doi.org/10.1093/nar/gkr349}.
-
-\hyperdef{}{ref-Edgar2013}{\label{ref-Edgar2013}}
+doi:\href{https://doi.org/10.1093/nar/gkr349}{10.1093/nar/gkr349}.
+
+\hypertarget{ref-Edgar2013}{}
 23. \textbf{Edgar RC}. 2013. UPARSE: Highly accurate OTU sequences from
 microbial amplicon reads. Nature Methods \textbf{10}:996--998.
-doi:\url{http://doi.org/10.1038/nmeth.2604}.
-
-\hyperdef{}{ref-Mah2015}{\label{ref-Mah2015}}
+doi:\href{https://doi.org/10.1038/nmeth.2604}{10.1038/nmeth.2604}.
+
+\hypertarget{ref-Mah2015}{}
 24. \textbf{Mahé F}, \textbf{Rognes T}, \textbf{Quince C},
@@
 Highly-scalable and high-resolution amplicon clustering. PeerJ
-\textbf{3}:e1420. doi:\url{http://doi.org/10.7717/peerj.1420}.
-
-\hyperdef{}{ref-Barriuso2011}{\label{ref-Barriuso2011}}
+\textbf{3}:e1420.
+doi:\href{https://doi.org/10.7717/peerj.1420}{10.7717/peerj.1420}.
+
+\hypertarget{ref-Barriuso2011}{}
 25. \textbf{Barriuso J}, \textbf{Valverde JR}, \textbf{Mellado RP}.
@@
 of 16S rDNA: A comparison of different workflows. BMC Bioinformatics
-\textbf{12}:473. doi:\url{http://doi.org/10.1186/1471-2105-12-473}.
-
-\hyperdef{}{ref-Bonder2012}{\label{ref-Bonder2012}}
+\textbf{12}:473.
+doi:\href{https://doi.org/10.1186/1471-2105-12-473}{10.1186/1471-2105-12-473}.
+
+\hypertarget{ref-Bonder2012}{}
 26. \textbf{Bonder MJ}, \textbf{Abeln S}, \textbf{Zaura E},
@@
 taxonomy analysis. Bioinformatics \textbf{28}:2891--2897.
-doi:\url{http://doi.org/10.1093/bioinformatics/bts552}.
-
-\hyperdef{}{ref-Chen2013}{\label{ref-Chen2013}}
+doi:\href{https://doi.org/10.1093/bioinformatics/bts552}{10.1093/bioinformatics/bts552}.
+
+\hypertarget{ref-Chen2013}{}
 27. \textbf{Chen W}, \textbf{Zhang CK}, \textbf{Cheng Y}, \textbf{Zhang
@@
 rRNA sequences into OTUs. PLOS ONE \textbf{8}:e70837.
-doi:\url{http://doi.org/10.1371/journal.pone.0070837}.
-
-\hyperdef{}{ref-Huse2010}{\label{ref-Huse2010}}
+doi:\href{https://doi.org/10.1371/journal.pone.0070837}{10.1371/journal.pone.0070837}.
+
+\hypertarget{ref-Huse2010}{}
 28. \textbf{Huse SM}, \textbf{Welch DM}, \textbf{Morrison HG},
@@
 \textbf{12}:1889--1898.
-doi:\url{http://doi.org/10.1111/j.1462-2920.2010.02193.x}.
-
-\hyperdef{}{ref-May2014}{\label{ref-May2014}}
+doi:\href{https://doi.org/10.1111/j.1462-2920.2010.02193.x}{10.1111/j.1462-2920.2010.02193.x}.
+
+\hypertarget{ref-May2014}{}
 29. \textbf{May A}, \textbf{Abeln S}, \textbf{Crielaard W},
@@
 Bioinformatics \textbf{30}:1530--1538.
-doi:\url{http://doi.org/10.1093/bioinformatics/btu085}.
-
-\hyperdef{}{ref-Sun2011}{\label{ref-Sun2011}}
+doi:\href{https://doi.org/10.1093/bioinformatics/btu085}{10.1093/bioinformatics/btu085}.
+
+\hypertarget{ref-Sun2011}{}
 30. \textbf{Sun Y}, \textbf{Cai Y}, \textbf{Huse SM}, \textbf{Knight R},
@@
 Bioinformatics \textbf{13}:107--121.
-doi:\url{http://doi.org/10.1093/bib/bbr009}.
-
-\hyperdef{}{ref-White2010}{\label{ref-White2010}}
+doi:\href{https://doi.org/10.1093/bib/bbr009}{10.1093/bib/bbr009}.
+
+\hypertarget{ref-White2010}{}
 31. \textbf{White JR}, \textbf{Navlakha S}, \textbf{Nagarajan N},
@@
 microbial diversity studies. BMC Bioinformatics \textbf{11}:152.
-doi:\url{http://doi.org/10.1186/1471-2105-11-152}.
-
-\hyperdef{}{ref-AlGhalith2016}{\label{ref-AlGhalith2016}}
+doi:\href{https://doi.org/10.1186/1471-2105-11-152}{10.1186/1471-2105-11-152}.
+
+\hypertarget{ref-AlGhalith2016}{}
 32. \textbf{Al-Ghalith GA}, \textbf{Montassier E}, \textbf{Ward HN},
@@
 \textbf{12}:e1004658.
-doi:\url{http://doi.org/10.1371/journal.pcbi.1004658}.
-
-\hyperdef{}{ref-He2015}{\label{ref-He2015}}
+doi:\href{https://doi.org/10.1371/journal.pcbi.1004658}{10.1371/journal.pcbi.1004658}.
+
+\hypertarget{ref-He2015}{}
 33. \textbf{He Y}, \textbf{Caporaso JG}, \textbf{Jiang X-T},
@@
 Microbiome \textbf{3}.
-doi:\url{http://doi.org/10.1186/s40168-015-0081-x}.
-
-\hyperdef{}{ref-Kopylova2016}{\label{ref-Kopylova2016}}
+doi:\href{https://doi.org/10.1186/s40168-015-0081-x}{10.1186/s40168-015-0081-x}.
+
+\hypertarget{ref-Kopylova2016}{}
 34. \textbf{Kopylova E}, \textbf{Navas-Molina JA}, \textbf{Mercier C},
@@
 mSystems \textbf{1}:e00003--15.
-doi:\url{http://doi.org/10.1128/msystems.00003-15}.
-
-\hyperdef{}{ref-Schmidt2014}{\label{ref-Schmidt2014}}
+doi:\href{https://doi.org/10.1128/msystems.00003-15}{10.1128/msystems.00003-15}.
+
+\hypertarget{ref-Schmidt2014}{}
 35. \textbf{Schmidt TSB}, \textbf{Rodrigues JFM}, \textbf{Mering C von}.
@@
 operational taxonomic units. Environ Microbiol \textbf{17}:1689--1706.
-doi:\url{http://doi.org/10.1111/1462-2920.12610}.
-
-\hyperdef{}{ref-Schloss2016a}{\label{ref-Schloss2016a}}
+doi:\href{https://doi.org/10.1111/1462-2920.12610}{10.1111/1462-2920.12610}.
+
+\hypertarget{ref-Schloss2016a}{}
 36. \textbf{Schloss PD}. 2016. Application of a database-independent
@@
 methods. mSystems \textbf{1}:e00027--16.
-doi:\url{http://doi.org/10.1128/msystems.00027-16}.
-
-\hyperdef{}{ref-Matthews1975}{\label{ref-Matthews1975}}
+doi:\href{https://doi.org/10.1128/msystems.00027-16}{10.1128/msystems.00027-16}.
+
+\hypertarget{ref-Matthews1975}{}
 37. \textbf{Matthews B}. 1975. Comparison of the predicted and observed
@@
 (BBA) - Protein Structure \textbf{405}:442--451.
-doi:\url{http://doi.org/10.1016/0005-2795(75)90109-9}.
-
-\hyperdef{}{ref-Pruesse2007}{\label{ref-Pruesse2007}}
+doi:\href{https://doi.org/10.1016/0005-2795(75)90109-9}{10.1016/0005-2795(75)90109-9}.
+
+\hypertarget{ref-Pruesse2007}{}
 38. \textbf{Pruesse E}, \textbf{Quast C}, \textbf{Knittel K},
@@
 ARB. Nucleic Acids Research \textbf{35}:7188--7196.
-doi:\url{http://doi.org/10.1093/nar/gkm864}.
-
-\hyperdef{}{ref-Baxter2016}{\label{ref-Baxter2016}}
+doi:\href{https://doi.org/10.1093/nar/gkm864}{10.1093/nar/gkm864}.
+
+\hypertarget{ref-Baxter2016}{}
 39. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
@@
 Genome Medicine \textbf{8}.
-doi:\url{http://doi.org/10.1186/s13073-016-0290-3}.
-
-\hyperdef{}{ref-Schloss2012}{\label{ref-Schloss2012}}
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
+\hypertarget{ref-Schloss2012}{}
 40. \textbf{Schloss PD}, \textbf{Schubert AM}, \textbf{Zackular JP},
@@
 Microbes \textbf{3}:383--393.
-doi:\url{http://doi.org/10.4161/gmic.21008}.
-
-\hyperdef{}{ref-Johnston2016}{\label{ref-Johnston2016}}
+doi:\href{https://doi.org/10.4161/gmic.21008}{10.4161/gmic.21008}.
+
+\hypertarget{ref-Johnston2016}{}
 41. \textbf{Johnston ER}, \textbf{Rodriguez-R LM}, \textbf{Luo C},
@@
 tundra ecosystem. Front Microbiol \textbf{7}.
-doi:\url{http://doi.org/10.3389/fmicb.2016.00579}.
-
-\hyperdef{}{ref-Henson2016}{\label{ref-Henson2016}}
+doi:\href{https://doi.org/10.3389/fmicb.2016.00579}{10.3389/fmicb.2016.00579}.
+
+\hypertarget{ref-Henson2016}{}
 42. \textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst JL},
@@
 \textbf{1}:e00028--16.
-doi:\url{http://doi.org/10.1128/msphere.00028-16}.
-
-\hyperdef{}{ref-Schloss2010}{\label{ref-Schloss2010}}
+doi:\href{https://doi.org/10.1128/msphere.00028-16}{10.1128/msphere.00028-16}.
+
+\hypertarget{ref-Schloss2010}{}
 43. \textbf{Schloss PD}. 2010. The effects of alignment quality,
@@
 \textbf{6}:e1000844.
-doi:\url{http://doi.org/10.1371/journal.pcbi.1000844}.
-
-\hyperdef{}{ref-language2015}{\label{ref-language2015}}
+doi:\href{https://doi.org/10.1371/journal.pcbi.1000844}{10.1371/journal.pcbi.1000844}.
+
+\hypertarget{ref-language2015}{}
 44. \textbf{R Core Team}. 2015. R: A language and environment for
@@
 
-\hyperdef{}{ref-wesanderson}{\label{ref-wesanderson}}
+\hypertarget{ref-wesanderson}{}
 45. \textbf{Ram K}, \textbf{Wickham H}. 2015. wesanderson: A wes
@@
 
-\hyperdef{}{ref-dplyr}{\label{ref-dplyr}}
+\hypertarget{ref-dplyr}{}
 46. \textbf{Wickham H}, \textbf{Francois R}. 2016. dplyr: A grammar of
@@
 
-\hyperdef{}{ref-tidyr}{\label{ref-tidyr}}
+\hypertarget{ref-tidyr}{}
 47. \textbf{Wickham H}. 2016. tidyr: Easily tidy data with `spread()`
@@
 
-\hyperdef{}{ref-cowplot}{\label{ref-cowplot}}
+\hypertarget{ref-cowplot}{}
 48. \textbf{Wilke CO}. cowplot: Streamlined plot theme and plot
@@
 
-\hyperdef{}{ref-ggplot2}{\label{ref-ggplot2}}
+\hypertarget{ref-ggplot2}{}
 49. \textbf{Wickham H}. 2009. ggplot2: Elegant graphics for data

### 2017-02-09 submission/Westcott_OptiClust_mSphere_2016.Rmd
Change from tiff to eps for figures

@@
 
-\includegraphics[width=6.0in]{../results/figures/performance.png}
-
 **Figure 1.
@@
 
-\newpage
-
-\includegraphics[width=3.5in]{../results/figures/speed_memory.png}
-
 **Figure 2.
@@
 The y-axis is scaled by the square root to demonstrate the relationship between the time and memory requirements relative to the number of unique sequences squared.
-
-\newpage
-
-\includegraphics[width=3.5in]{../results/figures/split_mcc.png}
 
@@
 
-\includegraphics[width=6.0in]{../results/figures/performance_other.png}
+**Table S1.
+Summary of the average number of true positives, true negatives, false positives, false negatives and the resulting Matthews correlation coefficient for each of the clustering methods that were analyzed in this study for each of the six datasets.** Blank values indicate that those conditions could not be completed in 50 hours with 45 GB of RAM.
 
@@
 
-\newpage
-
-\includegraphics[width=3.5in]{../results/figures/optimization.png}
-
 **Figure S2.
@@
 
-\newpage
-
-\newpage
-
 **Supplemental text.** Worked example of how OptiClust algorithm clusters sequences into OTUs.
-
-**Table S1.
-Summary of the average number of true positives, true negatives, false positives, false negatives and the resulting Matthews correlation coefficient for each of the clustering methods that were analyzed in this study for each of the six datasets.** Blank values indicate that those conditions could not be completed in 50 hours with 45 GB of RAM.
 

### 2017-02-09 submission/Westcott_OptiClust_mSphere_2016.tex
Change from tiff to eps for figures

@@
 
-\includegraphics[width=6.0in]{results/figures/performance.png}
-
 \textbf{Figure 1. Comparison of de novo clustering algorithms.} Plot of
@@
 
-\newpage
-
-\includegraphics[width=3.5in]{results/figures/speed_memory.png}
-
 \textbf{Figure 2. OptiClust performance} The average execution time (A)
@@
 
-\newpage
-
-\includegraphics[width=3.5in]{results/figures/split_mcc.png}
-
 \textbf{Figure 3. Effects of taxonomically splitting the datasets on
@@
 
-\includegraphics[width=6.0in]{results/figures/performance_other.png}
+\textbf{Table S1. Summary of the average number of true positives, true
+negatives, false positives, false negatives and the resulting Matthews
+correlation coefficient for each of the clustering methods that were
+analyzed in this study for each of the six datasets.} Blank values
+indicate that those conditions could not be completed in 50 hours with
+45 GB of RAM.
 
@@
 
-\newpage
-
-\includegraphics[width=3.5in]{results/figures/optimization.png}
-
 \textbf{Figure S2. The OptiClust algorithm rapidly converges to optimize
@@
 
-\newpage
-
-\newpage
-
 \textbf{Supplemental text.} Worked example of how OptiClust algorithm
 clusters sequences into OTUs.
-
-\textbf{Table S1. Summary of the average number of true positives, true
-negatives, false positives, false negatives and the resulting Matthews
-correlation coefficient for each of the clustering methods that were
-analyzed in this study for each of the six datasets.} Blank values
-indicate that those conditions could not be completed in 50 hours with
-45 GB of RAM.
 

### 2017-02-09 code/build_optimization_figure.R

@@ -59,5 +59,5 @@ all_step_data %>%
 	expand_limits(y=c(0.7,0.9), x=c(0,17)) +
 	scale_x_continuous(breaks=c(0,5,10,15,20), labels=c(0,5,10,15,20)) +
 	my_theme +
-	ggsave("results/figures/optimization.tiff", width=3.5, height=2.5, unit='in') +
+	ggsave("results/figures/optimization.eps", width=3.5, height=2.5, unit='in') +
 	ggsave("results/figures/optimization.png", width=3.5, height=2.5, unit='in')

### 2017-02-09 code/build_performance_figure.R

@@ -127,11 +127,10 @@ my_legend <- theme(
 		legend.margin = margin(t=0,4,4,4)
 	)
 
-
 ggdraw() +
 	draw_plot(mcc + theme(axis.title.y=element_text(margin=margin(r=13,l=0,t=0,b=0))),x=0,y=0.7,width=1,height=0.3) +
 	draw_plot(sobs + theme(axis.title.y=element_text(margin=margin(r=11,l=0,t=0,b=0))) + my_legend,x=0,y=0.4,1,0.3) +
 	draw_plot(secs,x=0,y=0.0,1,0.4) +
-	draw_plot_label(c("A", "B", "C"), x=c(0,0,0), y=c(1.00,0.71,0.41), size=12) +
-	ggsave('results/figures/performance.tiff', width=6.875, height=5.5, unit='in') +
+	draw_plot_label(c("A", "B", "C"), x=c(0,0,0), y=c(1.00,0.71,0.41), size=12)# +
+	ggsave('results/figures/performance.eps', width=6.875, height=5.5, unit='in') +
 	ggsave('results/figures/performance.png', width=6.875, height=5.5, unit='in')

### 2017-02-09 code/build_performance_other_figure.R

@@ -122,5 +122,5 @@ ggdraw() +
 	draw_plot(sobs + theme(axis.title.y=element_text(margin=margin(r=11,l=0,t=0,b=0))) + my_legend,x=0,y=0.45,1,0.275) +
 	draw_plot(secs,x=0,y=0.0,1,0.45) +
 	draw_plot_label(c("A", "B", "C"), x=c(0,0,0), y=c(1.00,0.735,0.46), size=12) +
-	ggsave('results/figures/performance_other.tiff', width=6.875, height=5.5, unit='in') +
-	ggsave('results/figures/performance_other.png', width=6.875, height=5.5, unit='in')
+	ggsave('results/figures/performance_other.eps', width=6.875, height=5.5, units='in') +
+	ggsave('results/figures/performance_other.png', width=6.875, height=5.5, units='in')

### 2017-02-09 code/build_scaling_figure.R

@@ -143,5 +143,5 @@ ggdraw() +
 	draw_plot(speed_plot,  x=0, y=0.535, 1.0, 0.465) +
 	draw_plot(memory_plot, x=0, y=0.000, 1.0, 0.535) +
 	draw_plot_label(c("A", "B"), x=c(0,0), y=c(1.00,0.54), size=12) +
-	ggsave('results/figures/speed_memory.tiff', width=3.5, height=4.0, unit='in') +
+	ggsave('results/figures/speed_memory.eps', width=3.5, height=4.0, unit='in') +
 	ggsave('results/figures/speed_memory.png', width=3.5, height=4.0, unit='in')

### 2017-02-09 code/build_split_figure.R

@@ -104,5 +104,5 @@ figure <- split_data %>% filter(metric == "mcc") %>%
 ggdraw() +
 	draw_plot(figure,  x=0, y=0, 1.0, 1) +
 	draw_plot_label(pretty_datasets, x=rep(0.15,4), y=c(0.825,0.6,0.375,0.15), size=10, hjust=0) +
-	ggsave("results/figures/split_mcc.tiff", width=3.5, height=5.0, unit="in") +
+	ggsave("results/figures/split_mcc.eps", width=3.5, height=5.0, unit="in") +
 	ggsave("results/figures/split_mcc.png", width=3.5, height=5.0, unit="in")

### 2017-02-09 submission/Westcott_OptiClust_mSphere_2016.tex
Render write.paper on axiom

--- 2017-02-03 submission/Westcott_OptiClust_mSphere_2016.tex
@@
 correlation coefficient. This was repeated 10 times for each dataset.
-
-\newpage
 
